### 2022-11-09 paper/paper.tex
[ADD] Draft outline.

\documentclass[conference,draft]{IEEEtran}

\usepackage{cite}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{graphicx}
\graphicspath{{figures/}}

\newcommand{\TG}[1]{\color{red}\textsc{From Tristan: }#1\color{black}}
\newcommand{\MD}[1]{\color{magenta}\textsc{From Tristan: }#1\color{black}}
\newcommand{\HL}[1]{\hl{#1}}

\title{Surveying the Performance Bottleneck of Neuroimaging Pre-Processing}

\author{Mathieu Dugr\'e, Tristan Glatard}

\begin{document}
\maketitle

\begin{abstract}
	% TODO

\end{abstract}

\section{Introduction}
% TODO

% MRI preprocessing

% Motivate the need to profile the bottleneck of applications.
% * Major part of neuroimaging study
% * Faster pre-processing would enable larger study and more clinical use.
% Identifying the major bottleneck will let us know what to optimize down the line. (compute, compression, etc.)
% Reduced precision + Compression ?

\section{Background}
% TODO
\subsection{MRI Pre-Processing}
% MRI modalities
% Pre-processing tools
% Pre-processing step (fMRIPrep figure, + something for DIPY)

\subsection{Profiling (VTune)}
% Either profiling in general, or VTune in particular.

% VTune
% Lightweight profiler
% Info on function and module runtime

\subsection{Reduced Precision}
% Quickly motivate the need for this paper
% Summarize the advantages and disavantages of using reduced precison.
% Talk about the potential benefits it could have to the field of neurorimaging,
% but that profiling the application hotspot is required to perform it.

\subsection{Compression}
% How it can help with intermediate file.
% Link to reduced precision, which could enable a sort of compression by of it self.

\section{Methods}
\subsection{Infrastructure}
% ??? Should we include that ???
% Doesn't really mattter for this paper, I think.
% But doesn't harm either. For example, the performance could vary depending on architecture.

\subsection{Dataset}
% TODO
% size
% distribution: age, sex, race?, etc.
% Look into CoRR and HPC datasets.
% Need to have quality, yet not prestine, data with subject info for at least age and sex.

\subsection{Pipelines}
% TODO 
% Motivation for those pipelines:

% MRI modalities
% fMRIprep (full) + sub-pipelines (sMRI + fMRI)
% DIPY pipelines (dMRI)

% Use case from papers
% LivingPark pipelines (sMRI)

\section{Results}
% TODO
% 2-step approach:
% 1. Which pipeline constitute the long execution time in preprocessing?
% 2. For those, which functions take the most time to execute? (hotspot)

\subsection{Makespan}
% Identify which tools take the longest to execute.

% ###    FIGURE    ###
%  Bar chart with the total makespan of the pipeline execution (mean + std)
% ### END FIGURE ####
% This will let us decided which pipeline to investigate more.

\subsection{Hotspot}
% (1-2 pages)
% TopK (approx. 10) longest makepsan application. The rest can go in supplemental.
% ###    FIGURE    ###
% (Large pipeline)
% fMRIPrep
% sMRIPrep
% DIPY
% 
% (Pipeline components)
% FreeSurfer
% ANTs ??
% SPM DARTEL
% ### END FIGURE ####

% Supplementary figure for the sub-components of large pipelines.

\section{Discussion}
% TODO
% Principal Bottlenecks

% Limitation
% Can only perform profiling of open-source pipelines
% Or pipelines with debug-info.

% Opportunity for optimisation (future work)
% Reduced precision for: Compute or Compression

\section{Conclusion}
% TODO

\section*{Acknowledgement}
% TODO

\section*{Data Availability}
% TODO

\section*{Conflict of Interests}
The authors report no conflict of interests.

\section*{Supplementary Figures}
% Pipelines' hotspot with bottomK makespan

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv, paper.bib}

\end{document}

### 2022-11-10 paper/paper.tex
[ADD] Draft for the introduction.
Plus some more outline for the other sections.

@@
-\documentclass[conference,draft]{IEEEtran}
+\documentclass[conference]{IEEEtran}
 
@@
 
-\title{Surveying the Performance Bottleneck of Neuroimaging Pre-Processing}
+\title{A Detailed Analysis of Performance Bottlenecks in MRI Pre-Processing}
 
@@
 \section{Introduction}
-% TODO
+Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to
+perform clinical diagnosis and for researchers to develop a better understanding
+of the brain.
+Three main MRI techniques exist: strutural MRI (sMRI), functional MRI (fMRI),
+and diffusion MRI (dMRI).
+While other modalities such as EEG, CT, and PET exist, we focus on MRIs for
+their broad adoption and non-invasive property.
+However, MRI data analysis is challeging due to computionally expensive
+pre-processing and large output and intermediate data size.
 
-% MRI preprocessing
+Many toolboxes have been developed to tackle the challenging task of
+pre-processing MRI data.
+Among those, some openly available and commonly accepted emerged:
+AFNI~\cite{Cox1996-sl}, ANTs~\cite{Avants_undated-fu},
+FreeSurfer~\cite{Fischl2012-bp}, FSL~\cite{Jenkinson2012-cq}, and
+SPM~\cite{Friston2007-ag}.
+While the previous pipelines focuse on specific steps of MRI pre-processing,
+tools like fMRIPrep~\cite{Esteban2019-og} and DIPY~\cite{Garyfallidis2014-ve}
+combine various pipelines to produce an end-to-end pre-processing workflow.
 
-% Motivate the need to profile the bottleneck of applications.
-% * Major part of neuroimaging study
-% * Faster pre-processing would enable larger study and more clinical use.
-% Identifying the major bottleneck will let us know what to optimize down the line. (compute, compression, etc.)
-% Reduced precision + Compression ?
+The MRI pre-processing of a subject can take several hours preventing the
+clinical use where data analysis might be required in timely manner.
+Furthermore, the lenghty pre-processing hinders researcher with limited 
+computational resources to perform research on large cohorts.
+
+The first step to improve performance of an application is to understand its
+performance bottleneck.
+Following, efforts can be focused on those performance critical parts.
+For example, better algorithm design or usage of hardware accelerator can 
+reduce computation time.
+Moreover, compression techniques can be used to lower data transfer time for
+output and intermediate data.
+Lastly, reduced precision arithmetic can lower computation time,
+memory footprint, and storage size.
+While these techniques can provide significant performance improvements, their
+implementation can require substantial effort.
+Therefore, it is critical to understand the performance bottleneck of MRI
+pre-processing pipelines to improve their performance.
+
+To the best of our knowledge, there are no comprehensive study of the
+bottlenecks in MRI pre-processing.
+In this paper, we characterize computational and I/O cost of several commonly
+adopted MRI pre-processing pipelines.
+The results of our analysis serve as a reference for future efforts to optimise
+the MRI pre-processing workflow.
 
 \section{Background}
-% TODO
+
 \subsection{MRI Pre-Processing}
-% MRI modalities
-% Pre-processing tools
-% Pre-processing step (fMRIPrep figure, + something for DIPY)
+% Discussion of the pre-processing steps.
+% Explain their computation and I/O requirements.
 
-\subsection{Profiling (VTune)}
-% Either profiling in general, or VTune in particular.
+% sMRI
+
+% fMRI
+
+% dMRI
+
+\subsection{Intel VTune profiler}
+% Requirements for profiling the Neuroimaging field.
+% Multi-language support
+% Lightweight
+% Info on function and module runtime
 
 % VTune
-% Lightweight profiler
-% Info on function and module runtime
 
-\subsection{Reduced Precision}
-% Quickly motivate the need for this paper
-% Summarize the advantages and disavantages of using reduced precison.
-% Talk about the potential benefits it could have to the field of neurorimaging,
-% but that profiling the application hotspot is required to perform it.
-
-\subsection{Compression}
-% How it can help with intermediate file.
-% Link to reduced precision, which could enable a sort of compression by of it self.
+% Challenges remaining
+% Debug information when compiling, for valuable information
+% Combining the results from multiple analysis. Different input data can 
+% vary the analysis significantly. (due to application branching and convergence)
 
@@
 \subsection{Infrastructure}
-% ??? Should we include that ???
-% Doesn't really mattter for this paper, I think.
-% But doesn't harm either. For example, the performance could vary depending on architecture.
+% TODO
 
@@
 % TODO
+% !!! TBD !!!
 % size
@@
 % TODO 
+% !!! TBD !!!
 % Motivation for those pipelines:
@@
 % LivingPark pipelines (sMRI)
+
+\subsection{Profiling}
+% How we did the profiling.
+% Code to repository
+% Aggregation of the results from different subjects.
 
@@
 % TODO
+The code to compile, profile, and execute the pipelines is publicly available at:
+
+\href{https://github.com/mathdugre/neuro-bottleneck}{https://github.com/mathdugre/neuro-bottleneck}
+% TODO Update URL to slashbin org, when created.
 
@@
 \bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv, paper.bib}
+% \bibliography{IEEEabrv, paper}
+\bibliography{paper}
 

### 2022-11-15 paper/paper.tex
[FIX] syntax and grammar for introduction.

@@
 \section{Introduction}
-Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to
-perform clinical diagnosis and for researchers to develop a better understanding
-of the brain.
-Three main MRI techniques exist: strutural MRI (sMRI), functional MRI (fMRI),
-and diffusion MRI (dMRI).
-While other modalities such as EEG, CT, and PET exist, we focus on MRIs for
-their broad adoption and non-invasive property.
-However, MRI data analysis is challeging due to computionally expensive
-pre-processing and large output and intermediate data size.
+Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain.
+Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI).
+While other modalities such as EEG, CT, and PET exist, we focus on MRIs for their broad adoption and non-invasive property.
+However, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.
 
-Many toolboxes have been developed to tackle the challenging task of
-pre-processing MRI data.
-Among those, some openly available and commonly accepted emerged:
-AFNI~\cite{Cox1996-sl}, ANTs~\cite{Avants_undated-fu},
-FreeSurfer~\cite{Fischl2012-bp}, FSL~\cite{Jenkinson2012-cq}, and
-SPM~\cite{Friston2007-ag}.
-While the previous pipelines focuse on specific steps of MRI pre-processing,
-tools like fMRIPrep~\cite{Esteban2019-og} and DIPY~\cite{Garyfallidis2014-ve}
-combine various pipelines to produce an end-to-end pre-processing workflow.
+Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data.
+Among those, some openly available and commonly accepted emerged: AFNI~\cite{Cox1996-sl}, ANTs~\cite{Avants_undated-fu}, FreeSurfer~\cite{Fischl2012-bp}, FSL~\cite{Jenkinson2012-cq}, and SPM~\cite{Friston2007-ag}.
+While the previous pipelines focus on specific steps of MRI pre-processing, tools like fMRIPrep~\cite{Esteban2019-og} and DIPY~\cite{Garyfallidis2014-ve} combine various pipelines to produce a complete pre-processing workflow.
 
-The MRI pre-processing of a subject can take several hours preventing the
-clinical use where data analysis might be required in timely manner.
-Furthermore, the lenghty pre-processing hinders researcher with limited 
-computational resources to perform research on large cohorts.
+The MRI pre-processing of a subject can take several hours, preventing the clinical use where data analysis might be required in timely manner.
+The lengthy pre-processing hinders researchers with limited computational resources to perform research on large cohorts.
 
-The first step to improve performance of an application is to understand its
-performance bottleneck.
-Following, efforts can be focused on those performance critical parts.
-For example, better algorithm design or usage of hardware accelerator can 
-reduce computation time.
-Moreover, compression techniques can be used to lower data transfer time for
-output and intermediate data.
-Lastly, reduced precision arithmetic can lower computation time,
-memory footprint, and storage size.
-While these techniques can provide significant performance improvements, their
-implementation can require substantial effort.
-Therefore, it is critical to understand the performance bottleneck of MRI
-pre-processing pipelines to improve their performance.
+The first step to improve performance of an application is to understand its performance bottleneck.
+Following, we can focus our efforts on the performance critical components.
+For example, better algorithm designs or usage of hardware accelerator can reduce computation time.
+Compression techniques can lower data transfer time for output and intermediate data.
+At last, reduced precision arithmetic can lower computation time, memory footprint, and storage size.
+While these techniques can provide significant performance improvements, their implementation can require substantial effort.
+Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.
 
-To the best of our knowledge, there are no comprehensive study of the
-bottlenecks in MRI pre-processing.
-In this paper, we characterize computational and I/O cost of several commonly
-adopted MRI pre-processing pipelines.
-The results of our analysis serve as a reference for future efforts to optimise
-the MRI pre-processing workflow.
+To the best of our knowledge, there is no comprehensive study of the bottlenecks in MRI pre-processing.
+In this paper, we characterize computational and I/O cost of several commonly adopted MRI pre-processing pipelines.
+The results of our analysis serve as a reference for future efforts to optimize the MRI pre-processing workflow.
 
@@
 \subsection{MRI Pre-Processing}
+% Look at the benchamrked tool and the information needed to abstract those tools.
 % Discussion of the pre-processing steps.
@@
 % distribution: age, sex, race?, etc.
-% Look into CoRR and HPC datasets.
+% Look into **CoRR** (more representative for this project) and HPC datasets.
 % Need to have quality, yet not prestine, data with subject info for at least age and sex.

### 2022-11-28 sbatch/utils.sh
[ADD] nthread override for slurm + [ADD] output dir for profiling data.

@@ -17,6 +17,8 @@ Usage: sh ${FUNC} <SIF_IMG> -i <INPUT_DIR> -o <OUTPUT_DIR> [-j <NTHREAD>]
     -j NTHREAD: Number of threads used for the application. By default use all.
 
     -o OUTPUT_DIR: Path to the output directory.
+
+    -p PROFILING_DIR: Path to write profiling data.
 "
 
 # Initalize env loading on either Slashbin and Compute Canada
@@ -26,7 +28,6 @@ then
     mdoule load python/3.10
 fi
 [[ -z ${SLURM_TMPDIR:+x} ]] && export SLURM_TMPDIR=/disk5/${USER}/mri-bottleneck
-NTHREAD=$(nproc)
 
 # Parse arguments
 function validate_opt(){
@@ -36,6 +37,9 @@ function validate_opt(){
         exit 1
     fi
 }
+
+NTHREAD=$(nproc)
+PROFILING_DIR="vtune_output"
 
 PARAMS=""
 while (( $# )); do
@@ -53,6 +57,10 @@ while (( $# )); do
     -o | --output)
         validate_opt $@
         OUTPUT_DIR=$2
+        shift 2;;
+    -p | --profiling)
+        validate_opt $@
+        PROFILING_DIR=$2
         shift 2;;
     -* | --*=) # unsupported flags
       echo "Error: Unsupported flag ${1}" >&2
@@ -80,10 +88,16 @@ if [[ ! -f ${SIF_IMG} ]]; then
     echo "Error: SIF_IMG path does not exist: ${SIF_IMG} "
     exit 1
 fi
+if [[ -n ${SLURM_NPROCS}]]; then
+    NTHREAD=$((NTHREAD>SLURM_NPROCS ? SLURM_NPROCS : NTHREAD))
+    echo "WARNING: ${SLURM_NPROCS} core(s) allocated by SLURM. NTHREAD will be overidden to ${SLURM_NPROCS}"
+fi
+mkdir -p ${PROFILING_DIR}
 
 export SIF_IMG
 export INPUT_DIR
 export OUTPUT_DIR
+export PROFILING_DIR
 export NTHREAD
 export DATASET=$(basename ${INPUT_DIR})
[…]

### 2022-11-28 sbatch/utils.sh
[REMOVE] profiling dir and nthread slurm overidde.
It was creating unessecary issues.

@@ -17,8 +17,6 @@ Usage: sh ${FUNC} <SIF_IMG> -i <INPUT_DIR> -o <OUTPUT_DIR> [-j <NTHREAD>]
     -j NTHREAD: Number of threads used for the application. By default use all.
 
     -o OUTPUT_DIR: Path to the output directory.
-
-    -p PROFILING_DIR: Path to write profiling data.
 "
 
 # Initalize env loading on either Slashbin and Compute Canada
@@ -58,10 +56,6 @@ while (( $# )); do
         validate_opt $@
         OUTPUT_DIR=$2
         shift 2;;
-    -p | --profiling)
-        validate_opt $@
-        PROFILING_DIR=$2
-        shift 2;;
     -* | --*=) # unsupported flags
       echo "Error: Unsupported flag ${1}" >&2
       exit 1
@@ -88,16 +82,10 @@ if [[ ! -f ${SIF_IMG} ]]; then
     echo "Error: SIF_IMG path does not exist: ${SIF_IMG} "
     exit 1
 fi
-if [[ -n ${SLURM_NPROCS}]]; then
-    NTHREAD=$((NTHREAD>SLURM_NPROCS ? SLURM_NPROCS : NTHREAD))
-    echo "WARNING: ${SLURM_NPROCS} core(s) allocated by SLURM. NTHREAD will be overidden to ${SLURM_NPROCS}"
-fi
-mkdir -p ${PROFILING_DIR}
 
 export SIF_IMG
 export INPUT_DIR
 export OUTPUT_DIR
-export PROFILING_DIR
 export NTHREAD
 export DATASET=$(basename ${INPUT_DIR})
 export SUBJECT_ID=$(sed -n $(( 1 + ${SLURM_ARRAY_TASK_ID} ))p ${INPUT_DIR}/participants.tsv | cut -f1)
@@ -109,7 +97,6 @@ Pipeline: ${FUNC}
 SIF_IMG: ${SIF_IMG}
 INPUT_DIR: ${INPUT_DIR}
 OUTPUT_DIR: ${OUTPUT_DIR}
-PROFILING_DIR: ${PROFILING_DIR}
 NTHREAD: ${NTHREAD}
 DATASET: ${DATASET}
 SUBJECT_ID: ${SUBJECT_ID}

### 2022-12-01 paper/paper.tex
[ADD] [DRAFT] Method section.

@@
 \section{Methods}
+In this section, we explain our choice of pipelines and dataset. Furthermore, we describe the methods used for profiling the dataset with the different applications.
+
+For a detailed view of our methods, we redirect the reader to our GitHub repository (see \ref{data-availability}) which contains the code for:
+\begin{itemize}
+	\item Acquiring the datasets.
+	\item Building the containerized image for the pipelines.
+	\item Executing the pipeline locally or on a SLURM cluster.
+	\item Visualizing the profiling data.
+\end{itemize}
+
+\subsection{Pipelines}
+In this analysis, we aim at covering the pre-processing performance for different MRI modalities.
+We study the performance bottleneck of large toolboxes as well as individual pipelines.
+Moreover, we use the pipelines from the LivingPark project as a real use case of MRI pre-processing.
+% TODO citation for LivingPark project. 
+% Q: Ask Tristan about it.
+
+First, we choose these commonly accepted workflows: fMRIPrep anatomical only without and without FreeSurfer, fMRIPrep complete workflow, and DIPY.
+This let us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.
+
+Second, we perform a finer grain analysis of the afore mentioned workflows' sub-components.
+The fMRIPrep anatomical workflow use these pipelines: \textit{ANTS Brain Extraction}, \textit{FSL FAST}, and \textit{FreeSurfer recon-all}.
+The fMRIPrep Bold workflow use those: \textit{AFNI 3dTShift}, \textit{FSL MELODIC}, \textit{FSL MCFLIRT}, and \textit{FSL FLIRT}.
+DIPY uses the following pipelines: \dots
+% TODO Decide on workflow sub-components.
+
+Last, the LivingPark project replicate several papers on Parkinson's Disease with a broad range of analysis.
+The following pipelines are used in the study: \textit{FSL SIENA}, \textit{SPM DARTEL}, and \textit{SPM DBM}.
+% TODO Validate which pipelines are used in the project.
+% Ideally use all the LivingPark pipelines. If some cannot be used, explain why.
+
+\subsection{Dataset}
+To ensure the performance profile of the application derived from our work are inclusive to different population, we use dataset with a wide range of age and equal distribution of sex (biological).
+Using a diverse dataset should help to prevent finding of potential performance bottleneck associated to a specific population.
+
+For the sMRI and fMRI pipelines, we use the Consortium for Reliability and Reproducibility (CoRR) dataset~\cite{Zuo2014-xa} publicly available on Datalad\footnote{\href{https://datasets.datalad.org/?dir=/corr/RawDataBIDS}{https://datasets.datalad.org/?dir=/corr/RawDataBIDS}}.
+The CoRR dataset contains 1629 subjects resting state fMRI (rfMRI) data (5093 rfMRI scans) collected across 18 international sites.
+For computational feasibility, we limit ourselves to the \textit{LMU 2} site.
+We chose the \textit{LMU 2} site for its diversity with 40 subjects of age 20-79 (mean 50.8) and female-to-male ratio of 45\%.
+% Q: Might want to discuss the data size and type (anat, func, dwi).
+
+% TODO Choose dataset for dMRI pipelines.
+% CoRR has some sites with dwi data. Could use subjects from those sites.
+
+% TODO Discuss the PPMI dataset.
+
+\subsection{Profiling}
+We use the Intel VTune profile for our analysis.
+On the one hand, it has a low performance overhead and offers reporting profiling data at various granularity.
+On the hand, it requires the debug information of a given application to make sense of its profiling data.
+Otherwise, function name are not available.
+Therefore, as a first step, we re-compiled each pipeline with debug information in a Docker image.
+For use on HPC systems, created Singularity images using Docker2Singularity.
+
+Afterwards, we profile a pipeline by mounting the VTune binary to the Singularity image at execution.
+The profiling result for each (pipeline, subject) pair is stored.
+Combining the Pareto's law and VTune bottom-up report, we identify the top 20\% of function calls that approximately account for 80\% of the pipeline execution time.
+This heuristic allow us to focus our efforts on the performance critical sections of a pipeline.
+% TODO Add code snippet to illustrate Vtune profiling of Singularity pipeline.
+
+Finally, the execution branching of pipeline can vary based on the input data.
+Therefore, we aggregate the profiling results of each pipeline across all subjects.
+We calculate the mean and standard deviation of total execution time of all top 20\% functions and modules of the pipeline.
+% TODO Decide how the mean is calculated.\
+% Either total number of subjects or only with function.
+
 \subsection{Infrastructure}
 % TODO
-
-\subsection{Dataset}
-% TODO
-% !!! TBD !!!
-% size
-% distribution: age, sex, race?, etc.
-% Look into **CoRR** (more representative for this project) and HPC datasets.
-% Need to have quality, yet not prestine, data with subject info for at least age and sex.
-
-\subsection{Pipelines}
-% TODO 
-% !!! TBD !!!
-% Motivation for those pipelines:
-
-% MRI modalities
-% fMRIprep (full) + sub-pipelines (sMRI + fMRI)
-% DIPY pipelines (dMRI)
-
-% Use case from papers
-% LivingPark pipelines (sMRI)
-
-\subsection{Profiling}
-% How we did the profiling.
-% Code to repository
-% Aggregation of the results from different subjects.
+% Decide between slashbin or compute canada. Most likely going to use CC.
 
@@
 
-\section*{Acknowledgement}
-% TODO
-
-\section*{Data Availability}
-% TODO
+\section{Data Availability}
+\label{data-availability}
 The code to compile, profile, and execute the pipelines is publicly available at:
@@
 % TODO Update URL to slashbin org, when created.
+
+\section*{Acknowledgement}
+% TODO
 

### 2022-12-02 paper/paper.tex
[UPDATE] edit draft for methods section.

@@
 \section{Methods}
-In this section, we explain our choice of pipelines and dataset. Furthermore, we describe the methods used for profiling the dataset with the different applications.
+In this section, we explain our choice of pipelines and dataset. We describe the methods used for profiling the dataset with the different applications.
 
-For a detailed view of our methods, we redirect the reader to our GitHub repository (see \ref{data-availability}) which contains the code for:
+For a detailed view of our methods, we redirect the reader to our GitHub repository (see Section~\ref{data-availability}), which contains the code for:
+
 \begin{itemize}
@@
 \subsection{Pipelines}
-In this analysis, we aim at covering the pre-processing performance for different MRI modalities.
-We study the performance bottleneck of large toolboxes as well as individual pipelines.
-Moreover, we use the pipelines from the LivingPark project as a real use case of MRI pre-processing.
+In this analysis, we aim to cover the pre-processing performance for different MRI modalities.
+We study the performance bottleneck of large toolboxes and individual pipelines.
+We use the pipelines from the LivingPark project as a real use case of MRI pre-processing.
 % TODO citation for LivingPark project. 
@@
 
-Second, we perform a finer grain analysis of the afore mentioned workflows' sub-components.
-The fMRIPrep anatomical workflow use these pipelines: \textit{ANTS Brain Extraction}, \textit{FSL FAST}, and \textit{FreeSurfer recon-all}.
-The fMRIPrep Bold workflow use those: \textit{AFNI 3dTShift}, \textit{FSL MELODIC}, \textit{FSL MCFLIRT}, and \textit{FSL FLIRT}.
+Second, we perform a finer grain analysis of the aforementioned workflows' sub-components.
+The fMRIPrep anatomical workflow uses these pipelines: ANTS Brain Extraction, FSL FAST, and FreeSurfer recon-all.
+The fMRIPrep Bold workflow use those: AFNI 3dTShift, FSL MELODIC, FSL MCFLIRT, and FSL FLIRT.
 DIPY uses the following pipelines: \dots
@@
 
-Last, the LivingPark project replicate several papers on Parkinson's Disease with a broad range of analysis.
-The following pipelines are used in the study: \textit{FSL SIENA}, \textit{SPM DARTEL}, and \textit{SPM DBM}.
+Last, the LivingPark project replicates several papers on Parkinson's Disease with a broad range of analysis. The study uses the following pipelines: FSL SIENA, SPM DARTEL, and SPM DBM.
 % TODO Validate which pipelines are used in the project.
@@
 \subsection{Dataset}
-To ensure the performance profile of the application derived from our work are inclusive to different population, we use dataset with a wide range of age and equal distribution of sex (biological).
-Using a diverse dataset should help to prevent finding of potential performance bottleneck associated to a specific population.
+To ensure the performance profile of the application derived from our work is inclusive of different population, we use dataset with a wide range of age and equal distribution of sex (biological).
+Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.
 
@@
 The CoRR dataset contains 1629 subjects resting state fMRI (rfMRI) data (5093 rfMRI scans) collected across 18 international sites.
-For computational feasibility, we limit ourselves to the \textit{LMU 2} site.
-We chose the \textit{LMU 2} site for its diversity with 40 subjects of age 20-79 (mean 50.8) and female-to-male ratio of 45\%.
+For computational feasibility, we limit ourselves to the LMU 2 site.
+We chose the LMU 2 site for its diversity, with 40 subjects of age 20-79 (mean 50.8) and a female-to-male ratio of 45\%.
 % Q: Might want to discuss the data size and type (anat, func, dwi).
@@
 We use the Intel VTune profile for our analysis.
-On the one hand, it has a low performance overhead and offers reporting profiling data at various granularity.
-On the hand, it requires the debug information of a given application to make sense of its profiling data.
-Otherwise, function name are not available.
+On the one hand, it has a low performance overhead and offers reporting profiling data at various granularities.
+On the hand, it requires the debug information of an application to make sense of its profiling data.
+Otherwise, function names are not available.
 Therefore, as a first step, we re-compiled each pipeline with debug information in a Docker image.
@@
 Afterwards, we profile a pipeline by mounting the VTune binary to the Singularity image at execution.
-The profiling result for each (pipeline, subject) pair is stored.
+We store the profiling result for each (pipeline, subject) pair.
 Combining the Pareto's law and VTune bottom-up report, we identify the top 20\% of function calls that approximately account for 80\% of the pipeline execution time.
-This heuristic allow us to focus our efforts on the performance critical sections of a pipeline.
-% TODO Add code snippet to illustrate Vtune profiling of Singularity pipeline.
+This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
+% TODO Add code snippet to illustrate VTune profiling of Singularity pipeline.
 
@@
 We calculate the mean and standard deviation of total execution time of all top 20\% functions and modules of the pipeline.
-% TODO Decide how the mean is calculated.\
+% TODO Decide how the mean is calculated.
 % Either total number of subjects or only with function.

### 2023-05-03 sbatch/ants/brainExtraction.sh
[REFACTOR] reworked output for results and profiling data.
- Allow multiple type of experiments (e.g. varying amount of threads).
- Isolate the preprocessed output from the input. This force
recomputation everytime a pipeline is launched.

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-#SBATCH -J ants-BrainExtraction
+#SBATCH -J ants-brainExtraction
 #SBATCH --array=1
 #SBATCH --time=2:00:00
 #SBATCH --exclusive
@@ -23,10 +23,10 @@ TMPLT="/opt/templates/OASIS"
 
 antsBrainExtraction.sh \
     -d 3 \
-    -a /data/input/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w.nii.gz \
-    -e ${TMPLT}/T_template0.nii.gz \
-    -m ${TMPLT}/T_template0_BrainCerebellumProbabilityMask.nii.gz \
-    -o /data/output/sub-${SUBJECT_ID}/
+    -a /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w.nii.gz \
+    -e \${TMPLT}/T_template0.nii.gz \
+    -m \${TMPLT}/T_template0_BrainCerebellumProbabilityMask.nii.gz \
+    -o /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-1/anat/
 EOT
 
 export SINGULARITYENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD

### 2023-05-03 sbatch/ants/registrationSyN-fp.sh

@@ -1,0 +1,36 @@
+#!/bin/bash
+
+#SBATCH -J ants-registrationSyN-fp
+#SBATCH --array=1
+#SBATCH --time=8:00:00
+#SBATCH --exclusive
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem=0
+# -------------------------
+#SBATCH -o log/%x-%A-%a.out
+#SBATCH -e log/%x-%A-%a.err
+# -------------------------
+set -e
+set -u
+
+# Setup environment and parse args.
+source ./sbatch/pre_run.sh ants registrationSyN-fp -j ${SLURM_CPUS_PER_TASK} $@
+
+export SIF_OPTION="-B $HOME/.cache/templateflow:/templateflow"
+cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+TMPLT="/templateflow/tpl-MNI152NLin2009cAsym/"
+
+antsRegistrationSyN.sh \
+    -d 3 \
+    -f \${TMPLT}/tpl-MNI152NLin2009cAsym_res-01_desc-brain_T1w.nii.gz\
+    -m /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain_seg.nii.gz \
+    -o /data/derivatives/ants/registrationSyN-fp/sub-${SUBJECT_ID}/ses-1/anat/ \
+    -p f
+EOT
+
+export SINGULARITYENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
+source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+
+source ./sbatch/post_run.sh

### 2023-05-03 sbatch/ants/registrationSyN.sh

@@ -1,8 +1,8 @@
 #!/bin/bash
 
-#SBATCH -J ants-brainExtraction-fp
+#SBATCH -J ants-registrationSyN
 #SBATCH --array=1
-#SBATCH --time=2:00:00
+#SBATCH --time=8:00:00
 #SBATCH --exclusive
 #SBATCH --nodes=1
 #SBATCH --ntasks=1
@@ -16,18 +16,17 @@ set -e
 set -u
 
 # Setup environment and parse args.
-source ./sbatch/pre_run.sh ants brainExtraction-fp -j ${SLURM_CPUS_PER_TASK} $@
+source ./sbatch/pre_run.sh ants registrationSyN -j ${SLURM_CPUS_PER_TASK} $@
 
+export SIF_OPTION="-B $HOME/.cache/templateflow:/templateflow"
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
-TMPLT="/opt/templates/OASIS"
+TMPLT="/templateflow/tpl-MNI152NLin2009cAsym/"
 
-antsBrainExtraction.sh \
-    -q 1 \
+antsRegistrationSyN.sh \
     -d 3 \
-    -a /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w.nii.gz \
-    -e \${TMPLT}/T_template0.nii.gz \
-    -m \${TMPLT}/T_template0_BrainCerebellumProbabilityMask.nii.gz \
-    -o /data/derivatives/ants/brainExtraction-fp/sub-${SUBJECT_ID}/ses-1/anat/
+    -f \${TMPLT}/tpl-MNI152NLin2009cAsym_res-01_desc-brain_T1w.nii.gz\
+    -m /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain_seg.nii.gz \
+    -o /data/derivatives/ants/registrationSyN/sub-${SUBJECT_ID}/ses-1/anat/
 EOT
 
 export SINGULARITYENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD

### 2023-05-03 sbatch/fsl/fast.sh

@@ -20,10 +20,9 @@ source ./sbatch/pre_run.sh fsl fast -j ${SLURM_CPUS_PER_TASK} $@
 
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 source /fsl_env.sh
-mkdir -p /data/output/sub-${SUBJECT_ID}/
 fast \
-    -o /data/output/sub-${SUBJECT_ID}/BrainExtractionBrain \
-    /data/input/sub-${SUBJECT_ID}/BrainExtractionBrain.nii.gz
+    -o /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain \
+    /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain.nii.gz
 EOT
 
 export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD

### 2023-05-03 sbatch/post_run.sh

@@ -29,14 +29,8 @@ tar czf ${PROFILING_DIR}.tar.gz -C $(dirname ${PROFILING_DIR}) $(basename ${PROF
 rm -r ${PROFILING_DIR}
 
 # Transfer pipeline output to storage node.
-mkdir -p ${OUTPUT_DIR}
-rsync -aLq --info=progress2 ${SLURM_TMPDIR}/output/ ${OUTPUT_DIR}/
+rsync -aLq --info=progress2 ${SLURM_TMPDIR}/ ${DATA_DIR}/
 rm -rf ${SLURM_TMPDIR}
-
-# Transfer `participant.tsv` file for further use.
-rsync -aLq --info=progress2 \
-    ${DATA_DIR}/participants.tsv \
-    ${OUTPUT_DIR}/participants.tsv
 
 # Delete temporary script
 rm ${TMP_SCRIPT}/${RANDOM_STRING}.sh

### 2023-05-03 sbatch/pre_run.sh

@@ -3,9 +3,9 @@ set -e
 set -u
 
 TOOLKIT=$1; shift
-FUNC=$1; shift
+PIPELINE=$1; shift
 USAGE="
-Usage: sh ${TOOLKIT} ${FUNC} <SIF_IMG> -i <DATA_DIR> -o <PROJECT_DIR> [-j <NTHREAD>]
+Usage: sh ${TOOLKIT} ${PIPELINE} <SIF_IMG> -i <DATA_DIR> -o <PROJECT_DIR> [-j <NTHREAD>]
 
     SIF_IMG: Path to the Singularity image to profile.
 
@@ -13,11 +13,11 @@ Usage: sh ${TOOLKIT} ${FUNC} <SIF_IMG> -i <DATA_DIR> -o <PROJECT_DIR> [-j <NTHRE
 
     -h: print this help message.
 
-    -i DATA_DIR: Path to the input data directory.
+    -d DATA_DIR: Path to the input data directory.
 
     -j NTHREAD: Number of threads used for the application. By default use all.
 
-    -o PROJECT_DIR: Path to the project directory.
+    -p PROJECT_DIR: Path to the project directory.
 "
 
 # Initalize env loading on either Slashbin and Compute Canada
@@ -56,15 +56,15 @@ while (( $# )); do
   case $1 in
     -h | --help)
       echo "${USAGE}" && exit 0;;
-    -i | --input )
+    -d | --dataset)
         validate_opt $@
         DATA_DIR=$2
         shift 2;;
-    -j | --nthread)
+    -j | --NTHREAD)
         validate_opt $@
-        NTHREAD=$(($2>NTHREAD ? NTRHEAD : $2))
+        NTHREAD=$(($2>NTHREAD ? NTHREAD : $2))
         shift 2;;
-    -o | --output)
+    -p | --project)
         validate_opt $@
         PROJECT_DIR=$2
         shift 2;;
@@ -88,7 +88,7 @@ if [[ -z ${1:+x} ]]; then
 fi
 SIF_IMG=$1
 if [[ ! -f ${SIF_IMG} ]]; then
-    echo "Error: SIF_IMG path does not exist: ${SIF_IMG} "
[…]

### 2023-05-03 sbatch/vtune.sh

@@ -4,10 +4,13 @@ set -u
 
 echo "Command to profile: $@"
 
+[[ -z ${SIF_OPTION:+x} ]] && SIF_OPTION=""
+
 singularity exec --cleanenv \
     -B ~/intel/oneapi/vtune/latest/:/vtune \
     -B ${SLURM_TMPDIR}:/data \
     -B ${PROJECT_DIR}:${PROJECT_DIR} \
+    ${SIF_OPTION} \
     ${SIF_IMG} \
     /vtune/bin64/vtune \
     -collect hpc-performance \

### 2023-05-03 run_experiments.sh
[RENAME] file to print_experiments.

@@ -1,10 +1,0 @@
-LFS_HOME=/mnt/lustre/mathdugre
-DATASET=LMU_2
-SIF_DIR=$HOME/containers
-
-N_SUBJECTS=$(awk 'END {print NR-1}' ${LFS_HOME}/datasets/CoRR/${DATASET}/participants.tsv)
-
-echo sbatch --array=1-${N_SUBJECTS} sbatch/ants/brainExtraction.sh $SIF_DIR/ants-debug.sif -i ${LFS_HOME}/datasets/CoRR/${DATASET} -o ${LFS_HOME}/mri-bottleneck
-echo sbatch --array=1-${N_SUBJECTS} sbatch/ants/brainExtraction-fp.sh $SIF_DIR/ants-debug.sif -i ${LFS_HOME}/datasets/CoRR/${DATASET} -o ${LFS_HOME}/mri-bottleneck
-
-echo sbatch --array=1-${N_SUBJECTS} sbatch/fsl/fast.sh $SIF_DIR/fsl-debug.sif -i ${LFS_HOME}/mri-bottleneck/ants/brainExtraction/${DATASET} -o ${LFS_HOME}/mri-bottleneck

### 2023-05-03 sbatch/pre_run.sh
[FIX] transfer files properly to exclude preprocessed output cache.

@@ -136,9 +136,21 @@ mkdir -p ${PROJECT_DIR}
 rm -rf ${SLURM_TMPDIR}
 mkdir -p ${SLURM_TMPDIR}/sub-${SUBJECT_ID}
 rsync -aLq --info=progress2 \
-    --exclude "derivatives" \
     ${DATA_DIR}/sub-${SUBJECT_ID}/ \
     ${SLURM_TMPDIR}/sub-${SUBJECT_ID}/
+
+# Transfer derivatives for a given subject, excluding the potential cache of
+# current pipeline.
+mkdir -p ${SLURM_TMPDIR}/derivatives
+rsync -aLq \
+    --exclude=${TOOLKIT}/${PIPELINE} \
+    --include="**/sub-${SUBJECT_ID}/***" \
+    --exclude="**/sub-*/" \
+    --include "*/" \
+    --exclude="*" \
+    ${DATA_DIR}/derivatives/ \
+    ${SLURM_TMPDIR}/derivatives/
+
 
 # Create temp directory for pipeline scripts.
 export TMP_SCRIPT=tmp-scripts

### 2023-05-03 sbatch/fsl/fast.sh
[FIX] output path must exist.

@@ -20,6 +20,7 @@ source ./sbatch/pre_run.sh fsl fast -j ${SLURM_CPUS_PER_TASK} $@
 
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 source /fsl_env.sh
+mkdir -p /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/
 fast \
     -o /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain \
     /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain.nii.gz

### 2023-05-05 paper/paper.tex
[ADD] example to run vtune in sigularity + [UPDATE] some sections.

@@
 \usepackage{graphicx}
-\graphicspath{{figures/}}
+\graphicspath{{figures}}
+\usepackage{listings}
+\renewcommand{\lstlistingname}{Algorithm}
+\usepackage[binary-units=true]{siunitx}
 
@@
 % Q: Ask Tristan about it.
+% A: GitHub page for now
+% -> Added to paperpile
 
@@
 
-Last, the LivingPark project replicates several papers on Parkinson's Disease with a broad range of analysis. The study uses the following pipelines: FSL SIENA, SPM DARTEL, and SPM DBM.
+Last, the LivingPark project replicates several papers on Parkinson's Disease with a broad range of analysis.
+The study uses the following pipelines: FSL SIENA, SPM DARTEL, and SPM pairwise registration.
 % TODO Validate which pipelines are used in the project.
@@
 We use the Intel VTune profile for our analysis.
-On the one hand, it has a low performance overhead and offers reporting profiling data at various granularities.
+On the one hand, it has a low performance overhead and offers reporting profiling data at various granularity.
 On the hand, it requires the debug information of an application to make sense of its profiling data.
@@
 Therefore, as a first step, we re-compiled each pipeline with debug information in a Docker image.
-For use on HPC systems, created Singularity images using Docker2Singularity.
+For use on HPC systems, we created Singularity images using Docker2Singularity.
 
-Afterwards, we profile a pipeline by mounting the VTune binary to the Singularity image at execution.
+\lstlistingname~\ref{vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Singularity image at execution.
 We store the profiling result for each (pipeline, subject) pair.
@@
 This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
-% TODO Add code snippet to illustrate VTune profiling of Singularity pipeline.
+
+\lstinputlisting[
+	language={bash},
+	caption={Profiling an application within a Singularity container, using VTune profiler.},
+	captionpos=b,
+	label=vtune_example,
+	numbers=left,
+	frame=single,
+	basicstyle=\footnotesize,
+	numbersep=5pt,
+	numberstyle=\tiny\color{gray},
+	rulecolor=\color{black},
+	tabsize=2,
+]{vtune-example.sh}
 
@@
 
+% TODO explain profiling with single-thread vs. multi-thread.
+We set all applications to use a single thread, while profiling. This let
+
 \subsection{Infrastructure}
 % TODO
-% Decide between slashbin or compute canada. Most likely going to use CC.
+% slashbin description
+For our experiments, we used the \textit{Slashbin cluster} at Concordia University.
+The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-240.1.1.el8\_lustre.x86\_64}.
+
+Before profiling each application, we transfer the input data from the shared file system to the compute node.
+After profiling, we transfer back the output to the shared file system.
+This limits profiling variability due to I/O contingency.
+
+The data generated by the VTune profiler is written to a shared file system, for administrative reasons.
+Theoretically, this increase the profiling overhead.
+However, in practice there is no impact since the data generated by the VTune profiler is very small.
 
@@
 % (Pipeline components)
-% FreeSurfer
-% ANTs ??
 % SPM DARTEL
+% SPM pairwise registration
 % ### END FIGURE ####
@@
 % Supplementary figure for the sub-components of large pipelines.
+% Or include them in #(Pipeline components), if space allows.
 
@@
 
-\href{https://github.com/mathdugre/neuro-bottleneck}{https://github.com/mathdugre/neuro-bottleneck}
+\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
 % TODO Update URL to slashbin org, when created.

### 2023-08-11 paper/paper.tex
latest change to paper and experiments.

@@
 \graphicspath{{figures}}
+\usepackage{caption}
+\usepackage{subcaption}
+\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
 \usepackage{listings}
@@
 \usepackage[binary-units=true]{siunitx}
+\usepackage{csvsimple}
 
@@
 \section{Introduction}
+% Intro: Large dataset, long processing time, I/O intensive (derivatives), focus on MRI, no current study, pipelines?
+% **Main question:** What are the main bottleneck in MRI pre-processing?
+% **Why:** Performance of neuroimaging pipeline is critical (larger study, clinical use). However, limited study on performance profile.
 Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain.
@@
 Among those, some openly available and commonly accepted emerged: AFNI~\cite{Cox1996-sl}, ANTs~\cite{Avants_undated-fu}, FreeSurfer~\cite{Fischl2012-bp}, FSL~\cite{Jenkinson2012-cq}, and SPM~\cite{Friston2007-ag}.
-While the previous pipelines focus on specific steps of MRI pre-processing, tools like fMRIPrep~\cite{Esteban2019-og} and DIPY~\cite{Garyfallidis2014-ve} combine various pipelines to produce a complete pre-processing workflow.
+While the previous pipelines focus on specific steps of MRI pre-processing, tools like fMRIPrep~\cite{Esteban2019-og} and dMRIPrep~\cite{Michael2021-4515513} combine various pipelines to produce a complete pre-processing workflow.
 
 The MRI pre-processing of a subject can take several hours, preventing the clinical use where data analysis might be required in timely manner.
-The lengthy pre-processing hinders researchers with limited computational resources to perform research on large cohorts.
+Moreover, the lengthy pre-processing hinders researchers with limited computational resources to perform research on large cohorts.
 
@@
 
-To the best of our knowledge, there is no comprehensive study of the bottlenecks in MRI pre-processing.
-In this paper, we characterize computational and I/O cost of several commonly adopted MRI pre-processing pipelines.
-The results of our analysis serve as a reference for future efforts to optimize the MRI pre-processing workflow.
+The neuroimaging community is well aware that MRI pre-precessing pipelines require long computation time.
+However, to the best of our knowledge, there is no comprehensive study of the performance bottlenecks in MRI pre-processing.
+In this paper, we characterize computational cost of several commonly adopted MRI pre-processing pipelines.
+The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.
 
-\section{Background}
+\section{Methods}
+\subsection{Infrastructure}
+For our experiments, we used the \textit{Slashbin cluster} at Concordia University.
+The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-240.1.1.el8\_lustre.x86\_64}.
 
-\subsection{MRI Pre-Processing}
-% Look at the benchamrked tool and the information needed to abstract those tools.
+\subsection{Data}
+To ensure the performance profiles of the applications derived from our work are inclusive of different population, we used data with a wide range of age and equal distribution of sex (biological).
+Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.
+
+We used the the OpenNeuro ds004513 v.1.0.2 dataset~\cite{ds004513:1.0.2}.
+The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts.
+The within-subject cohort has nine participants (mean age=43, std=7; 4 females) with two session: eyes open and eyes closed.
+The replication cohort has eleven participants (mean age=27, std=5; 6 females) with only the eyes open session.
+We profiled the application using both cohort on the eyes open session, unless explicitly specified.
+
+\subsection{Pipelines}
+% TODO rework
+In this analysis, we aim to cover the pre-processing performance for different MRI modalities.
+We study the performance bottleneck of large toolboxes and individual pipelines.
+Moreover, we use several pipelines from the LivingPark project~\cite{livingpark} as a real use case of MRI pre-processing.
+
+First, we profile these commonly accepted workflows: fMRIPrep anatomical only without and without FreeSurfer, fMRIPrep complete workflow, and dMRIPrep.
+This let us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.
+
+Second, we perform a finer grain analysis of the aforementioned workflows' sub-components. We list the different sub-workflow and describe the key steps for each modality in the section below.
+
+Last, the LivingPark project replicates several papers on Parkinson's Disease with a broad range of analysis.
+The study uses the following pipelines: FSL SIENA/VIENA, SPM VBM, SPM DARTEL, and SPM pairwise registration.
+% TODO Decide which pipelines to use from the project.
+
+% ############################################################
+% Pre-Processing
+% ############################################################
+% Look at the benchmarked tool and the information needed to abstract those tools.
 % Discussion of the pre-processing steps.
 % Explain their computation and I/O requirements.
+\subsection{sMRI Pre-Processing}
+\subsubsection{Intensity correction}
+% NU4 bias correction (Check ANTS brainExtraction)
 
-% sMRI
+\subsubsection{Reference Image}
+% Longitudinal study
+% FreeSurfer mri_robust_template
 
-% fMRI
+\subsubsection{Skull-Stripping}
+% ANTS brainExtraction
 
-% dMRI
+\subsubsection{Spatial Normalization}
+% ANTS registrationSyN
 
-\subsection{Intel VTune profiler}
-% Requirements for profiling the Neuroimaging field.
-% Multi-language support
-% Lightweight
-% Info on function and module runtime
+\subsubsection{Segmentation}
+% FSL FAST
 
-% VTune
+\subsubsection{Surface Reconstruction}
+% FreeSurfer recon-all
 
-% Challenges remaining
-% Debug information when compiling, for valuable information
-% Combining the results from multiple analysis. Different input data can 
-% vary the analysis significantly. (due to application branching and convergence)
+\subsection{fMRI Pre-Processing}
+\subsubsection{Slice-Timing Correction}
+% AFNI 3dTShift
 
-\section{Methods}
-In this section, we explain our choice of pipelines and dataset. We describe the methods used for profiling the dataset with the different applications.
+\subsubsection{Head-Motion Correction}
+% FSL MCFLIRT
 
-For a detailed view of our methods, we redirect the reader to our GitHub repository (see Section~\ref{data-availability}), which contains the code for:
+\subsubsection{Susceptibility Distortion Correction}
+% SDC workflow
 
-\begin{itemize}
-	\item Acquiring the datasets.
-	\item Building the containerized image for the pipelines.
-	\item Executing the pipeline locally or on a SLURM cluster.
-	\item Visualizing the profiling data.
-\end{itemize}
+\subsubsection{Registration}
+% FSL FLIRT
 
-\subsection{Pipelines}
-In this analysis, we aim to cover the pre-processing performance for different MRI modalities.
-We study the performance bottleneck of large toolboxes and individual pipelines.
-We use the pipelines from the LivingPark project as a real use case of MRI pre-processing.
-% TODO citation for LivingPark project. 
-% Q: Ask Tristan about it.
-% A: GitHub page for now
-% -> Added to paperpile
-
-First, we choose these commonly accepted workflows: fMRIPrep anatomical only without and without FreeSurfer, fMRIPrep complete workflow, and DIPY.
-This let us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.
-
-Second, we perform a finer grain analysis of the aforementioned workflows' sub-components.
-The fMRIPrep anatomical workflow uses these pipelines: ANTS Brain Extraction, FSL FAST, and FreeSurfer recon-all.
-The fMRIPrep Bold workflow use those: AFNI 3dTShift, FSL MELODIC, FSL MCFLIRT, and FSL FLIRT.
-DIPY uses the following pipelines: \dots
-% TODO Decide on workflow sub-components.
-
-Last, the LivingPark project replicates several papers on Parkinson's Disease with a broad range of analysis.
-The study uses the following pipelines: FSL SIENA, SPM DARTEL, and SPM pairwise registration.
-% TODO Validate which pipelines are used in the project.
-% Ideally use all the LivingPark pipelines. If some cannot be used, explain why.
-
-\subsection{Dataset}
-To ensure the performance profile of the application derived from our work is inclusive of different population, we use dataset with a wide range of age and equal distribution of sex (biological).
-Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.
-
-For the sMRI and fMRI pipelines, we use the Consortium for Reliability and Reproducibility (CoRR) dataset~\cite{Zuo2014-xa} publicly available on Datalad\footnote{\href{https://datasets.datalad.org/?dir=/corr/RawDataBIDS}{https://datasets.datalad.org/?dir=/corr/RawDataBIDS}}.
-The CoRR dataset contains 1629 subjects resting state fMRI (rfMRI) data (5093 rfMRI scans) collected across 18 international sites.
-For computational feasibility, we limit ourselves to the LMU 2 site.
-We chose the LMU 2 site for its diversity, with 40 subjects of age 20-79 (mean 50.8) and a female-to-male ratio of 45\%.
-% Q: Might want to discuss the data size and type (anat, func, dwi).
-
-% TODO Choose dataset for dMRI pipelines.
-% CoRR has some sites with dwi data. Could use subjects from those sites.
-
-% TODO Discuss the PPMI dataset.
+\subsection{dMRI Pre-Processing}
+\subsubsection{Eddy Currents Distortion Estimation}
+% FSL eddy
 
 \subsection{Profiling}
-We use the Intel VTune profile for our analysis.
-On the one hand, it has a low performance overhead and offers reporting profiling data at various granularity.
-On the hand, it requires the debug information of an application to make sense of its profiling data.
-Otherwise, function names are not available.
-Therefore, as a first step, we re-compiled each pipeline with debug information in a Docker image.
-For use on HPC systems, we created Singularity images using Docker2Singularity.
+Profiling MRI pre-processing pipelines present a set of challenges.
+The neuroimaging community develops pipelines using several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Python}, and \textit{Matlab}, with several pipelines utilizing a combination of multiple languages.
+Pipelines are generally complex and computationally expensive.
 
-\lstlistingname~\ref{vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Singularity image at execution.
-We store the profiling result for each (pipeline, subject) pair.
-Combining the Pareto's law and VTune bottom-up report, we identify the top 20\% of function calls that approximately account for 80\% of the pipeline execution time.
-This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
+The Intel VTune profiler~\cite{vtune_profiler} satisfies these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level.
+However, ome general profiling challenges remain.
+Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules.
+Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds.
+We discuss our approach to these challenges in the next section.
+% TODO add citation on effect of different input data on numerical stability and reduced precision.
 
+First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
+For use on HPC systems, we created Singularity images using Docker2Singularity \lstlistingname~\ref{vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Singularity image at execution.
 \lstinputlisting[
-	language={bash},
+	language={sh},
 	caption={Profiling an application within a Singularity container, using VTune profiler.},
@@
 
-Finally, the execution branching of pipeline can vary based on the input data.
-Therefore, we aggregate the profiling results of each pipeline across all subjects.
-We calculate the mean and standard deviation of total execution time of all top 20\% functions and modules of the pipeline.
-% TODO Decide how the mean is calculated.
-% Either total number of subjects or only with function.
+Second, to estimate an average performance profile, we aggregate the profiling results of each pipeline across all subjects.
+Combining the Pareto's law and VTune hotspots report, we present the functions accounting for most of the CPU time up to 80\% of the total cpu time of the application.
+This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
+We report the mean and standard deviation of the execution time for those functions.
 
-% TODO explain profiling with single-thread vs. multi-thread.
-We set all applications to use a single thread, while profiling. This let
-
-\subsection{Infrastructure}
-% TODO
-% slashbin description
-For our experiments, we used the \textit{Slashbin cluster} at Concordia University.
-The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-240.1.1.el8\_lustre.x86\_64}.
+Last, we profile the application two threading configuration: single-threaded and 32 threads.
+On the one hand, the single-threaded approach limits the potential overhead from I/O on the profiling.
+On the other hand, we evaluate the multi-threading performance of the applications using maximum number of CPU core available on our system (i.e.
+32).
 
@@
 Theoretically, this increase the profiling overhead.
-However, in practice there is no impact since the data generated by the VTune profiler is very small.
+However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as the data from applications.
 
@@
 % 1. Which pipeline constitute the long execution time in preprocessing?
-% 2. For those, which functions take the most time to execute? (hotspot)
+% 2. For those, which functions take the most time to execute? (hotspots)
+% Any key finding: common point between functions, mem bound, fops, etc.
+
+% In this section, analyze the aggregated profiling data obtained with VTune.
+First, a coarse summary of the workflows and pipelines is presented using their makespan.
+Second, we depict detailed analysis for the key functions contributing to the CPU time of each application (hotspots).
 
@@
 % Identify which tools take the longest to execute.
+% This will let us decided which pipeline to investigate more.
+Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan, in seconds, while using 1 thread for one execution of all subjects.
+Unexpectedly, we observe that both ANTS~BrainExtration and ANTS~RegistrationSyN are slower when using single precision compared to double precision.
+% TODO Stat test between mean for single and double precision.
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
+	\caption{Profiling of the pipelines using 1 thread.}
+	\label{fig:makespan-1thread}
+\end{figure}
 
-% ###    FIGURE    ###
-%  Bar chart with the total makespan of the pipeline execution (mean + std)
-% ### END FIGURE ####
-% This will let us decided which pipeline to investigate more.
+\subsection{Hotspots}
+Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspot observed from profiling the fMRIPrep anatomical sub-workflows.
+In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module.
+The function ID are found in the supplementary material Section~\ref{sec:supplementary}.
+The bars show the average CPU time in seconds of a function and the error bars indicate the standard deviation, with the left y-axis showing the scale.
+The light-green dots represent the functions cumulative percentage makespan of the application, with the right y-axis showing the scale.
+The color of the bar indicate the percentage of calls, for the function, that were ``stalled due to demand memory loads and stores''\footnote{\href{https://www.intel.com/content/www/us/en/docs/vtune-profiler/user-guide/2023-0/cpu-metrics-reference.html\#MEMORY-BOUND}{Vtune Memory Bound metric}}, with the left color bar showing the scale.
 
-\subsection{Hotspot}
-% (1-2 pages)
-% TopK (approx. 10) longest makepsan application. The rest can go in supplemental.
-% ###    FIGURE    ###
-% (Large pipeline)
-% fMRIPrep
-% sMRIPrep
-% DIPY
-% 
-% (Pipeline components)
-% SPM DARTEL
-% SPM pairwise registration
-% ### END FIGURE ####
+In figure~\ref{subfig:hotspots-ants-brainExtraction}, we observe that the function 1 and 2 account for a significant portion of the application execution time.
+Moreover, calls from function~1 are memory bound $51.84\pm{4.55}$ percent of the time.
+Table~\ref{extab:hotspots-1thread-ants-brainExtraction} shows that both functions are part of the ITK linear interpolation module.
 
-% Supplementary figure for the sub-components of large pipelines.
-% Or include them in #(Pipeline components), if space allows.
+Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two function accounting for the most time, although in flipped order.
+Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time.
+Compared to the double precision version of ANTS brainExtraction, the third function runtime is significantly larger.
+It is an ITK function transforming points while calculating the displacement field of the image.
+
+Figure~\ref{subfig:hotspots-ants-registrationSyN} and figure~\ref{subfig:hotspots-ants-registrationSyN-fp} show that the execution time for ANTS registrationSyN is dominated by a single function.
+In both case, the ITK function evaluating point during linear interpolation.
+The next two function are also shared between the application, although in flipped order.
+These functions show some memory bound in the range of 16--27.5.
+We also note the three most expensive functions are the same as in both versions of ANTS brainExtraction.
+
+% TODO FreeSurfer 1-thread sub-figure
+
+In figure~\ref{subfig:hotspots-32threads-freesurfer-reconall}, we observe that OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime.
+The only function shown from FreeSurfer recon-all is the same as the first function from figure~\ref{subfig:hotspots-1thread-freesurfer-reconall}.
+However, it only accounts for 2\% of the total makespan compared to 3.8\% while running sequentially.
+
+\begin{figure*}[ht!]
+	\centering
+	% ANTS brainExtraction
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-ants-brainExtraction}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-ants-brainExtraction-fp}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
+	\end{subfigure}
+
+	% ANTS registrationSyN
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-ants-registrationSyN}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-ants-registrationSyN-fp}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
+	\end{subfigure}
+
+	% FreeSurfer recon-all
+	\begin{subfigure}[t]{\textwidth}
+		\caption{}
+		\label{subfig:hotspots-1thread-freesurfer-reconall}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-freesurfer-reconall.png}
+	\end{subfigure}
+
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-32threads-freesurfer-reconall}
+		\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall.png}
+	\end{subfigure}
+	\hfill
+	% FSL FAST
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-fsl-fast}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-fast.png}
+	\end{subfigure}
+
+	\caption{Hotspots profiling of fMRIPrep's anatomical sub-workflows.}
+	\label{fig:hotspots-fmriprep-anat-subworkflow}
+	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
+\end{figure*}
+
+\begin{figure*}[ht!]
+	\centering
+	% FSL MCFLIRT
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-fsl-mcflirt}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-mcflirt.png}
+	\end{subfigure}
+	% FSL FLIRT
+	\hfill
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-fsl-flirt}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-flirt.png}
+	\end{subfigure}
+
+	\caption{Hotspots profiling of fMRIPrep's functional sub-workflows.}
+	\label{fig:hotspots-fmriprep-func-subworkflow}
+	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
+\end{figure*}
 
@@
 % TODO
+
 % Principal Bottlenecks
+% What are common key points between the different applications?
+% Possibly move this to the Results section.
+% - Top 3 functions in ANTS application are the same, although in varying order.
+% -
 
-% Limitation
-% Can only perform profiling of open-source pipelines
-% Or pipelines with debug-info.
+% # Limitation
+% - Can only perform profiling of open-source pipelines or pipelines with debug-info.
+% - No information about space or time for function call.
+% - Dataset only contains healthy subjects (no pathology).
+% - Dataset is from a single site. Possible bias from scanner or protocol.
 
-% Opportunity for optimisation (future work)
-% Reduced precision for: Compute or Compression
+% # Opportunity for optimization (future work)
+% Reduced precision for: Compute or Compression. Reduce memory footprint and compute time. -> Mention the applications with high FP-ops and MEM-bound.
+
+% # Linear Interpolation
+% Theoretical improvements using reduced precision. If time permits, toy experiment with ITK single precision data type at different locations.
+
+% # Single precision slowdown?
+% Possibly due to slower convergence and less optimized code. (Quick drop-in of ITK single precision data type.)
+% Need to investigate more.
+
+% OpenOMP bottleneck (FreeSurfer with 32 threads)
+% Majority of the time is spent waiting to synchronize threads.
+% If times permit, changes the scheduling type and run for a subject to see if we can get performance improvements.
 
@@
 \section{Data Availability}
-\label{data-availability}
-The code to compile, profile, and execute the pipelines is publicly available at:
+\label{sec:data-availability}
+The entire OpenNeuro ds004513 v1.0.2 dataset is freely available to download at:
+\\\href{https://openneuro.org/datasets/ds004513/versions/1.0.2}{https://openneuro.org/datasets/ds004513/versions/1.0.2}.
+% TODO add compiled containers.
+% TODO add csv files from profiled runs.
+% The above can be distributed on DockerHub and Zenodo respectively.
 
-\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
+\section{Code Availability}
+\label{sec:code-availability}
+The code to compile, profile the pipelines, and generate figures is publicly available at:
+\\\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
 % TODO Update URL to slashbin org, when created.
@@
 
+\newpage
+\onecolumn
+\section*{Supplementary material}
+\label{sec:supplementary}
+
+\csvnames{csvcol}{2=\module, 3=\func, 4=\mean, 5=\std}
+\csvstyle{hotspot}{
+	tabular = |r|l|l|c|,
+	table head = \hline & Module & Function & CPU Time (mean$\pm$std)\\\hline\hline,
+	late after line = \\\hline,
+	respect all,
+	csvcol
+}
+\newcommand{\csvtable}[3]{
+	\begin{table}[h]
+		\resizebox*{\textwidth}{!}{
+			\csvreader[hotspot]{#1}{}
+			% \csvreader[hotspot, filter={\value{csvrow}<60}]{#1}{}
+			{\thecsvrow & \module & \func & \tablenum[round-precision=2, round-mode=places]{\mean}$\pm$\tablenum[round-precision=2, round-mode=places]{\std}}
+		}
+		\caption{#2}
+		\label{#3}
+	\end{table}
+}
+
+% CSV tables
+% ANTS brainExtraction
+\csvtable{tables/hotspots-1thread-ants-brainExtraction.csv}{ANTS brainExtraction (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-brainExtraction}
+\csvtable{tables/hotspots-1thread-ants-brainExtraction-fp.csv}{ANTS brainExtraction-fp (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-brainExtraction-fp}
+
+% ANTS registrationSyN
+\csvtable{tables/hotspots-1thread-ants-registrationSyN.csv}{ANTS registrationSyN (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-registrationSyN}
+\csvtable{tables/hotspots-1thread-ants-registrationSyN-fp.csv}{ANTS registrationSyN-fp (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-registrationSyN-fp}
+
+% FreeSurfer recon-all
+\csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
+\csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}
+
+% FSL FAST
+\csvtable{tables/hotspots-1thread-fsl-fast.csv}{FSL FAST (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-fast}
+
+% FSL FLIRT
+\csvtable{tables/hotspots-1thread-fsl-flirt.csv}{FSL FLIRT (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-flirt}
+
+% FSL MCFLIRT
+\csvtable{tables/hotspots-1thread-fsl-mcflirt.csv}{FSL MCFLIRT (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-mcflirt}
+
 \end{document}

### 2023-08-11 paper/vtune-example.sh

@@ -1,14 +1,12 @@
-#!/bin/bash
-
+#!/bin/sh
 singularity exec --cleanenv \
     -B ~/intel/oneapi/vtune/latest/:/vtune \
     -B ${SLURM_TMPDIR}:/data \
-    -B ${PROFILING_DIR}:${PROFILING_DIR} \
+    -B ${PROJECT_DIR}:${PROJECT_DIR} \
     ${SIF_IMG} \
     /vtune/bin64/vtune \
     -collect hpc-performance \
     -knob enable-stack-collection=true \
-    -knob collect-memory-bandwidth=true \
-    -knob analyze-openmp=false \
-    -result-dir ${VTUNE_DIR} \
+    -knob analyze-openmp=true \
+    -result-dir ${PROFILING_DIR} \
     <script.sh>

### 2023-08-11 print_experiments.sh

@@ -3,8 +3,9 @@ set -e
 set -u
 
 NTHREAD=32
+MAX_JOBS=8
 PROJECT_DIR=/mnt/lustre/mathdugre/mri-bottleneck
-SIF_DIR=$HOME/containers
+SIF_DIR=/mnt/lustre/mathdugre/containers
 
 # Datatset preparation
 cat << EOF
@@ -12,11 +13,13 @@ cat << EOF
 # Datatset preparation #
 ########################
 EOF
-DATA_DIR=${PROJECT_DIR}/datasets/LMU_2
-DATALAD_URL="///corr/RawDataBIDS/LMU_2"
+DATA_DIR=${PROJECT_DIR}/datasets/ds004513
+DATALAD_URL="https://github.com/OpenNeuroDatasets/ds004513.git"
 echo "datalad install -gr -J\$(nproc) --source ${DATALAD_URL} ${DATA_DIR}"
+## Convert symlink to hardlink to prevent issue with preprocessing
+echo "find ${DATA_DIR} -type l -exec bash -c 'ln -f \$(readlink -m \$0) \$0' {} \;"
 
-N_SUBJECTS=$(awk 'END {print NR-1}' ${DATA_DIR}/participants.tsv)
+N_SUBJECTS=$(find ${DATA_DIR} -maxdepth 1 -name "sub-*" | wc -l )
 
 # Pipeline profiling
 # PROJECT_DIR=${PROJECT_DIR}/test
@@ -27,22 +30,27 @@ cat << EOF
 ######################
 EOF
 ## Anatomical MRI preprocessing
-echo sbatch --array=1-${N_SUBJECTS}%6 --cpus-per-task=${NTHREAD} sbatch/ants/brainExtraction.sh $SIF_DIR/ants-debug.sif -d ${DATA_DIR} -p ${PROJECT_DIR}
-echo sbatch --array=1-${N_SUBJECTS}%6 --cpus-per-task=${NTHREAD} sbatch/ants/brainExtraction-fp.sh $SIF_DIR/ants-debug.sif -d ${DATA_DIR} -p ${PROJECT_DIR}
[…]

### 2023-08-11 sbatch/afni/3dTshift.sh

@@ -1,0 +1,31 @@
+#!/bin/bash
+
+#SBATCH -J afni-3dTshift
+#SBATCH --array=1
+#SBATCH --time=24:00:00
+#SBATCH --exclusive
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=32
+#SBATCH --mem=0
+# -------------------------
+#SBATCH -o log/%x-%A-%a.out
+#SBATCH -e log/%x-%A-%a.err
+# -------------------------
+set -e
+set -u
+
+# Setup environment and parse args.
+source ./sbatch/pre_run.sh afni 3dTshift -j ${SLURM_CPUS_PER_TASK} $@
+
+cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+3dTshift \
+    -tzero 0 \
+    -quintic \
+    /data/sub-${SUBJECT_ID}/ses-open/func/sub-${SUBJECT_ID}_ses-open_task-rest_bold.nii.gz
+EOT
+
+export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+
+source ./sbatch/post_run.sh

### 2023-08-11 sbatch/fmriprep/anat-fs.sh

@@ -1,0 +1,61 @@
+#!/bin/bash
+
+#SBATCH -J fmriprep-anat-fs
+#SBATCH --array=1
+#SBATCH --time=UNLIMITED
+#SBATCH --exclusive
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem=0
+# -------------------------
+#SBATCH -o log/%x-%A-%a.out
+#SBATCH -e log/%x-%A-%a.err
+# -------------------------
+set -e
+set -u
+
+# Setup environment and parse args.
+source ./sbatch/pre_run.sh fmriprep anat-fs -j ${SLURM_CPUS_PER_TASK} $@
+
+# Create an empty directory for output
+OUTPUT_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-outputs
+SUBJECT_OUTPUT=${OUTPUT_DIR}/sub-${SUBJECT_ID}
+if [ -d "${SUBJECT_OUTPUT}" ]; then rm -Rf ${SUBJECT_OUTPUT}; fi
+mkdir -p ${SUBJECT_OUTPUT}
+
+# Create WORK_DIR for fMRIPrep
+WORK_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-${RANDOM_STRING}
+mkdir -p ${WORK_DIR}
+
+SIF_OPTION+=" -B ${SUBJECT_OUTPUT}:/output"
+SIF_OPTION+=" -B ${WORK_DIR}:/workdir"
+export SIF_OPTION+=" -B $HOME/.freesurfer.txt:/opt/freesurfer/.license"
+cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+FS_LICENSE=/opt/freesurfer/.license
+fmriprep \
+    /data \
+    /output \
+    participant \
+    --work-dir /workdir \
+    --skip_bids_validation \
+    --participant-label ${SUBJECT_ID} \
+    --nthreads ${NTHREAD} \
+    --omp-nthreads ${NTHREAD} \
+    --anat-only
+EOT
+
+export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+
+export SKIP_REPORT=1
[…]

### 2023-08-11 sbatch/fmriprep/anat-nofs.sh

@@ -1,0 +1,62 @@
+#!/bin/bash
+
+#SBATCH -J fmriprep-anat-anat-nofs
+#SBATCH --array=1
+#SBATCH --time=UNLIMITED
+#SBATCH --exclusive
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem=0
+# -------------------------
+#SBATCH -o log/%x-%A-%a.out
+#SBATCH -e log/%x-%A-%a.err
+# -------------------------
+set -e
+set -u
+
+# Setup environment and parse args.
+source ./sbatch/pre_run.sh fmriprep anat-nofs -j ${SLURM_CPUS_PER_TASK} $@
+
+# Create an empty directory for output
+OUTPUT_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-outputs
+SUBJECT_OUTPUT=${OUTPUT_DIR}/sub-${SUBJECT_ID}
+if [ -d "${SUBJECT_OUTPUT}" ]; then rm -Rf ${SUBJECT_OUTPUT}; fi
+mkdir -p ${SUBJECT_OUTPUT}
+
+# Create WORK_DIR for fMRIPrep
+WORK_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-${RANDOM_STRING}
+mkdir -p ${WORK_DIR}
+
+SIF_OPTION+=" -B ${SUBJECT_OUTPUT}:/output"
+SIF_OPTION+=" -B ${WORK_DIR}:/workdir"
+export SIF_OPTION+=" -B $HOME/.freesurfer.txt:/opt/freesurfer/.license"
+cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+FS_LICENSE=/opt/freesurfer/.license
+fmriprep \
+    /data \
+    /output \
+    participant \
+    --work-dir /workdir \
+    --skip_bids_validation \
+    --participant-label ${SUBJECT_ID} \
+    --nthreads ${NTHREAD} \
+    --omp-nthreads ${NTHREAD} \
+    --anat-only \
+    --fs-no-reconall
+EOT
+
+export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+
[…]

### 2023-08-11 sbatch/fmriprep/full.sh

@@ -1,0 +1,59 @@
+#!/bin/bash
+
+#SBATCH -J fmriprep-full
+#SBATCH --array=1
+#SBATCH --time=UNLIMITED
+#SBATCH --exclusive
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem=0
+# -------------------------
+#SBATCH -o log/%x-%A-%a.out
+#SBATCH -e log/%x-%A-%a.err
+# -------------------------
+set -e
+set -u
+
+# Setup environment and parse args.
+source ./sbatch/pre_run.sh fmriprep full -j ${SLURM_CPUS_PER_TASK} $@
+
+# Create an empty directory for output
+OUTPUT_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-outputs
+SUBJECT_OUTPUT=${OUTPUT_DIR}/sub-${SUBJECT_ID}
+if [ -d "${SUBJECT_OUTPUT}" ]; then rm -Rf ${SUBJECT_OUTPUT}; fi
+mkdir -p ${SUBJECT_OUTPUT}
+
+# Create WORK_DIR for fMRIPrep
+WORK_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-${RANDOM_STRING}
+mkdir -p ${WORK_DIR}
+
+SIF_OPTION+=" -B ${SUBJECT_OUTPUT}:/output"
+SIF_OPTION+=" -B ${WORK_DIR}:/workdir"
+export SIF_OPTION+=" -B $HOME/.freesurfer.txt:/opt/freesurfer/.license"
+cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+FS_LICENSE=/opt/freesurfer/.license
+fmriprep \
+    /data \
+    /output \
+    participant \
+    --work-dir /workdir \
+    --skip_bids_validation \
+    --participant-label ${SUBJECT_ID} \
+    --nthreads ${NTHREAD} \
+    --omp-nthreads ${NTHREAD}
+EOT
+
+export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+
+source ./sbatch/post_run.sh
+
+# Extra cleanup
[…]

### 2023-08-11 sbatch/fsl/flirt.sh

@@ -1,0 +1,36 @@
+#!/bin/bash
+
+#SBATCH -J fsl-flirt
+#SBATCH --array=1
+#SBATCH --time=4:00:00
+#SBATCH --exclusive
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem=0
+# -------------------------
+#SBATCH -o log/%x-%A-%a.out
+#SBATCH -e log/%x-%A-%a.err
+# -------------------------
+set -e
+set -u
+
+# Setup environment and parse args.
+source ./sbatch/pre_run.sh fsl flirt -j ${SLURM_CPUS_PER_TASK} $@
+
+export SIF_OPTION="-B $HOME/.cache/templateflow:/templateflow"
+cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+source /fsl_env.sh
+TMPLT="/templateflow/tpl-MNI152NLin2009cAsym/"
+mkdir -p /data/derivatives/fsl/flirt/sub-${SUBJECT_ID}/ses-1/anat/
+
+flirt \
+    -in /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w.nii.gz \
+    -ref \${TMPLT}/tpl-MNI152NLin2009cAsym_res-01_desc-brain_T1w.nii.gz \
+    -out /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w_flirt.nii.gz
+EOT
+
+export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+
+source ./sbatch/post_run.sh

### 2023-08-11 sbatch/fsl/mcflirt.sh

@@ -1,0 +1,32 @@
+#!/bin/bash
+
+#SBATCH -J fsl-mcflirt
+#SBATCH --array=1
+#SBATCH --time=4:00:00
+#SBATCH --exclusive
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem=0
+# -------------------------
+#SBATCH -o log/%x-%A-%a.out
+#SBATCH -e log/%x-%A-%a.err
+# -------------------------
+set -e
+set -u
+
+# Setup environment and parse args.
+source ./sbatch/pre_run.sh fsl mcflirt -j ${SLURM_CPUS_PER_TASK} $@
+
+cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+source /fsl_env.sh
+mkdir -p /data/derivatives/fsl/mcflirt/sub-${SUBJECT_ID}/ses-1/func/
+mcflirt \
+    -o /data/derivatives/fsl/mcflirt/sub-${SUBJECT_ID}/ses-1/func/sub-${SUBJECT_ID}_ses-1_task-rest_run-1_bold_mcf.nii.gz \
+    /data/sub-${SUBJECT_ID}/ses-1/func/sub-${SUBJECT_ID}_ses-1_task-rest_run-1_bold.nii.gz
+EOT
+
+export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
+
+source ./sbatch/post_run.sh

### 2023-08-11 sbatch/post_run.sh

@@ -9,20 +9,19 @@ if [[ ${EXIT_CODE} -ne 0 ]]; then
     exit ${EXIT_CODE}
 fi
 
-# Generate VTune report
-singularity exec --cleanenv \
-    -B ~/intel/oneapi/vtune/latest/:/vtune \
-    -B ${PROJECT_DIR}:${PROJECT_DIR} \
-    ${SIF_IMG} \
-    /vtune/bin64/vtune \
-    -report hotspots \
-    -r ${PROFILING_DIR} \
-    -report-output ${PROFILING_DIR}.csv \
-    -format csv \
-    -csv-delimiter tab \
-    -group-by module,function \
-    -loop-mode function-only
-
+# # Generate VTune report
+# singularity exec --cleanenv \
+#     -B ~/intel/oneapi/vtune/latest/:/vtune \
+#     -B ${PROJECT_DIR}:${PROJECT_DIR} \
+#     ${SIF_IMG} \
+#     /vtune/bin64/vtune \
+#     -report hotspots \
+#     -r ${PROFILING_DIR} \
+#     -report-output ${PROFILING_DIR}.csv \
+#     -format csv \
+#     -csv-delimiter tab \
+#     -group-by module,function \
+#     -loop-mode function-only
 
 # Tarball VTune profiling data to reduce inode usage.
 tar czf ${PROFILING_DIR}.tar.gz -C $(dirname ${PROFILING_DIR}) $(basename ${PROFILING_DIR})

### 2023-08-11 sbatch/pre_run.sh

@@ -109,7 +109,7 @@ export NTHREAD
 export DATASET=$(basename ${DATA_DIR})
 
 [[ -z ${SLURM_ARRAY_TASK_ID:+x} ]] && SLURM_ARRAY_TASK_ID=1
-export SUBJECT_ID=$(sed -n $(( 1 + ${SLURM_ARRAY_TASK_ID} ))p ${DATA_DIR}/participants.tsv | cut -f1)
+export SUBJECT_ID=$(find ${DATA_DIR} -maxdepth 1 -name "sub-*" -exec basename {} \;| sed -n 5p | sed -e "s/^sub-//")
 
 export RANDOM_STRING=$(cat /dev/urandom | tr -dc 'a-zA-Z0-9' | fold -w 32 | head -n 1)
 export PROFILING_DIR=${PROJECT_DIR}/vtune_output/${NTHREAD}-threads/${TOOLKIT}/${PIPELINE}/${DATASET}/sub-${SUBJECT_ID}/${RANDOM_STRING}
@@ -138,6 +138,12 @@ mkdir -p ${SLURM_TMPDIR}/sub-${SUBJECT_ID}
 rsync -aLq --info=progress2 \
     ${DATA_DIR}/sub-${SUBJECT_ID}/ \
     ${SLURM_TMPDIR}/sub-${SUBJECT_ID}/
+# Transfer dataset metadata to compute node.
+count=`ls -1 ${DATA_DIR}/*.tsv 2>/dev/null | wc -l`
+[[ $count != 0 ]] && scp ${DATA_DIR}/*.tsv ${SLURM_TMPDIR}
+
+count=`ls -1 ${DATA_DIR}/*.json 2>/dev/null | wc -l`
+[[ $count != 0 ]] && scp ${DATA_DIR}/*.json ${SLURM_TMPDIR}
 
 # Transfer derivatives for a given subject, excluding the potential cache of
 # current pipeline.

### 2023-08-11 sbatch/vtune.sh

@@ -14,7 +14,9 @@ singularity exec --cleanenv \
     ${SIF_IMG} \
     /vtune/bin64/vtune \
     -collect hpc-performance \
+    -no-auto-finalize \
+    -data-limit=0 \
     -knob enable-stack-collection=true \
-    -knob analyze-openmp=false \
+    -knob analyze-openmp=true \
     -result-dir ${PROFILING_DIR} \
     $@

### 2023-09-06 paper/paper.tex
First comments

@@
 
-\newcommand{\TG}[1]{\color{red}\textsc{From Tristan: }#1\color{black}}
-\newcommand{\MD}[1]{\color{magenta}\textsc{From Tristan: }#1\color{black}}
+\newcommand{\TG}[1]{\color{blue}\textsc{From Tristan: }#1\color{black}}
+\newcommand{\MD}[1]{\color{magenta}\textsc{From Mathieu: }#1\color{black}}
 \newcommand{\HL}[1]{\hl{#1}}
@@
 
-\author{Mathieu Dugr\'e, Tristan Glatard}
+\author{Mathieu Dugr\'e, Tristan Glatard \TG{I think Yohan should be co-author given his technical input}}
 
 \begin{document}
+
 \maketitle
@@
 	% TODO																
-
 \end{abstract}
@@
 % **Why:** Performance of neuroimaging pipeline is critical (larger study, clinical use). However, limited study on performance profile.
+
+\TG{Good intro but I think it would deserve a better logic. You could start by explaining that computational bottlenecks 
+have been one of the main bottlenecks to the processing of large cohorts in neuroimaging, that the community is constantly searching for ways to process data faster (HPC, cloud, GPUs, DL models), and that your paper takes a step back by analyzing the computational bottlenecks in order to subsequently focus on their optimization.
+}
+
 Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain.
 Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI).
-While other modalities such as EEG, CT, and PET exist, we focus on MRIs for their broad adoption and non-invasive property.
+While other modalities such as EEG, CT, and PET exist \TG{define acronyms}, we focus on MRIs for their broad adoption and non-invasive property.
 However, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.
@@
 
+\TG{you could summarize the method (vtune profiler) in the intro.}
+
 \section{Methods}
@@
 The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-240.1.1.el8\_lustre.x86\_64}.
+\TG{I would defer this subsection to later in the methods as it's not the most central one}
 
 \subsection{Data}
-To ensure the performance profiles of the applications derived from our work are inclusive of different population, we used data with a wide range of age and equal distribution of sex (biological).
+To ensure the performance profiles of the applications derived from our work are inclusive of different populations, we used data with a wide range of age and equal distribution of sex (biological).
 Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.
@@
 The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts.
-The within-subject cohort has nine participants (mean age=43, std=7; 4 females) with two session: eyes open and eyes closed.
-The replication cohort has eleven participants (mean age=27, std=5; 6 females) with only the eyes open session.
+The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed.
+The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session.
 We profiled the application using both cohort on the eyes open session, unless explicitly specified.
@@
 First, we profile these commonly accepted workflows: fMRIPrep anatomical only without and without FreeSurfer, fMRIPrep complete workflow, and dMRIPrep.
-This let us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.
+This lets us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.
 
@@
 Second, to estimate an average performance profile, we aggregate the profiling results of each pipeline across all subjects.
-Combining the Pareto's law and VTune hotspots report, we present the functions accounting for most of the CPU time up to 80\% of the total cpu time of the application.
+Combining the Pareto's law and VTune hotspots report \TG{It would be useful to define these terms}, we present the functions accounting for most of the CPU time up to 80\% of the total CPU time of the application \TG{check wording}.
 This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
@@
 \section{Results}
+\TG{Interesting results but presentation needs to be worked out.
+This section should be structured with sub-headings summarizing the main findings, for instance: "Interpolation is the main bottleneck", etc}
+
 % TODO
@@
 % This will let us decided which pipeline to investigate more.
-Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan, in seconds, while using 1 thread for one execution of all subjects.
+Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan \TG{define makespan}, in seconds, while using 1 thread for one execution of all subjects.
 Unexpectedly, we observe that both ANTS~BrainExtration and ANTS~RegistrationSyN are slower when using single precision compared to double precision.
@@
 	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
-	\caption{Profiling of the pipelines using 1 thread.}
+	\caption{Profiling of the pipelines using 1 thread. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
 	\label{fig:makespan-1thread}
@@
 \subsection{Hotspots}
-Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspot observed from profiling the fMRIPrep anatomical sub-workflows.
+
+Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspots observed from profiling the fMRIPrep anatomical sub-workflows.
 In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module.
@@
 
+\TG{I'd move the graphical explanations to the figure caption so that the text can focus on observations}
+
 In figure~\ref{subfig:hotspots-ants-brainExtraction}, we observe that the function 1 and 2 account for a significant portion of the application execution time.
@@
 
-Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two function accounting for the most time, although in flipped order.
+Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two functions accounting for the most time, although in flipped order.
 Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time.
@@
 	\label{fig:hotspots-fmriprep-anat-subworkflow}
+
 	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
@@
 		\label{subfig:hotspots-fsl-flirt}
+
 		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-flirt.png}
 	\end{subfigure}
-
 	\caption{Hotspots profiling of fMRIPrep's functional sub-workflows.}
 	\label{fig:hotspots-fmriprep-func-subworkflow}
+
 	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.

### 2023-10-25 paper/paper.tex
Updated writing.

--- 2023-08-11 paper/paper.tex
@@
 \usepackage{csvsimple}
+\usepackage{booktabs}
 
@@
 \subsubsection{Intensity correction}
-% NU4 bias correction (Check ANTS brainExtraction)
+% N4 bias correction (Check ANTS brainExtraction)
+% https://ieeexplore.ieee.org/document/5445030
 
@@
 
+\subsubsection{Spatial Normalization}
+% ANTS registrationSyN
+% SyN Registration: https://pubmed.ncbi.nlm.nih.gov/17659998/
+
 \subsubsection{Skull-Stripping}
 % ANTS brainExtraction
-
-\subsubsection{Spatial Normalization}
-% ANTS registrationSyN
+% 1. Truncate image intensity
+% 2. N4 bias correction
+% 3. SyN Registration
+% 4. Warp prior to image using affine
+% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/
 
@@
 \subsection{dMRI Pre-Processing}
-\subsubsection{Eddy Currents Distortion Estimation}
-% FSL eddy
+% TODO: Decide which pipelines to use
+% \subsubsection{Eddy Currents Distortion Estimation}
+% % FSL eddy
 
@@
 \section{Results}
+\subsection{Makespan}
+Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan, in seconds, while using 1 thread for one execution of all subjects.
+We observe significant computation requirements between the pipelines.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
+	\caption{Profiling of the pipelines using 1 thread.}
+	\label{fig:makespan-1thread}
+\end{figure}
+
+\subsection{Long-Tail Distribution}
+Figure~\ref{fig:longtail} depicts the distribution of average CPU time for each function across all pipelines. We observe a long tail behavior for the execution time. 
+% TODO add statistics about the long tail distribution
+This confirms that only a small fraction of the functions contribute to a majority of the execution time of MRI preprocessing pipelines.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\linewidth]{figures/global_longtail.png}
+	\caption{Distribution of CPU Time for each function across all pipelines.}
+	\label{fig:longtail}
+\end{figure}
+
+\subsection{Linear Interpolation}
+% TODO
+% - Hotspot analysis for ANTS brainExtraction, ANTS registerSyN, and FreeSurfer recon-all
+% Interpolation plays major role for volumetric. Mention top function in ANTS and FreeSurfer.
+% - What is the cumulative contribution of interpolation in both applications?
+% -
+
+\subsection{FreeSurfer parallelism}
+% TODO
+% - Add results from dummy OMP dynamic scheduling.
+% Might want to add those to the discussion instead.
+Our results showed poor parallelism while using FreeSurfer with 32 threads.
+Therefore, we expanded our methods to execute FreeSurfer recon-all with 1, 2, 4, 8, 16, and 32 threads.
+This allow us to study the multi-threading performance of the application.
+
+Figure~\ref{fig:freesurfer-parallelism} shows the makespan (blue) and parallelism efficiency (red) while varying the allocated number of threads for FreeSurfer.
+We set the number of threads using the FreeSurfer ``-threads'' flag and the ``OMP\_NUM\_THREADS'' environment variable.
+While the makespan reduce with the number of threads, the application shows poor parallelism performance.
+Table~\ref{tab:freesurfer-parallelism} shows decreasing parallelism efficiency from 56.16 to 4.93, for 2 to 32 threads respectively.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\linewidth]{figures/makespan-freesurfer.png}
+	\caption{FreeSurfer: Makespan (Blue) and parallelism efficiency (Red) while varying the number of threads.}
+	\label{fig:freesurfer-parallelism}
+\end{figure}
+\begin{table}[h]
+	\centering
+	\begin{tabular}{@{}lcccccc@{}}
+		\toprule
+		\# of Threads       & \multicolumn{1}{l}{1} & \multicolumn{1}{l}{2} & \multicolumn{1}{l}{4} & \multicolumn{1}{l}{8} & \multicolumn{1}{l}{16} & \multicolumn{1}{l}{32} \\ \midrule
+		Speedup             & -                     & 1.12                  & 1.33                  & 1.47                  & 1.58                   & 1.58                   \\
+		Parallel Efficiency & -                     & 56.16                 & 33.17                 & 18.33                 & 9.88                   & 4.93                   \\ \bottomrule
+	\end{tabular}
+	\caption{FreeSurfer: Makespan and parallelism efficiency while varying the number of threads}
+	\label{tab:freesurfer-parallelism}
+\end{table}
+
+The hotspot analysis for FreeSurfer recon-all with 1 and 32 threads are shown in Figure~\ref{fig:hotspots-freesurfer-reconall}.
+We observe a clear difference in the profiling between the sequential and multi-threaded executions of the pipeline.
+On the one hand, figure~\ref{subfig:hotspots-1thread-freesurfer-reconall} shows that the sequential execution has dozens of functions and modules contributing to 80\% application execution time.
+On the other hand, figure~\ref{subfig:hotspots-32threads-freesurfer-reconall} shows that the multi-threaded execution only has 2 modules and 3 functions contributing a majority of the application runtime.
+
+The primary bottleneck in multi-threaded FreeSurfer is the OMP library, which awaits thread synchronization throughout the pipeline.
+This accounts for at least 76\% of the application runtime.
+We think this is due to imbalance in the task scheduling and suppose that optimizing the OMP scheduling could significantly benefit FreeSurfer.
+\begin{figure}[ht!]
+	\centering
+	% FreeSurfer recon-all
+	\begin{subfigure}[t]{\linewidth}
+		\caption{}
+		\label{subfig:hotspots-1thread-freesurfer-reconall}
+		\includegraphics[width=\linewidth]{figures/hotspots-1thread-freesurfer-reconall.png}
+	\end{subfigure}
+	\begin{subfigure}[t]{\linewidth}
+		\caption{}
+		\label{subfig:hotspots-32threads-freesurfer-reconall}
+		\includegraphics[width=\linewidth]{figures/hotspots-32threads-freesurfer-reconall.png}
+	\end{subfigure}
+	\caption{FreeSurfer recon-all: Hotspot analysis for 1 and 32 threads.}
+	\label{fig:hotspots-freesurfer-reconall}
+\end{figure}
+
+FreeSurfer plays an integral role in the fMRIPrep workflow.
+Therefore, the OMP bottleneck shown previously are also found in multi-threaded fMRIPrep, as Figure~\ref{fig:hotspots-fmriprep-32threads} depicts.
+\begin{figure}[ht!]
+	\centering
+	\includegraphics[width=\linewidth]{figures/hotspots-32threads-fmriprep-full.png}
+	\caption{fMRIPrep (32 threads): Hotspot analysis.}
+	\label{fig:hotspots-fmriprep-32threads}
+\end{figure}
+
+\subsection{ANTS single vs. double precision}
+% TODO
+% - Create makespan plot with only ANTS pipeline
+% There's no need to show the hotspot analysis as there are no differences between single and double precision.
+% - Results from toy example with single/double precision in ITK
+\begin{figure*}[ht!]
+	\centering
+	% ANTS brainExtraction
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-ants-brainExtraction}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-ants-brainExtraction-fp}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
+	\end{subfigure}
+	% ANTS registrationSyN
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-ants-registrationSyN}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[t]{0.49\textwidth}
+		\caption{}
+		\label{subfig:hotspots-ants-registrationSyN-fp}
+		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
+	\end{subfigure}
+	\caption{ANTS: Single vs. Double precision}
+	\label{fig:ants-fp-comparison}
+\end{figure*}
+
+\section{old Results}
 % TODO
@@
 Second, we depict detailed analysis for the key functions contributing to the CPU time of each application (hotspots).
-
-\subsection{Makespan}
-% Identify which tools take the longest to execute.
-% This will let us decided which pipeline to investigate more.
-Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan, in seconds, while using 1 thread for one execution of all subjects.
-Unexpectedly, we observe that both ANTS~BrainExtration and ANTS~RegistrationSyN are slower when using single precision compared to double precision.
-% TODO Stat test between mean for single and double precision.
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
-	\caption{Profiling of the pipelines using 1 thread.}
-	\label{fig:makespan-1thread}
-\end{figure}
 
@@
 % TODO FreeSurfer 1-thread sub-figure
-
-In figure~\ref{subfig:hotspots-32threads-freesurfer-reconall}, we observe that OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime.
-The only function shown from FreeSurfer recon-all is the same as the first function from figure~\ref{subfig:hotspots-1thread-freesurfer-reconall}.
-However, it only accounts for 2\% of the total makespan compared to 3.8\% while running sequentially.
+% Note:
+% - Way more functions
+% - Top 5 accounts for only 15% of total time. Much less than other applications.
+% - Way more memory bound that previous application.
+% - Focusing on Top 5 and memory bound could give good performance boost.
 
@@
 	\centering
-	% ANTS brainExtraction
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-ants-brainExtraction}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-ants-brainExtraction-fp}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
-	\end{subfigure}
+	% % ANTS brainExtraction
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-ants-brainExtraction}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
+	% \end{subfigure}
+	% \hfill
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-ants-brainExtraction-fp}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
+	% \end{subfigure}
 
-	% ANTS registrationSyN
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-ants-registrationSyN}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-ants-registrationSyN-fp}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
-	\end{subfigure}
+	% % ANTS registrationSyN
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-ants-registrationSyN}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
+	% \end{subfigure}
+	% \hfill
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-ants-registrationSyN-fp}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
+	% \end{subfigure}
 
-	% FreeSurfer recon-all
-	\begin{subfigure}[t]{\textwidth}
-		\caption{}
-		\label{subfig:hotspots-1thread-freesurfer-reconall}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-freesurfer-reconall.png}
-	\end{subfigure}
+	% % FreeSurfer recon-all
+	% \begin{subfigure}[t]{\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-1thread-freesurfer-reconall}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-freesurfer-reconall.png}
+	% \end{subfigure}
 
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-32threads-freesurfer-reconall}
-		\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall.png}
-	\end{subfigure}
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-32threads-freesurfer-reconall}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall.png}
+	% \end{subfigure}
 	\hfill
@@
 % FreeSurfer recon-all
-\csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
-\csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}
+% \csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
+% \csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}
 

### 2023-11-30 paper/paper.tex
latest results.

@@
 For our experiments, we used the \textit{Slashbin cluster} at Concordia University.
-The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-240.1.1.el8\_lustre.x86\_64}.
+The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}.
 

### 2023-11-30 print_experiments.sh

@@ -2,7 +2,7 @@
 set -e
 set -u
 
-NTHREAD=32
+NTHREAD=1
 MAX_JOBS=8
 PROJECT_DIR=/mnt/lustre/mathdugre/mri-bottleneck
 SIF_DIR=/mnt/lustre/mathdugre/containers
@@ -30,27 +30,31 @@ cat << EOF
 ######################
 EOF
 ## Anatomical MRI preprocessing
-echo sbatch --array=1-${N_SUBJECTS}%${MAX_JOBS} --cpus-per-task=${NTHREAD} sbatch/ants/brainExtraction.sh $SIF_DIR/ants-debug.sif -d ${DATA_DIR} -p ${PROJECT_DIR}
-echo sbatch --array=1-${N_SUBJECTS}%${MAX_JOBS} --cpus-per-task=${NTHREAD} sbatch/ants/brainExtraction-fp.sh $SIF_DIR/ants-debug.sif -d ${DATA_DIR} -p ${PROJECT_DIR}
+echo sbatch --array=1-${N_SUBJECTS}%${MAX_JOBS} --cpus-per-task=${NTHREAD} sbatch/ants/brainExtraction.sh $SIF_DIR/ants-debug.simg -d ${DATA_DIR} -p ${PROJECT_DIR}
+echo sbatch --array=1-${N_SUBJECTS}%${MAX_JOBS} --cpus-per-task=${NTHREAD} sbatch/ants/brainExtraction-fp.sh $SIF_DIR/ants-debug.simg -d ${DATA_DIR} -p ${PROJECT_DIR}
 
-echo sbatch --array=1-${N_SUBJECTS}%${MAX_JOBS} --cpus-per-task=${NTHREAD} sbatch/fsl/fast.sh $SIF_DIR/fsl-debug.sif -d ${DATA_DIR} -p ${PROJECT_DIR}
+echo sbatch --array=1-${N_SUBJECTS}%${MAX_JOBS} --cpus-per-task=${NTHREAD} sbatch/fsl/fast.sh $SIF_DIR/fsl-debug.simg -d ${DATA_DIR} -p ${PROJECT_DIR}
 
-echo sbatch --array=1-${N_SUBJECTS}%${MAX_JOBS} --cpus-per-task=${NTHREAD} sbatch/ants/registrationSyN.sh $SIF_DIR/ants-debug.sif -d ${DATA_DIR} -p ${PROJECT_DIR}
[…]

### 2023-11-30 sbatch/ants/brainExtraction-fp.sh

@@ -24,13 +24,13 @@ TMPLT="/opt/templates/OASIS"
 antsBrainExtraction.sh \
     -q 1 \
     -d 3 \
-    -a /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w.nii.gz \
+    -a /data/sub-${SUBJECT_ID}/ses-open/anat/sub-${SUBJECT_ID}_ses-open_T1w.nii.gz \
     -e \${TMPLT}/T_template0.nii.gz \
     -m \${TMPLT}/T_template0_BrainCerebellumProbabilityMask.nii.gz \
-    -o /data/derivatives/ants/brainExtraction-fp/sub-${SUBJECT_ID}/ses-1/anat/
+    -o /data/derivatives/ants/brainExtraction-fp/sub-${SUBJECT_ID}/ses-open/anat/
 EOT
 
-export SINGULARITYENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
+export APPTAINERENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2023-11-30 sbatch/ants/brainExtraction.sh

@@ -23,13 +23,13 @@ TMPLT="/opt/templates/OASIS"
 
 antsBrainExtraction.sh \
     -d 3 \
-    -a /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w.nii.gz \
+    -a /data/sub-${SUBJECT_ID}/ses-open/anat/sub-${SUBJECT_ID}_ses-open_T1w.nii.gz \
     -e \${TMPLT}/T_template0.nii.gz \
     -m \${TMPLT}/T_template0_BrainCerebellumProbabilityMask.nii.gz \
-    -o /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-1/anat/
+    -o /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-open/anat/
 EOT
 
-export SINGULARITYENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
+export APPTAINERENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2023-11-30 sbatch/ants/registrationSyN-fp.sh

@@ -25,12 +25,12 @@ TMPLT="/templateflow/tpl-MNI152NLin2009cAsym/"
 antsRegistrationSyN.sh \
     -d 3 \
     -f \${TMPLT}/tpl-MNI152NLin2009cAsym_res-01_desc-brain_T1w.nii.gz\
-    -m /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain_seg.nii.gz \
-    -o /data/derivatives/ants/registrationSyN-fp/sub-${SUBJECT_ID}/ses-1/anat/ \
+    -m /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-open/anat/BrainExtractionBrain_seg.nii.gz \
+    -o /data/derivatives/ants/registrationSyN-fp/sub-${SUBJECT_ID}/ses-open/anat/ \
     -p f
 EOT
 
-export SINGULARITYENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
+export APPTAINERENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2023-11-30 sbatch/ants/registrationSyN.sh

@@ -25,11 +25,11 @@ TMPLT="/templateflow/tpl-MNI152NLin2009cAsym/"
 antsRegistrationSyN.sh \
     -d 3 \
     -f \${TMPLT}/tpl-MNI152NLin2009cAsym_res-01_desc-brain_T1w.nii.gz\
-    -m /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain_seg.nii.gz \
-    -o /data/derivatives/ants/registrationSyN/sub-${SUBJECT_ID}/ses-1/anat/
+    -m /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-open/anat/BrainExtractionBrain_seg.nii.gz \
+    -o /data/derivatives/ants/registrationSyN/sub-${SUBJECT_ID}/ses-open/anat/
 EOT
 
-export SINGULARITYENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
+export APPTAINERENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2023-11-30 sbatch/freesurfer/reconall.sh

@@ -28,7 +28,7 @@ mkdir -p /data/derivatives/freesurfer/reconall
 recon-all \
     -sd /data/derivatives/freesurfer/reconall \
     -subjid sub-${SUBJECT_ID} \
-    -i /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w.nii.gz \
+    -i /data/sub-${SUBJECT_ID}/ses-open/anat/sub-${SUBJECT_ID}_ses-open_T1w.nii.gz \
     -autorecon1 \
     -noskullstrip -noT2pial -noFLAIRpial \
     -threads ${NTHREAD}
@@ -36,7 +36,7 @@ recon-all \
 # Step 2
 python3 ./sbatch/freesurfer/skull_strip_extern.py \
     /data/derivatives/freesurfer/reconall/sub-${SUBJECT_ID}/mri \
-    /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain.nii.gz
+    /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-open/anat/BrainExtractionBrain.nii.gz
 
 # Step 3
 recon-all \
@@ -83,8 +83,8 @@ recon-all \
     -threads ${NTHREAD}
 EOT
 
-export SINGULARITYENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
-export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+export APPTAINERENV_ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS=$NTHREAD
+export APPTAINERENV_OMP_NUM_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2023-11-30 sbatch/fsl/fast.sh

@@ -20,13 +20,13 @@ source ./sbatch/pre_run.sh fsl fast -j ${SLURM_CPUS_PER_TASK} $@
 
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 source /fsl_env.sh
-mkdir -p /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/
+mkdir -p /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-open/anat/
 fast \
-    -o /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain \
-    /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-1/anat/BrainExtractionBrain.nii.gz
+    -o /data/derivatives/fsl/fast/sub-${SUBJECT_ID}/ses-open/anat/BrainExtractionBrain \
+    /data/derivatives/ants/brainExtraction/sub-${SUBJECT_ID}/ses-open/anat/BrainExtractionBrain.nii.gz
 EOT
 
-export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+export APPTAINERENV_OMP_NUM_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2023-11-30 sbatch/fsl/flirt.sh

@@ -22,15 +22,15 @@ export SIF_OPTION="-B $HOME/.cache/templateflow:/templateflow"
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 source /fsl_env.sh
 TMPLT="/templateflow/tpl-MNI152NLin2009cAsym/"
-mkdir -p /data/derivatives/fsl/flirt/sub-${SUBJECT_ID}/ses-1/anat/
+mkdir -p /data/derivatives/fsl/flirt/sub-${SUBJECT_ID}/ses-open/anat/
 
 flirt \
-    -in /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w.nii.gz \
+    -in /data/sub-${SUBJECT_ID}/ses-open/anat/sub-${SUBJECT_ID}_ses-open_T1w.nii.gz \
     -ref \${TMPLT}/tpl-MNI152NLin2009cAsym_res-01_desc-brain_T1w.nii.gz \
-    -out /data/sub-${SUBJECT_ID}/ses-1/anat/sub-${SUBJECT_ID}_ses-1_run-1_T1w_flirt.nii.gz
+    -out /data/sub-${SUBJECT_ID}/ses-open/anat/sub-${SUBJECT_ID}_ses-open_T1w_flirt.nii.gz
 EOT
 
-export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+export APPTAINERENV_OMP_NUM_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2023-11-30 sbatch/fsl/mcflirt.sh

@@ -20,13 +20,13 @@ source ./sbatch/pre_run.sh fsl mcflirt -j ${SLURM_CPUS_PER_TASK} $@
 
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 source /fsl_env.sh
-mkdir -p /data/derivatives/fsl/mcflirt/sub-${SUBJECT_ID}/ses-1/func/
+mkdir -p /data/derivatives/fsl/mcflirt/sub-${SUBJECT_ID}/ses-open/func/
 mcflirt \
-    -o /data/derivatives/fsl/mcflirt/sub-${SUBJECT_ID}/ses-1/func/sub-${SUBJECT_ID}_ses-1_task-rest_run-1_bold_mcf.nii.gz \
-    /data/sub-${SUBJECT_ID}/ses-1/func/sub-${SUBJECT_ID}_ses-1_task-rest_run-1_bold.nii.gz
+    -o /data/derivatives/fsl/mcflirt/sub-${SUBJECT_ID}/ses-open/func/sub-${SUBJECT_ID}_ses-open_task-rest_bold_mcf.nii.gz \
+    /data/sub-${SUBJECT_ID}/ses-open/func/sub-${SUBJECT_ID}_ses-open_task-rest_bold.nii.gz
 EOT
 
-export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+export APPTAINERENV_OMP_NUM_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2023-11-30 sbatch/post_run.sh

@@ -9,19 +9,30 @@ if [[ ${EXIT_CODE} -ne 0 ]]; then
     exit ${EXIT_CODE}
 fi
 
-# # Generate VTune report
-# singularity exec --cleanenv \
-#     -B ~/intel/oneapi/vtune/latest/:/vtune \
-#     -B ${PROJECT_DIR}:${PROJECT_DIR} \
-#     ${SIF_IMG} \
-#     /vtune/bin64/vtune \
-#     -report hotspots \
-#     -r ${PROFILING_DIR} \
-#     -report-output ${PROFILING_DIR}.csv \
-#     -format csv \
-#     -csv-delimiter tab \
-#     -group-by module,function \
-#     -loop-mode function-only
+# Generate VTune report
+singularity exec --cleanenv \
+    -B ~/intel/oneapi/vtune/latest/:/vtune \
+    -B ${PROJECT_DIR}:${PROJECT_DIR} \
+    ${SIF_IMG} \
+    /vtune/bin64/vtune \
+    -report hotspots \
+    -r ${PROFILING_DIR} \
+    -report-output ${PROFILING_DIR}.csv \
+    -format csv \
+    -csv-delimiter tab \
+    -group-by module,function \
+    -loop-mode function-only
+
+singularity exec --cleanenv \
+    -B ~/intel/oneapi/vtune/latest/:/vtune \
+    -B ${PROJECT_DIR}:${PROJECT_DIR} \
+    ${SIF_IMG} \
+    /vtune/bin64/vtune \
+    -report summary \
+    -r ${PROFILING_DIR} \
+    -report-output ${PROFILING_DIR}.makespan \
+    -format csv \
+    -csv-delimiter tab
 
 # Tarball VTune profiling data to reduce inode usage.
 tar czf ${PROFILING_DIR}.tar.gz -C $(dirname ${PROFILING_DIR}) $(basename ${PROFILING_DIR})
@@ -29,7 +40,6 @@ rm -r ${PROFILING_DIR}
 
 # Transfer pipeline output to storage node.
 rsync -aq --info=progress2 ${SLURM_TMPDIR}/ ${DATA_DIR}/
-rm -rf ${SLURM_TMPDIR}
 
 # Delete temporary script
 rm ${TMP_SCRIPT}/${RANDOM_STRING}.sh

### 2023-11-30 sbatch/pre_run.sh

@@ -109,7 +109,7 @@ export NTHREAD
 export DATASET=$(basename ${DATA_DIR})
 
 [[ -z ${SLURM_ARRAY_TASK_ID:+x} ]] && SLURM_ARRAY_TASK_ID=1
-export SUBJECT_ID=$(find ${DATA_DIR} -maxdepth 1 -name "sub-*" -exec basename {} \;| sed -n 5p | sed -e "s/^sub-//")
+export SUBJECT_ID=$(find ${DATA_DIR} -maxdepth 1 -name "sub-*" -exec basename {} \;| sed -n ${SLURM_ARRAY_TASK_ID}p | sed -e "s/^sub-//")
 
 export RANDOM_STRING=$(cat /dev/urandom | tr -dc 'a-zA-Z0-9' | fold -w 32 | head -n 1)
 export PROFILING_DIR=${PROJECT_DIR}/vtune_output/${NTHREAD}-threads/${TOOLKIT}/${PIPELINE}/${DATASET}/sub-${SUBJECT_ID}/${RANDOM_STRING}
@@ -122,6 +122,7 @@ NTHREAD: ${NTHREAD}
 
 DATASET: ${DATASET}
 SUBJECT_ID: ${SUBJECT_ID}
+SLURM_ARRAY_TASK_ID: ${SLURM_ARRAY_TASK_ID}
 
 SIF_IMG: ${SIF_IMG}
 DATA_DIR: ${DATA_DIR}
@@ -133,7 +134,6 @@ PROFILING_DIR: ${PROFILING_DIR}
 mkdir -p ${PROJECT_DIR}
 
 # Transfer dataset to compute node.
-rm -rf ${SLURM_TMPDIR}
 mkdir -p ${SLURM_TMPDIR}/sub-${SUBJECT_ID}
 rsync -aLq --info=progress2 \
     ${DATA_DIR}/sub-${SUBJECT_ID}/ \
@@ -157,6 +157,8 @@ rsync -aLq \
     ${DATA_DIR}/derivatives/ \
     ${SLURM_TMPDIR}/derivatives/
 
+# Prevent data caching
+rm -rf ${SLURM_TMPDIR}/derivatives/${TOOLKIT}/${PIPELINE}/sub-${SUBJECT_ID}
 
 # Create temp directory for pipeline scripts.
 export TMP_SCRIPT=tmp-scripts

### 2023-11-30 sbatch/vtune.sh

@@ -6,7 +6,7 @@ echo "Command to profile: $@"
 
 [[ -z ${SIF_OPTION:+x} ]] && SIF_OPTION=""
 
-singularity exec --cleanenv \
+apptainer exec --cleanenv \
     -B ~/intel/oneapi/vtune/latest/:/vtune \
     -B ${SLURM_TMPDIR}:/data \
     -B ${PROJECT_DIR}:${PROJECT_DIR} \
@@ -14,7 +14,6 @@ singularity exec --cleanenv \
     ${SIF_IMG} \
     /vtune/bin64/vtune \
     -collect hpc-performance \
-    -no-auto-finalize \
     -data-limit=0 \
     -knob enable-stack-collection=true \
     -knob analyze-openmp=true \

### 2023-11-30 paper/paper.tex

--- 2023-09-06 paper/paper.tex
@@
 For our experiments, we used the \textit{Slashbin cluster} at Concordia University.
-The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-240.1.1.el8\_lustre.x86\_64}.
+The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}.
 \TG{I would defer this subsection to later in the methods as it's not the most central one}
@@
 	\end{subfigure}
-
 	% ANTS registrationSyN
@@
 	\end{subfigure}
+	\caption{ANTS: Single vs. Double precision}
+	\label{fig:ants-fp-comparison}
+\end{figure*}
 
-	% FreeSurfer recon-all
-	\begin{subfigure}[t]{\textwidth}
-		\caption{}
-		\label{subfig:hotspots-1thread-freesurfer-reconall}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-freesurfer-reconall.png}
-	\end{subfigure}
+\section{old Results}
+% TODO
+% 2-step approach:
+% 1. Which pipeline constitute the long execution time in preprocessing?
+% 2. For those, which functions take the most time to execute? (hotspots)
+% Any key finding: common point between functions, mem bound, fops, etc.
 
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-32threads-freesurfer-reconall}
-		\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall.png}
-	\end{subfigure}
+% In this section, analyze the aggregated profiling data obtained with VTune.
+First, a coarse summary of the workflows and pipelines is presented using their makespan.
+Second, we depict detailed analysis for the key functions contributing to the CPU time of each application (hotspots).
+
+\subsection{Hotspots}
+Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspot observed from profiling the fMRIPrep anatomical sub-workflows.
+In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module.
+The function ID are found in the supplementary material Section~\ref{sec:supplementary}.
+The bars show the average CPU time in seconds of a function and the error bars indicate the standard deviation, with the left y-axis showing the scale.
+The light-green dots represent the functions cumulative percentage makespan of the application, with the right y-axis showing the scale.
+The color of the bar indicate the percentage of calls, for the function, that were ``stalled due to demand memory loads and stores''\footnote{\href{https://www.intel.com/content/www/us/en/docs/vtune-profiler/user-guide/2023-0/cpu-metrics-reference.html\#MEMORY-BOUND}{Vtune Memory Bound metric}}, with the left color bar showing the scale.
+
+In figure~\ref{subfig:hotspots-ants-brainExtraction}, we observe that the function 1 and 2 account for a significant portion of the application execution time.
+Moreover, calls from function~1 are memory bound $51.84\pm{4.55}$ percent of the time.
+Table~\ref{extab:hotspots-1thread-ants-brainExtraction} shows that both functions are part of the ITK linear interpolation module.
+
+Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two function accounting for the most time, although in flipped order.
+Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time.
+Compared to the double precision version of ANTS brainExtraction, the third function runtime is significantly larger.
+It is an ITK function transforming points while calculating the displacement field of the image.
+
+Figure~\ref{subfig:hotspots-ants-registrationSyN} and figure~\ref{subfig:hotspots-ants-registrationSyN-fp} show that the execution time for ANTS registrationSyN is dominated by a single function.
+In both case, the ITK function evaluating point during linear interpolation.
+The next two function are also shared between the application, although in flipped order.
+These functions show some memory bound in the range of 16--27.5.
+We also note the three most expensive functions are the same as in both versions of ANTS brainExtraction.
+
+% TODO FreeSurfer 1-thread sub-figure
+% Note:
+% - Way more functions
+% - Top 5 accounts for only 15% of total time. Much less than other applications.
+% - Way more memory bound that previous application.
+% - Focusing on Top 5 and memory bound could give good performance boost.
+
+\begin{figure*}[ht!]
+	\centering
+	% % ANTS brainExtraction
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-ants-brainExtraction}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
+	% \end{subfigure}
+	% \hfill
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-ants-brainExtraction-fp}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
+	% \end{subfigure}
+
+	% % ANTS registrationSyN
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-ants-registrationSyN}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
+	% \end{subfigure}
+	% \hfill
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-ants-registrationSyN-fp}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
+	% \end{subfigure}
+
+	% % FreeSurfer recon-all
+	% \begin{subfigure}[t]{\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-1thread-freesurfer-reconall}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-freesurfer-reconall.png}
+	% \end{subfigure}
+
+	% \begin{subfigure}[t]{0.49\textwidth}
+	% 	\caption{}
+	% 	\label{subfig:hotspots-32threads-freesurfer-reconall}
+	% 	\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall.png}
+	% \end{subfigure}
 	\hfill

### 2024-02-13 paper/fmriprep-citations/CITATION.tex
[WIP]

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

Results included in this manuscript come from preprocessing performed
using \emph{fMRIPrep} 23.0.2 (\citet{fmriprep1}; \citet{fmriprep2};
RRID:SCR\_016216), which is based on \emph{Nipype} 1.8.6
(\citet{nipype1}; \citet{nipype2}; RRID:SCR\_002502).

\begin{description}
\item[Anatomical data preprocessing]
A total of 1 T1-weighted (T1w) images were found within the input BIDS
dataset.The T1-weighted (T1w) image was corrected for intensity
non-uniformity (INU) with \texttt{N4BiasFieldCorrection} \citep{n4},
distributed with ANTs 2.4.2 \citep[RRID:SCR\_004757]{ants}, and used as
T1w-reference throughout the workflow. The T1w-reference was then
skull-stripped with a \emph{Nipype} implementation of the
\texttt{antsBrainExtraction.sh} workflow (from ANTs), using OASIS30ANTs
as target template. Brain tissue segmentation of cerebrospinal fluid
(CSF), white-matter (WM) and gray-matter (GM) was performed on the
brain-extracted T1w using \texttt{fast} \citep[FSL 6.0.6.4,
RRID:SCR\_002823,][]{fsl_fast}. Brain surfaces were reconstructed using
\texttt{recon-all} \citep[FreeSurfer 7.3.3-20231213-debug,
RRID:SCR\_001847,][]{fs_reconall}, and the brain mask estimated
previously was refined with a custom variation of the method to
reconcile ANTs-derived and FreeSurfer-derived segmentations of the
cortical gray-matter of Mindboggle
\citep[RRID:SCR\_002438,][]{mindboggle}. Volume-based spatial
normalization to one standard space (MNI152NLin2009cAsym) was performed
through nonlinear registration with \texttt{antsRegistration} (ANTs
2.4.2), using brain-extracted versions of both T1w reference and the T1w
template. The following template was were selected for spatial
normalization and accessed with \emph{TemplateFlow}
\citep[23.0.0,][]{templateflow}: \emph{ICBM 152 Nonlinear Asymmetrical
template version 2009c} {[}\citet{mni152nlin2009casym},
RRID:SCR\_008796; TemplateFlow ID: MNI152NLin2009cAsym{]}.
\item[Functional data preprocessing]
For each of the 1 BOLD runs found per subject (across all tasks and
sessions), the following preprocessing was performed. First, a reference
volume and its skull-stripped version were generated using a custom
methodology of \emph{fMRIPrep}. Head-motion parameters with respect to
the BOLD reference (transformation matrices, and six corresponding
rotation and translation parameters) are estimated before any
spatiotemporal filtering using \texttt{mcflirt} \citep[FSL
6.0.6.4,][]{mcflirt}. BOLD runs were slice-time corrected to 0.961s (0.5
of slice acquisition range 0s-1.92s) using \texttt{3dTshift} from AFNI
20230309 \citep[RRID:SCR\_005927]{afni}. The BOLD time-series (including
slice-timing correction when applied) were resampled onto their
original, native space by applying the transforms to correct for
head-motion. These resampled BOLD time-series will be referred to as
\emph{preprocessed BOLD in original space}, or just \emph{preprocessed
BOLD}. The BOLD reference was then co-registered to the T1w reference
using \texttt{bbregister} (FreeSurfer) which implements boundary-based
registration \citep{bbr}. Co-registration was configured with six
degrees of freedom. Several confounding time-series were calculated
based on the \emph{preprocessed BOLD}: framewise displacement (FD),
DVARS and three region-wise global signals. FD was computed using two
formulations following Power (absolute sum of relative motions,
\citet{power_fd_dvars}) and Jenkinson (relative root mean square
displacement between affines, \citet{mcflirt}). FD and DVARS are
calculated for each functional run, both using their implementations in
\emph{Nipype} \citep[following the definitions by][]{power_fd_dvars}.
The three global signals are extracted within the CSF, the WM, and the
whole-brain masks. Additionally, a set of physiological regressors were
extracted to allow for component-based noise correction
\citep[\emph{CompCor},][]{compcor}. Principal components are estimated
after high-pass filtering the \emph{preprocessed BOLD} time-series
(using a discrete cosine filter with 128s cut-off) for the two
\emph{CompCor} variants: temporal (tCompCor) and anatomical (aCompCor).
tCompCor components are then calculated from the top 2\% variable voxels
within the brain mask. For aCompCor, three probabilistic masks (CSF, WM
and combined CSF+WM) are generated in anatomical space. The
implementation differs from that of Behzadi et al.~in that instead of
eroding the masks by 2 pixels on BOLD space, a mask of pixels that
likely contain a volume fraction of GM is subtracted from the aCompCor
masks. This mask is obtained by dilating a GM mask extracted from the
FreeSurfer's \emph{aseg} segmentation, and it ensures components are not
extracted from voxels containing a minimal fraction of GM. Finally,
these masks are resampled into BOLD space and binarized by thresholding
at 0.99 (as in the original implementation). Components are also
calculated separately within the WM and CSF masks. For each CompCor
decomposition, the \emph{k} components with the largest singular values
are retained, such that the retained components' time series are
sufficient to explain 50 percent of variance across the nuisance mask
(CSF, WM, combined, or temporal). The remaining components are dropped
from consideration. The head-motion estimates calculated in the
correction step were also placed within the corresponding confounds
file. The confound time series derived from head motion estimates and
global signals were expanded with the inclusion of temporal derivatives
and quadratic terms for each \citep{confounds_satterthwaite_2013}.
Frames that exceeded a threshold of 0.5 mm FD or 1.5 standardized DVARS
were annotated as motion outliers. Additional nuisance timeseries are
calculated by means of principal components analysis of the signal found
within a thin band (\emph{crown}) of voxels around the edge of the
brain, as proposed by \citep{patriat_improved_2017}. The BOLD
time-series were resampled into standard space, generating a
\emph{preprocessed BOLD run in MNI152NLin2009cAsym space}. First, a
reference volume and its skull-stripped version were generated using a
custom methodology of \emph{fMRIPrep}. All resamplings can be performed
with \emph{a single interpolation step} by composing all the pertinent
transformations (i.e.~head-motion transform matrices, susceptibility
distortion correction when available, and co-registrations to anatomical
and output spaces). Gridded (volumetric) resamplings were performed
using \texttt{antsApplyTransforms} (ANTs), configured with Lanczos
interpolation to minimize the smoothing effects of other kernels
\citep{lanczos}. Non-gridded (surface) resamplings were performed using
\texttt{mri\_vol2surf} (FreeSurfer).
\end{description}

Many internal operations of \emph{fMRIPrep} use \emph{Nilearn} 0.9.1
\citep[RRID:SCR\_001362]{nilearn}, mostly within the functional
processing workflow. For more details of the pipeline, see
\href{https://fmriprep.readthedocs.io/en/latest/workflows.html}{the
section corresponding to workflows in \emph{fMRIPrep}'s documentation}.

\hypertarget{copyright-waiver}{%
\subsubsection{Copyright Waiver}\label{copyright-waiver}}

The above boilerplate text was automatically generated by fMRIPrep with
the express intention that users should copy and paste this text into
their manuscripts \emph{unchanged}. It is released under the
\href{https://creativecommons.org/publicdomain/zero/1.0/}{CC0} license.

\hypertarget{references}{%
\subsubsection{References}\label{references}}

  \bibliography{/output/logs/CITATION.bib}

\end{document}

### 2024-02-13 paper/paper.tex
[WIP]

@@
 \usepackage{booktabs}
+\usepackage{amsmath}
+\usepackage{import}
 
@@
 
-\title{A Detailed Analysis of Performance Bottlenecks in MRI Pre-Processing}
-
-\author{Mathieu Dugr\'e, Tristan Glatard \TG{I think Yohan should be co-author given his technical input}}
+\title{An Analysis of Performance Bottlenecks in MRI Pre-Processing}
+\author{Mathieu Dugr\'e, Yohan Chatelain, Tristan Glatard}
 
 \begin{document}
-
 \maketitle
@@
 \section{Introduction}
-% Intro: Large dataset, long processing time, I/O intensive (derivatives), focus on MRI, no current study, pipelines?
-% **Main question:** What are the main bottleneck in MRI pre-processing?
-% **Why:** Performance of neuroimaging pipeline is critical (larger study, clinical use). However, limited study on performance profile.
+% Computational bottleneck
+Pre-processing of Magnetic Resonance Imaging (MRI) data is a critical step in neuroimaging.
+Although established pre-processing pipelines exist, they commonly require expensive computation and produce large output and intermediate data.
+This has been a major bottleneck to study large subject cohorts in neuroimaging.
+Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed.
+For those reasons, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.
 
-\TG{Good intro but I think it would deserve a better logic. You could start by explaining that computational bottlenecks 
-have been one of the main bottlenecks to the processing of large cohorts in neuroimaging, that the community is constantly searching for ways to process data faster (HPC, cloud, GPUs, DL models), and that your paper takes a step back by analyzing the computational bottlenecks in order to subsequently focus on their optimization.
-}
+% HPCs
+Researcher broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts.
+HPC systems provide a large number of CPU cores and high-bandwidth memory to process subjects in parallel.
+However, HPCs might not be available to researchers in developing country.
+Similarly, cloud computing allows scaling the number of core and memory to process large cohort.
+However, cloud computing requires expertise to setup and can be costly for large cohort.
+Moreover, data transfer can be slow and might be impossible due to ethic concerns related to human data.
+In both cases, the benefits from parallelization are limited by the parallel efficiency of the pipeline and the number of subjects to process.
+Therefore, it can lead to under-utilization of the available resources.
 
-Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain.
-Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI).
-While other modalities such as EEG, CT, and PET exist \TG{define acronyms}, we focus on MRIs for their broad adoption and non-invasive property.
-However, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.
+%  GPUs
+\MD{TODO GPUs / Accelerators}
 
-Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data.
-Among those, some openly available and commonly accepted emerged: AFNI~\cite{Cox1996-sl}, ANTs~\cite{Avants_undated-fu}, FreeSurfer~\cite{Fischl2012-bp}, FSL~\cite{Jenkinson2012-cq}, and SPM~\cite{Friston2007-ag}.
-While the previous pipelines focus on specific steps of MRI pre-processing, tools like fMRIPrep~\cite{Esteban2019-og} and dMRIPrep~\cite{Michael2021-4515513} combine various pipelines to produce a complete pre-processing workflow.
+% DL
+In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing.
+For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} with the effort to provide Deep Learning alternative to traditional algorithms.
+The main benefit of those tools is the faster runtime they provide.
+Moreover, A study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
+However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive.
+Additionally, the results from DL models can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.
 
-The MRI pre-processing of a subject can take several hours, preventing the clinical use where data analysis might be required in timely manner.
-Moreover, the lengthy pre-processing hinders researchers with limited computational resources to perform research on large cohorts.
-
-The first step to improve performance of an application is to understand its performance bottleneck.
-Following, we can focus our efforts on the performance critical components.
-For example, better algorithm designs or usage of hardware accelerator can reduce computation time.
-Compression techniques can lower data transfer time for output and intermediate data.
-At last, reduced precision arithmetic can lower computation time, memory footprint, and storage size.
-While these techniques can provide significant performance improvements, their implementation can require substantial effort.
+% Alternative solutions
+While those techniques can provide significant performance improvements, their additional hardware requirement can be costly and unaccessible to some researchers.
+Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements.
+However, their implementation can require substantial effort.
 Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.
 
+% - No comprehensive study of the performance bottleneck of MRI pre-processing.
 The neuroimaging community is well aware that MRI pre-precessing pipelines require long computation time.
@@
 
-\TG{you could summarize the method (vtune profiler) in the intro.}
+% - Focus on MRI pre-processing.
+Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain.
+Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI).
+While other modalities such as electroencephalogram (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRIs for their broad adoption and non-invasive property.
+Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.
 
-\section{Methods}
-\subsection{Infrastructure}
-For our experiments, we used the \textit{Slashbin cluster} at Concordia University.
-The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}.
-\TG{I would defer this subsection to later in the methods as it's not the most central one}
+% - Pipelines
+Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data.
+We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} workflow as it a comprehensive pre-processing pipeline for both sMRI and fMRI.
+fMRIPrep combines several pipelines to produce a complete pre-processing workflow.
+We also study its sub-workflow separately to provide a finer grain analysis of the performance bottleneck.
+
+% - data: OpenNeuro ds004513
+We use the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
+We chose this dataset for its wide range of age and equal distribution of biological sex.
+Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.
+
+% - VTune profiler
+To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}.
+The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level.
+We profile the application using two threading configuration: single-threaded and 32 threads.
+We aggregate subjects' results for each pipeline to obtain an average performance profile.
+We compiled each application with debug information to obtain human readable information on function and module names in VTune.
+
+\section{Materials \& Methods}
+\subsection{Pipelines}
+In this paper, we study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow.
+First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance.
+Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}.
+We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset.
+Moreover, the statistical analysis of the fMRIPrep workflow are not computationally expensive and are not the focus of this paper.
+
+\subsubsection{ANTS registrationSyN}
+% Spatial Normalization
+% SyN Registration: https://doi.org/10.1016/j.media.2007.06.004
+% TODO Mention the option for single and double precision.
+
+\subsubsection{ANTS brainExtraction}
+% ANTS brainExtraction
+% 1. Truncate image intensity
+% 2. N4 bias correction: https://doi.org/10.1109/TMI.2010.2046908
+% 3. SyN Registration: : https://doi.org/10.1016/j.media.2007.06.004
+% 4. Warp prior to image using affine
+% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/
+% TODO Mention the option for single and double precision.
+% TODO OASIS template citation
+
+\subsubsection{FSL FAST}
+% DOI: https://doi.org/10.1109/42.906424
+% Segmentation
+
+\subsubsection{FreeSurfer recon-all}
+% Surface Reconstruction: https://doi.org/10.1006/nimg.1998.0395
+% Refer to the fMRIPrep specific 3 stages workflow.
+
+\subsubsection{FSL FLIRT}
+% Registration
+% FSL FLIRT: https://doi.org/10.1006/nimg.2002.1132, https://doi.org/10.1016/S1361-8415(01)00036-6, https://doi.org/10.1016/j.neuroimage.2009.06.060
+
+\subsubsection{FSL MCFLIRT}
+% Head-Motion Correction
+% FSL MCFLIRT: https://doi.org/10.1006/nimg.2002.1132
+
+\subsubsection{fMRIPrep}
+\MD{TODO import fMRIPrep citation boilerplate from "./fmriprep-citations"}
 
@@
 
-We used the the OpenNeuro ds004513 v.1.0.2 dataset~\cite{ds004513:1.0.2}.
+We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
 The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts.
@@
 
-\subsection{Pipelines}
-% TODO rework
-In this analysis, we aim to cover the pre-processing performance for different MRI modalities.
-We study the performance bottleneck of large toolboxes and individual pipelines.
-Moreover, we use several pipelines from the LivingPark project~\cite{livingpark} as a real use case of MRI pre-processing.
-
-First, we profile these commonly accepted workflows: fMRIPrep anatomical only without and without FreeSurfer, fMRIPrep complete workflow, and dMRIPrep.
-This lets us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.
-
-Second, we perform a finer grain analysis of the aforementioned workflows' sub-components. We list the different sub-workflow and describe the key steps for each modality in the section below.
-
-Last, the LivingPark project replicates several papers on Parkinson's Disease with a broad range of analysis.
-The study uses the following pipelines: FSL SIENA/VIENA, SPM VBM, SPM DARTEL, and SPM pairwise registration.
-% TODO Decide which pipelines to use from the project.
-
-% ############################################################
-% Pre-Processing
-% ############################################################
-% Look at the benchmarked tool and the information needed to abstract those tools.
-% Discussion of the pre-processing steps.
-% Explain their computation and I/O requirements.
-\subsection{sMRI Pre-Processing}
-\subsubsection{Intensity correction}
-% N4 bias correction (Check ANTS brainExtraction)
-% https://ieeexplore.ieee.org/document/5445030
-
-\subsubsection{Reference Image}
-% Longitudinal study 
-% FreeSurfer mri_robust_template
-
-\subsubsection{Spatial Normalization}
-% ANTS registrationSyN
-% SyN Registration: https://pubmed.ncbi.nlm.nih.gov/17659998/
-
-\subsubsection{Skull-Stripping}
-% ANTS brainExtraction
-% 1. Truncate image intensity
-% 2. N4 bias correction
-% 3. SyN Registration
-% 4. Warp prior to image using affine
-% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/
-
-\subsubsection{Segmentation}
-% FSL FAST
-
-\subsubsection{Surface Reconstruction}
-% FreeSurfer recon-all
-
-\subsection{fMRI Pre-Processing}
-\subsubsection{Slice-Timing Correction}
-% AFNI 3dTShift
-
-\subsubsection{Head-Motion Correction}
-% FSL MCFLIRT
-
-\subsubsection{Susceptibility Distortion Correction}
-% SDC workflow
-
-\subsubsection{Registration}
-% FSL FLIRT
-
-\subsection{dMRI Pre-Processing}
-% TODO: Decide which pipelines to use
-% \subsubsection{Eddy Currents Distortion Estimation}
-% % FSL eddy
-
 \subsection{Profiling}
 Profiling MRI pre-processing pipelines present a set of challenges.
-The neuroimaging community develops pipelines using several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Python}, and \textit{Matlab}, with several pipelines utilizing a combination of multiple languages.
+The neuroimaging community develops pipelines using several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages.
 Pipelines are generally complex and computationally expensive.
 
-The Intel VTune profiler~\cite{vtune_profiler} satisfies these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level.
-However, ome general profiling challenges remain.
+The Intel VTune profiler solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level.
+However, common profiling challenges remain.
 Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules.
@@
 We discuss our approach to these challenges in the next section.
-% TODO add citation on effect of different input data on numerical stability and reduced precision.
+\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for refereces}
 
 First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
-For use on HPC systems, we created Singularity images using Docker2Singularity \lstlistingname~\ref{vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Singularity image at execution.
+For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Apptainer image at execution.
+
+\begin{minipage}{\linewidth}
 	\lstinputlisting[
 		language={sh},
-	caption={Profiling an application within a Singularity container, using VTune profiler.},
+		caption={Profiling an application within a Apptainer container, using VTune profiler.},
 		captionpos=b,
-	label=vtune_example,
+		label=alg:vtune_example,
 		numbers=left,
@@
 	]{vtune-example.sh}
+\end{minipage}
 
-Second, to estimate an average performance profile, we aggregate the profiling results of each pipeline across all subjects.
-Combining the Pareto's law and VTune hotspots report \TG{It would be useful to define these terms}, we present the functions accounting for most of the CPU time up to 80\% of the total CPU time of the application \TG{check wording}.
+Second, we aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile.
+The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes.
+We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}.
+Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application.
 This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
@@
 
-Last, we profile the application two threading configuration: single-threaded and 32 threads.
-On the one hand, the single-threaded approach limits the potential overhead from I/O on the profiling.
-On the other hand, we evaluate the multi-threading performance of the applications using maximum number of CPU core available on our system (i.e.
-32).
+Last, we profile the application using two threading configuration: single-threaded and 32 threads.
+On the one hand, the single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling.
+On the other hand, we use all available CPU cores from a compute node (i.e.
+32) to study the multi-threading performance of the application.
+This is a more realistic usage for the user of the application.
 
@@
 
+\subsection{Metrics Definition}
+We define the makespan as the elapsed time from the start to the end of the application.
+The CPU time refers to the total time spent by the CPU, across all core, to execute the application.
+VTune provides several metrics to characterize the performance of an application.
+We describe the memory metrics used in this paper.
+
+\begin{figure*}[h!]
+	\noindent
+	\begin{minipage}
+		{0.5\textwidth}
+		\begin{align*}
+			C & = CPU\_CLK\_UNHALTED.THREAD                \\
+			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
+			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
+			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
+			W & = MEM\_L3\_WEIGHT                          \\
+			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
+			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
+			M & = \frac{W \times R}{H + R}                 \\
+			N & = \frac{H}{H + W \times R}                 
+		\end{align*}
+	\end{minipage}
+	\begin{minipage}
+		{0.5\textwidth}
+		\begin{align}
+			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
+			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
+			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
+			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
+			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
+		\end{align}
+	\end{minipage}
+	\caption{Intel performance monitoring events and derived memory metrics}
+	\label{fig:memory-metrics}
+\end{figure*}
+
+The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load.
+L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache.
+Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively.
+The store bound occurs when a store operation stall the CPU.
+Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory boudnd characterization~\cite{Intel2006-lc}.
+We use the convention from~\cite{Kukunas2015-jd} for the equations.
+
+The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level; i.e.
+the data is fetched from disk.
+
+\subsection{Infrastructure}
+For our experiments, we used the \textit{Slashbin cluster} at Concordia University.
+The compute nodes are configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), Rocky~Linux~8.9 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}.
+
 \section{Results}
-\TG{Interesting results but presentation needs to be worked out.
-This section should be structured with sub-headings summarizing the main findings, for instance: "Interpolation is the main bottleneck", etc}
+In this section we present the aggregated profiling data obtained with VTune.
+First, we present a performance overview of the pipelines.
+We then dive into more detailed analysis observed from the hotspots of the pipelines.
+Unless specified explicitly, we only present the profile data from using 1-thread since the results from 32-threads are similar.
 
-% TODO
-% 2-step approach:
-% 1. Which pipeline constitute the long execution time in preprocessing?
-% 2. For those, which functions take the most time to execute? (hotspots)
-% Any key finding: common point between functions, mem bound, fops, etc.
+\subsection{Long-tail Distribution}
+Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time is spent in less than 1.5\% of the function across all pipelines.
+While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a pareto distribution.
+We plotted the distribution of the functions' CPU time on a log-log scale and found it to be closer to a Zipf's law distribution with exponent $\alpha=1.8$.
 
-% In this section, analyze the aggregated profiling data obtained with VTune.
-First, a coarse summary of the workflows and pipelines is presented using their makespan.
-Second, we depict detailed analysis for the key functions contributing to the CPU time of each application (hotspots).
-
-\subsection{Makespan}
-% Identify which tools take the longest to execute.
-% This will let us decided which pipeline to investigate more.
-Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan \TG{define makespan}, in seconds, while using 1 thread for one execution of all subjects.
-Unexpectedly, we observe that both ANTS~BrainExtration and ANTS~RegistrationSyN are slower when using single precision compared to double precision.
-% TODO Stat test between mean for single and double precision.
 \begin{figure}[h]
 	\centering
+	\includegraphics[width=\linewidth]{figures/global-longtail.png}
+	\caption{Distribution of the functions' CPU time.
+The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage.
+The x-axis percentage of functions ordered by decreasing CPU time.
+The data includes all functions from all pipelines.}
+	\label{fig:long-tail-distribution}
+\end{figure}
+
+\subsection{Makespan: A performance overview}
+Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e.
+makspan.
+There is a significant difference between the makespan of different applications.
+This is expected as the pipelines perform different tasks.
+The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation.
+Overall, this shows that the input data does not substantially affect the makespan of those pipelines.
+
+\begin{figure}[t]
+	\centering
 	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
-	\caption{Profiling of the pipelines using 1 thread. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
+	\caption{Average makespan of the pipeline while using 1-thread. The error bars show the standard deviation. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
 	\label{fig:makespan-1thread}
@@
 
-\subsection{Hotspots}
+\subsection{Impact of Memory Bounded Functions}
+Table~\ref{extab:memory-bound} shows that fetching data from disk is the main source of cycles where the CPU is starved.
+Following, The L1 bound has the highest cache-level contribution to the memory bound of the pipelines.
+ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT.
+L2 and L3 bound is low for all pipeline, although highest for FreeSurfer recon-all.
+The DRAM bound is high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision.
+Store bound is low for all pipelines.
+We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.
 
-Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspots observed from profiling the fMRIPrep anatomical sub-workflows.
-In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module.
-The function ID are found in the supplementary material Section~\ref{sec:supplementary}.
-The bars show the average CPU time in seconds of a function and the error bars indicate the standard deviation, with the left y-axis showing the scale.
-The light-green dots represent the functions cumulative percentage makespan of the application, with the right y-axis showing the scale.
-The color of the bar indicate the percentage of calls, for the function, that were ``stalled due to demand memory loads and stores''\footnote{\href{https://www.intel.com/content/www/us/en/docs/vtune-profiler/user-guide/2023-0/cpu-metrics-reference.html\#MEMORY-BOUND}{Vtune Memory Bound metric}}, with the left color bar showing the scale.
+\csvnames{csvcol}{1=\pipeline, 2=\mem, 3=\la, 4=\lb, 5=\lc, 6=\dram, 7=\store}
+\csvstyle{memory_bound}{
+	tabular = |r|c|c|c|c|c|c|,
+	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound\\\hline\hline,
+	late after line = \\\hline,
+	respect all,
+	csvcol
+}
+\begin{table*}[ht]
+	\centering
+	\csvreader[memory_bound]{tables/memory_bound.csv}{}
+	{\pipeline & \tablenum[round-precision=2]{\mem} & \tablenum[round-precision=2]{\la} & \tablenum[round-precision=2]{\lb} & \tablenum[round-precision=2]{\lc} & \tablenum[round-precision=2]{\dram} & \tablenum[round-precision=2]{\store}}
+	\caption{Impact of data load on stalled CPU cycles.
+The values are the summation of the metric weighted by each function CPU time.
+This represent the percentage of the total CPU time stalled by each metrics.
+We show the average value across all subjects execution with one thread.}
+	\label{extab:memory-bound}
+\end{table*}
 
-\TG{I'd move the graphical explanations to the figure caption so that the text can focus on observations}
+\subsection{Main Bottleneck: Linear Interpolation}
+Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all.
+For those pipelines, interpolation contributes between 32.20\% to 62.70\% of the total CPU time.
+Overall, the number of functions using interpolation is low with less than 20 in each applications.
+In other words, few interpolation functions are contribute to a large portion of the CPU time.
+% There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)
 
-In figure~\ref{subfig:hotspots-ants-brainExtraction}, we observe that the function 1 and 2 account for a significant portion of the application execution time.
-Moreover, calls from function~1 are memory bound $51.84\pm{4.55}$ percent of the time.
-Table~\ref{extab:hotspots-1thread-ants-brainExtraction} shows that both functions are part of the ITK linear interpolation module.
+\csvnames{csvcol}{1=\pipeline, 2=\nfunc, 3=\cputime}
+\csvstyle{interpolation}{
+	tabular = |r|c|c|,
+	table head = \hline Pipeline & \# of functions & \% CPU Time \\
+	& with interpolation & from interpolation\\\hline\hline,
+	late after line = \\\hline,
+	respect all,
+	csvcol
+}
+\begin{table*}[h]
+	\centering
+	\csvreader[interpolation]{tables/interpolation.csv}{}
+	{\pipeline & \nfunc & \tablenum[round-precision=2]{\cputime}}
+	\caption{Contribution of interpolation to the applications' total CPU time.
+The percentage is the average sum of CPU time of functions using interpolation.
+The data includes all functions; not only the top 80\% of the CPU time.}
+	\label{extab:interpolation}
+\end{table*}
 
-Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two functions accounting for the most time, although in flipped order.
-Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time.
-Compared to the double precision version of ANTS brainExtraction, the third function runtime is significantly larger.
-It is an ITK function transforming points while calculating the displacement field of the image.
+\subsection{ANTS: Single vs. Double Precision}
+Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision.
+For both pipelines, the makespan is significantly higher in single precision.
+This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv}.
 
-Figure~\ref{subfig:hotspots-ants-registrationSyN} and figure~\ref{subfig:hotspots-ants-registrationSyN-fp} show that the execution time for ANTS registrationSyN is dominated by a single function.
-In both case, the ITK function evaluating point during linear interpolation.
-The next two function are also shared between the application, although in flipped order.
-These functions show some memory bound in the range of 16--27.5.
-We also note the three most expensive functions are the same as in both versions of ANTS brainExtraction.
+\begin{figure}[h]
+	\includegraphics[width=\linewidth]{figures/makespan-ants.png}
+	\caption{Comparison of makespan between double (blue) and single (orange) precision for ANTS brainExtraction and ANTS registrationSyN.}
+	\label{fig:makespan-ants}
+\end{figure}
 
-% TODO FreeSurfer 1-thread sub-figure
+The number of iteration between both version of ANTS registrationSyN is similar.
+In fact, for the latest two stages, they were the same across all but one execution.
+However, The single precision version of ANTS registrationSyN took longer per iteration than the double precision version (Figure~\ref{fig:mean-time-per-iteration-ants}).
+Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.
 
-In figure~\ref{subfig:hotspots-32threads-freesurfer-reconall}, we observe that OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime.
-The only function shown from FreeSurfer recon-all is the same as the first function from figure~\ref{subfig:hotspots-1thread-freesurfer-reconall}.
-However, it only accounts for 2\% of the total makespan compared to 3.8\% while running sequentially.
+\begin{figure}
+	\includegraphics[width=\linewidth]{figures/ants-registrationSyN-iteration-mean.png}
+	\caption{Mean time per iteration for ANTS registrationSyN in double and single precision.
+Only the SyN Registration stage is shown as the two earlier stage were near zero time.
+The error bars show the standard deviation.}
+	\label{fig:mean-time-per-iteration-ants}
+\end{figure}
 
-\begin{figure*}[ht!]
+\subsection{FreeSurfer: Thread-Synchronization}
+Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant different in the pipeline profile between single-threaded and multi-threaded execution.
+The multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime.
+FreeSurfer uses the multi-processing API provided by OpenMP.
+Moreover, the y-axis scale is multi folds larger than for the single-threaded executions.
+This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.
+
+\begin{figure*}
 	\centering
-	% ANTS brainExtraction
 	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-ants-brainExtraction}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
+		\caption{Single-threaded}
+		\label{subfig:hotspots-freesurfer-reconall-1thread}
+		\includegraphics[width=\textwidth]{figures/hotspots-1threads-freesurfer-reconall-simple.png}
 	\end{subfigure}
-	\hfill
 	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-ants-brainExtraction-fp}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
+		\caption{Multi-threaded (32 threads)}
+		\label{subfig:hotspots-freesurfer-reconall-32threads}
+		\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall-simple.png}
 	\end{subfigure}
-	% ANTS registrationSyN
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-ants-registrationSyN}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-ants-registrationSyN-fp}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
-	\end{subfigure}
-	\caption{ANTS: Single vs. Double precision}
-	\label{fig:ants-fp-comparison}
+	\caption{FreeSurfer recon-all hotspots analysis.
+The y-axis show the average CPU time spent in each function, with error bars showing the standard deviation.
+The x-axis show the function ordered by decreasing CPU time grouped by module.
+We omit the function names for clarity.
+The function ID are dependent to each plot.
+The supplementary materials depicts the mapping of the function ID to the function name for each plot.}
+	\label{fig:hotspots-freesurfer-reconall}
 \end{figure*}
 
-\section{old Results}
-% TODO
-% 2-step approach:
-% 1. Which pipeline constitute the long execution time in preprocessing?
-% 2. For those, which functions take the most time to execute? (hotspots)
-% Any key finding: common point between functions, mem bound, fops, etc.
+While the makespan for FreeSurfer recon-all decrease with an increase in the number of threads, the benefits are limited (Figure~\ref{fig:freesurfer-threading}).
+The parallel efficiency decreases from 66\% with two threads down to 5\% with 32 threads.
+While the Amdhals' law plays an impact on the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} indicates that the thread Synchronization has a larger impact.
 
-% In this section, analyze the aggregated profiling data obtained with VTune.
-First, a coarse summary of the workflows and pipelines is presented using their makespan.
-Second, we depict detailed analysis for the key functions contributing to the CPU time of each application (hotspots).
-
-\subsection{Hotspots}
-Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspot observed from profiling the fMRIPrep anatomical sub-workflows.
-In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module.
-The function ID are found in the supplementary material Section~\ref{sec:supplementary}.
-The bars show the average CPU time in seconds of a function and the error bars indicate the standard deviation, with the left y-axis showing the scale.
-The light-green dots represent the functions cumulative percentage makespan of the application, with the right y-axis showing the scale.
-The color of the bar indicate the percentage of calls, for the function, that were ``stalled due to demand memory loads and stores''\footnote{\href{https://www.intel.com/content/www/us/en/docs/vtune-profiler/user-guide/2023-0/cpu-metrics-reference.html\#MEMORY-BOUND}{Vtune Memory Bound metric}}, with the left color bar showing the scale.
-
-In figure~\ref{subfig:hotspots-ants-brainExtraction}, we observe that the function 1 and 2 account for a significant portion of the application execution time.
-Moreover, calls from function~1 are memory bound $51.84\pm{4.55}$ percent of the time.
-Table~\ref{extab:hotspots-1thread-ants-brainExtraction} shows that both functions are part of the ITK linear interpolation module.
-
-Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two function accounting for the most time, although in flipped order.
-Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time.
-Compared to the double precision version of ANTS brainExtraction, the third function runtime is significantly larger.
-It is an ITK function transforming points while calculating the displacement field of the image.
-
-Figure~\ref{subfig:hotspots-ants-registrationSyN} and figure~\ref{subfig:hotspots-ants-registrationSyN-fp} show that the execution time for ANTS registrationSyN is dominated by a single function.
-In both case, the ITK function evaluating point during linear interpolation.
-The next two function are also shared between the application, although in flipped order.
-These functions show some memory bound in the range of 16--27.5.
-We also note the three most expensive functions are the same as in both versions of ANTS brainExtraction.
-
-% TODO FreeSurfer 1-thread sub-figure
-% Note:
-% - Way more functions
-% - Top 5 accounts for only 15% of total time. Much less than other applications.
-% - Way more memory bound that previous application.
-% - Focusing on Top 5 and memory bound could give good performance boost.
-
-\begin{figure*}[ht!]
-	\centering
-	% % ANTS brainExtraction
-	% \begin{subfigure}[t]{0.49\textwidth}
-	% 	\caption{}
-	% 	\label{subfig:hotspots-ants-brainExtraction}
-	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
-	% \end{subfigure}
-	% \hfill
-	% \begin{subfigure}[t]{0.49\textwidth}
-	% 	\caption{}
-	% 	\label{subfig:hotspots-ants-brainExtraction-fp}
-	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
-	% \end{subfigure}
-
-	% % ANTS registrationSyN
-	% \begin{subfigure}[t]{0.49\textwidth}
-	% 	\caption{}
-	% 	\label{subfig:hotspots-ants-registrationSyN}
-	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
-	% \end{subfigure}
-	% \hfill
-	% \begin{subfigure}[t]{0.49\textwidth}
-	% 	\caption{}
-	% 	\label{subfig:hotspots-ants-registrationSyN-fp}
-	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
-	% \end{subfigure}
-
-	% % FreeSurfer recon-all
-	% \begin{subfigure}[t]{\textwidth}
-	% 	\caption{}
-	% 	\label{subfig:hotspots-1thread-freesurfer-reconall}
-	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-freesurfer-reconall.png}
-	% \end{subfigure}
-
-	% \begin{subfigure}[t]{0.49\textwidth}
-	% 	\caption{}
-	% 	\label{subfig:hotspots-32threads-freesurfer-reconall}
-	% 	\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall.png}
-	% \end{subfigure}
-	\hfill
-	% FSL FAST
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-fsl-fast}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-fast.png}
-	\end{subfigure}
-
-	\caption{Hotspots profiling of fMRIPrep's anatomical sub-workflows.}
-	\label{fig:hotspots-fmriprep-anat-subworkflow}
-
-	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
-\end{figure*}
-
-\begin{figure*}[ht!]
-	\centering
-	% FSL MCFLIRT
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-fsl-mcflirt}
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-mcflirt.png}
-	\end{subfigure}
-	% FSL FLIRT
-	\hfill
-	\begin{subfigure}[t]{0.49\textwidth}
-		\caption{}
-		\label{subfig:hotspots-fsl-flirt}
-
-		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-flirt.png}
-	\end{subfigure}
-	\caption{Hotspots profiling of fMRIPrep's functional sub-workflows.}
-	\label{fig:hotspots-fmriprep-func-subworkflow}
-
-	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
-\end{figure*}
+\begin{figure}
+	\includegraphics[width=\linewidth]{figures/makespan-freesurfer.png}
+	\caption{Makespan and parallel efficiency of FreeSurfer recon-all while varying the number of threads between 1 to 32.
+The left y-axis shows the makespan in seconds, while the right y-axis shows the parallel efficiency in percent.
+The log-scaled x-axis shows the number of threads.}
+	\label{fig:freesurfer-threading}
+\end{figure}
 
 \section{Discussion}
-% TODO
+\subsection{Few functions account for most of the CPU time}
+Our initial assumption was that the distribution of CPU time would follow a Pareto distribution.
+Thus, 20\% of the functions would contribute to 80\% of the total CPU time.
+However, Figure~\ref{fig:long-tail-distribution} shows that the distribution of CPU time is closer to a Zipf's law with exponent $\alpha=1.8$.
+This means that the CPU time of a function is approximately inversely proportional to its rank.
+This is consistent with the Pareto's principle, but is more extreme.
+Therefore, future efforts to optimize the main bottlenecks of MRI pre-processing pipelines could lead to enormous performance improvements.
 
-% Principal Bottlenecks
-% What are common key points between the different applications?
-% Possibly move this to the Results section.
-% - Top 3 functions in ANTS application are the same, although in varying order.
-% -
+\subsection{Wasted CPU cycles due to Memory Bound}
+Overall, our results show a high percent of memory bound across all pipelines.
+Optimizing the data access of the application could reduce the number of data load.
+This would potentially have a large impact on the performance of the pipelines.
+Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound.
+To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing.
+In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
 
-% # Limitation
-% - Can only perform profiling of open-source pipelines or pipelines with debug-info.
-% - No information about space or time for function call.
-% - Dataset only contains healthy subjects (no pathology).
-% - Dataset is from a single site. Possible bias from scanner or protocol.
+%%%%%%%%%% # Linear Interpolation
+% Theoretical improvements using reduced precision. If time permits, toy experiment with ITK single precision data type at different locations.
+% 1. Winograd algorithm for fast convolution.
+% - It could be used to speed up the current pipeline. Tradeoff with precision? How does it compare to reduced precision?
+% 2. Improvement to interpolation would help the current algorithms, as well as the DL algorithm using CNN.
+\subsection{Importance of Linear Interpolation}
+\MD{TODO}
 
-% # Opportunity for optimization (future work)
-% Reduced precision for: Compute or Compression. Reduce memory footprint and compute time. -> Mention the applications with high FP-ops and MEM-bound.
+%%%%%%%%%% # Single vs. Double Precision
+% 1. Not as simple as using different data type. It could affect
+% - Convergence
+% - Accuracy
+% - Runtime (maybe even slower)
+% 1.5 Convergence is the sane in our case (same number of iterations).
+% 2. Why is it slower in our case?
+% -? Possibly less optimized code for single precision.
+% -? Possibly memory placement
+% -? ...
+% 3. 
+\subsection{ANTS: A tale of precision}
+\MD{TODO + change title}
 
-% # Linear Interpolation
-% Theoretical improvements using reduced precision. If time permits, toy experiment with ITK single precision data type at different locations.
+\subsection{OpenMP bottleneck in FreeSurfer recon-all}
+In our experiments, FreeSurfer recon-all showed a significant decrease in parallel efficiency with an increase in the number of threads.
+While the Amdhals' law plays an impact on the maximum efficiency, the OpenMP thread synchronization played the largest impact in our results.
+The codebase use 83 \textit{static} scheduling policy out of 92 total.
+\textit{Static} scheduling assign chunks to threads in round-robin fashion.
+This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency.
+We supposed that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads.
+Unfortunately, we failed to recompile FreeSurfer by naively changing the \textit{static} OpenMP scheduling to \textit{dynamic}.
+This remain challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
 
-% # Single precision slowdown?
-% Possibly due to slower convergence and less optimized code. (Quick drop-in of ITK single precision data type.)
-% Need to investigate more.
+Future work should study the different configuration of OpenMP scheduling in FreeSurfer recon-all to improve the parallel efficiency.
+We believe that significant performance improvements could be achieved by optimizing the thread synchronization.
+This would lead to faster runtime time and higher CPU usage ration in HPC allocations.
 
-% OpenOMP bottleneck (FreeSurfer with 32 threads)
-% Majority of the time is spent waiting to synchronize threads.
-% If times permit, changes the scheduling type and run for a subject to see if we can get performance improvements.
+\subsection{Limitations}
+The dataset we chose contains wide range of age and equal distribution of sex.
+However, it only contains healthy subjects.
+Therefore, the performance bottleneck we found might not be applicable to data pre-processing of pathological patients.
+Moreover, the dataset is from a single site.
+This could also introduce bias from the scanner model and data acquisition protocol used.
+Future work should include data from multiple sites to ensure the generalization of the performance bottleneck we found.
+
+Our analysis omits the space or time of execution for function calls.
+This information could provide additional insight for optimization.
+For example, spatial and temporal dimensions were shown to be important in the field of reduced precision.
+Future work could extent the depth of the analysis by including this information.
+\MD{TODO Cite paper on spatial/temporal for RP (ask yohan for references)}
 
 \section{Conclusion}
-% TODO
+\MD{TODO}
 
@@
 \\\href{https://openneuro.org/datasets/ds004513/versions/1.0.2}{https://openneuro.org/datasets/ds004513/versions/1.0.2}.
-% TODO add compiled containers.
-% TODO add csv files from profiled runs.
-% The above can be distributed on DockerHub and Zenodo respectively.
+
+The container images for our profiling experiments are publicly on Docker Hub:
+\begin{itemize}
+	\item mathdugre/cmake:debug-info
+	\item mathdugre/intel-compilers:debug-info
+	\item mathdugre/afni:debug-info
+	\item mathdugre/ants:debug-info
+	\item mathdugre/fsl:debug-info
+	\item mathdugre/freesurfer:debug-info
+	\item mathdugre/fmriprep:debug-info
+\end{itemize}
+
+Our profiling results from VTune are publicly available at:
+\\\href{URL}{\MD{TODO add Zenodo link}}
 
@@
 \\\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
-% TODO Update URL to slashbin org, when created.
+\MD{TODO Update URL to slashbin org, when created.}
 
 \section*{Acknowledgement}
-% TODO
+\MD{TODO}
+% NSERC PGS-D
+% Concordia Bursary (find name)
+% Any from others?
+% ...
 
@@
 The authors report no conflict of interests.
-
-\section*{Supplementary Figures}
-% Pipelines' hotspot with bottomK makespan
 
@@
 \onecolumn
-\section*{Supplementary material}
+\section*{Supplementary Materials}
+\begin{figure*}
+	\centering
+	\includegraphics{figures/global-zipf_law.png}
+	\caption{Distribution of the functions' CPU time compare to a Zipf's law distribution with $\alpha=1.8$.
+The y-axis shows the average CPU time for a function.
+The x-axis shows the percentage of functions ordered by decreasing CPU time.
+The data includes all functions from all pipelines.}
+	\label{sup-fig:zips-law}
+\end{figure*}
+
 \label{sec:supplementary}
@@
 }
-
 % CSV tables
-% ANTS brainExtraction
-\csvtable{tables/hotspots-1thread-ants-brainExtraction.csv}{ANTS brainExtraction (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-brainExtraction}
-\csvtable{tables/hotspots-1thread-ants-brainExtraction-fp.csv}{ANTS brainExtraction-fp (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-brainExtraction-fp}
-
-% ANTS registrationSyN
-\csvtable{tables/hotspots-1thread-ants-registrationSyN.csv}{ANTS registrationSyN (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-registrationSyN}
-\csvtable{tables/hotspots-1thread-ants-registrationSyN-fp.csv}{ANTS registrationSyN-fp (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-registrationSyN-fp}
-
-% FreeSurfer recon-all
 % \csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
-% \csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}
-
-% FSL FAST
-\csvtable{tables/hotspots-1thread-fsl-fast.csv}{FSL FAST (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-fast}
-
-% FSL FLIRT
-\csvtable{tables/hotspots-1thread-fsl-flirt.csv}{FSL FLIRT (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-flirt}
-
-% FSL MCFLIRT
-\csvtable{tables/hotspots-1thread-fsl-mcflirt.csv}{FSL MCFLIRT (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-mcflirt}
+\csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}
 

### 2024-02-13 sbatch/fmriprep/anat-nofs.sh

@@ -1,62 +1,0 @@
-#!/bin/bash
-
-#SBATCH -J fmriprep-anat-anat-nofs
-#SBATCH --array=1
-#SBATCH --time=UNLIMITED
-#SBATCH --exclusive
-#SBATCH --nodes=1
-#SBATCH --ntasks=1
-#SBATCH --cpus-per-task=1
-#SBATCH --mem=0
-# -------------------------
-#SBATCH -o log/%x-%A-%a.out
-#SBATCH -e log/%x-%A-%a.err
-# -------------------------
-set -e
-set -u
-
-# Setup environment and parse args.
-source ./sbatch/pre_run.sh fmriprep anat-nofs -j ${SLURM_CPUS_PER_TASK} $@
-
-# Create an empty directory for output
-OUTPUT_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-outputs
-SUBJECT_OUTPUT=${OUTPUT_DIR}/sub-${SUBJECT_ID}
-if [ -d "${SUBJECT_OUTPUT}" ]; then rm -Rf ${SUBJECT_OUTPUT}; fi
-mkdir -p ${SUBJECT_OUTPUT}
-
-# Create WORK_DIR for fMRIPrep
-WORK_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-${RANDOM_STRING}
-mkdir -p ${WORK_DIR}
-
-SIF_OPTION+=" -B ${SUBJECT_OUTPUT}:/output"
-SIF_OPTION+=" -B ${WORK_DIR}:/workdir"
-export SIF_OPTION+=" -B $HOME/.freesurfer.txt:/opt/freesurfer/.license"
-cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
-FS_LICENSE=/opt/freesurfer/.license
-fmriprep \
-    /data \
-    /output \
-    participant \
-    --work-dir /workdir \
-    --skip_bids_validation \
-    --participant-label ${SUBJECT_ID} \
-    --nthreads ${NTHREAD} \
-    --omp-nthreads ${NTHREAD} \
-    --anat-only \
-    --fs-no-reconall
-EOT
-
-export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
-source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
-
[…]

### 2024-02-13 sbatch/vtune.sh

@@ -17,5 +17,6 @@ apptainer exec --cleanenv \
     -data-limit=0 \
     -knob enable-stack-collection=true \
     -knob analyze-openmp=true \
+    ${VTUNE_OPTION} \
     -result-dir ${PROFILING_DIR} \
     $@

### 2024-02-13 paper/paper.tex
[WIP]

@@
 \usepackage{amsmath}
-\usepackage{import}
 
@@
 % DL
-In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing.
-For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} with the effort to provide Deep Learning alternative to traditional algorithms.
-The main benefit of those tools is the faster runtime they provide.
-Moreover, A study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
-However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive.
-Additionally, the results from DL models can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.
+% SynthMorph, SynthSeg, etc.
+% Not broadly adopted yet
+\MD{TODO Deep Learning approach}
 
@@
 Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data.
-We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} workflow as it a comprehensive pre-processing pipeline for both sMRI and fMRI.
+We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-og} workflow as it a comprehensive pre-processing pipeline for both sMRI and fMRI.
 fMRIPrep combines several pipelines to produce a complete pre-processing workflow.
-We also study its sub-workflow separately to provide a finer grain analysis of the performance bottleneck.
+We also study the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN, FreeSurfer recon-all, FSL FAST, FSL FLIRT, and FSL MCFLIRT.
+\MD{TODO Cite the sub-workflows}
 
@@
 % - VTune profiler
-To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}.
+To profile the pipelines, we use the Intel VTune profiler~\cite{vtune_profiler}.
 The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level.
@@
 
+% - Outline of the paper.
+\MD{TODO outline?}
+
 \section{Materials \& Methods}
 \subsection{Pipelines}
-In this paper, we study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow.
+In this paper, we study the performance bottleneck of fMRIPrep~\cite{Esteban2019-og}; a commonly used fMRI and sMRI pre-processing workflow.
 First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance.
-Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}.
+Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTS brainExtraction, ANTS registrationSyN, FreeSurfer recon-all, FSL FAST, FSL FLIRT, and FSL MCFLIRT.
 We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset.
@@
 % Spatial Normalization
-% SyN Registration: https://doi.org/10.1016/j.media.2007.06.004
+% SyN Registration: https://pubmed.ncbi.nlm.nih.gov/17659998/
 % TODO Mention the option for single and double precision.
@@
 % 1. Truncate image intensity
-% 2. N4 bias correction: https://doi.org/10.1109/TMI.2010.2046908
-% 3. SyN Registration: : https://doi.org/10.1016/j.media.2007.06.004
+% 2. N4 bias correction: https://ieeexplore.ieee.org/document/5445030
+% 3. SyN Registration: : https://pubmed.ncbi.nlm.nih.gov/17659998/
 % 4. Warp prior to image using affine
@@
 % TODO Mention the option for single and double precision.
-% TODO OASIS template citation
 
 \subsubsection{FSL FAST}
-% DOI: https://doi.org/10.1109/42.906424
 % Segmentation
@@
 \subsubsection{FreeSurfer recon-all}
-% Surface Reconstruction: https://doi.org/10.1006/nimg.1998.0395
+% Surface Reconstruction
 % Refer to the fMRIPrep specific 3 stages workflow.
@@
 % Registration
-% FSL FLIRT: https://doi.org/10.1006/nimg.2002.1132, https://doi.org/10.1016/S1361-8415(01)00036-6, https://doi.org/10.1016/j.neuroimage.2009.06.060
 
@@
 % Head-Motion Correction
-% FSL MCFLIRT: https://doi.org/10.1006/nimg.2002.1132
-
-\subsubsection{fMRIPrep}
-\MD{TODO import fMRIPrep citation boilerplate from "./fmriprep-citations"}
+% Refer to FSL FLIRT
 
@@
 
-The Intel VTune profiler solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level.
+The Intel VTune profiler~\cite{vtune_profiler} solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level.
 However, common profiling challenges remain.
@@
 We discuss our approach to these challenges in the next section.
-\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for refereces}
+% TODO add citation on effect of different input data on numerical stability and reduced precision.
 
 First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
-For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Apptainer image at execution.
-
-\begin{minipage}{\linewidth}
+For use on HPC systems, we created Singularity images using Docker2Singularity \lstlistingname~\ref{vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Singularity image at execution.
+% TODO make sure the algo span one page.
 \lstinputlisting[
 	language={sh},
-		caption={Profiling an application within a Apptainer container, using VTune profiler.},
+	caption={Profiling an application within a Singularity container, using VTune profiler.},
 	captionpos=b,
-		label=alg:vtune_example,
+	label=vtune_example,
 	numbers=left,
@@
 ]{vtune-example.sh}
-\end{minipage}
 
@@
 The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes.
-We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}.
+We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines.
 Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application.
@@
 
-\begin{figure*}[h!]
+% cpu_clk_unhalted.thread: [Core cycles when the thread is not in halt state]
+% cycle_activity.stalls_ldm_pending: [Execution stalls while memory subsystem has an outstanding load]
+% cycle_activity.stalls_l1d_pending: [Execution stalls while L1 cache miss demand load is outstanding]
+% cycle_activity.stalls_l2_pending: [Execution stalls while L2 cache miss demand load is outstanding]
+% MEM_L3_WEIGHT: Approximately the external memory to L3 cache latency ratio
+% MEM_LOAD_UOPS_RETIRED.LLC_HIT: [Retired load uops which data sources were data hits in L3 without snoops required Supports address when precise. Spec update: BDM100 (Precise event)]
+% MEM_LOAD_UOPS_MISC_RETIRED.LLC_MISS: [Miss in last-level (L3) cache. Excludes Unknown data-source Supports address when precise. Spec update: BDM100, BDE70 (Precise event)]
+The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load.
+L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache.
+Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively.
+The store bound occurs when a store operation stall the CPU.
+Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive those metrics.
+\MD{TODO cite intel reference manual and "Intro to Profiling" ch. 6. Should I cite here or in the figure caption?}
+
+The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level; i.e.
+the data is fetched from disk.
+
+\begin{figure*}
 	\noindent
@@
 	\end{minipage}
-	\caption{Intel performance monitoring events and derived memory metrics}
+	\caption{Intel performance monitoring events and derived memory metrics~\MD{TODO cite intel reference manual and "Intro to Profiling" ch. 6}.}
 	\label{fig:memory-metrics}
 \end{figure*}
-
-The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load.
-L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache.
-Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively.
-The store bound occurs when a store operation stall the CPU.
-Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory boudnd characterization~\cite{Intel2006-lc}.
-We use the convention from~\cite{Kukunas2015-jd} for the equations.
-
-The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level; i.e.
-the data is fetched from disk.
 
@@
 \section{Results}
+\MD{TODO Find better titles for the subsections.}
 In this section we present the aggregated profiling data obtained with VTune.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht!]
 	\centering
@@
 Overall, this shows that the input data does not substantially affect the makespan of those pipelines.
-
-\begin{figure}[t]
+\begin{figure}[h]
 	\centering
@@
 }
-\begin{table*}[ht]
+\begin{table*}[h]
 	\centering
@@
 
-\subsection{Main Bottleneck: Linear Interpolation}
+\subsection{Linear Interpolation is the Main Bottleneck}
 Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all.
@@
 In other words, few interpolation functions are contribute to a large portion of the CPU time.
-% There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)
+
+\MD{TODO Check why FreeSurfer recon-all does not use interpolation. Possibly hidden in other functions or using different algorithm.}
 
@@
 
-\subsection{ANTS: Single vs. Double Precision}
+\subsection{Single vs. Double Precision}
 Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision.
 For both pipelines, the makespan is significantly higher in single precision.
-This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv}.
-
+This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision \MD{TODO Cite quadratic of bit-witdh}.
 \begin{figure}[h]
@@
 
+\MD{TODO Discuss the potential reasons for the slowdown. The difference in time for iterations in single and double precision.}
 The number of iteration between both version of ANTS registrationSyN is similar.
@@
 Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.
-
 \begin{figure}
@@
 
-\subsection{FreeSurfer: Thread-Synchronization}
+\subsection{Thread-Synchronization using OpenMP}
 Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant different in the pipeline profile between single-threaded and multi-threaded execution.
@@
 This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.
+\MD{TODO Mention the OpenMP scheduling used in majority.}
 
@@
 
+\MD{TODO}
 %%%%%%%%%% # Linear Interpolation
@@
 % 2. Improvement to interpolation would help the current algorithms, as well as the DL algorithm using CNN.
-\subsection{Importance of Linear Interpolation}
-\MD{TODO}
 
@@
 % 3.
-\subsection{ANTS: A tale of precision}
-\MD{TODO + change title}
 
-\subsection{OpenMP bottleneck in FreeSurfer recon-all}
-In our experiments, FreeSurfer recon-all showed a significant decrease in parallel efficiency with an increase in the number of threads.
-While the Amdhals' law plays an impact on the maximum efficiency, the OpenMP thread synchronization played the largest impact in our results.
-The codebase use 83 \textit{static} scheduling policy out of 92 total.
-\textit{Static} scheduling assign chunks to threads in round-robin fashion.
-This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency.
-We supposed that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads.
-Unfortunately, we failed to recompile FreeSurfer by naively changing the \textit{static} OpenMP scheduling to \textit{dynamic}.
-This remain challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
-
-Future work should study the different configuration of OpenMP scheduling in FreeSurfer recon-all to improve the parallel efficiency.
-We believe that significant performance improvements could be achieved by optimizing the thread synchronization.
-This would lead to faster runtime time and higher CPU usage ration in HPC allocations.
+%%%%%%%%%% OpenOMP bottleneck (FreeSurfer with 32 threads)
+% Hypothesis: OpenMP scheduling is not optimal for the pipeline.
+% - What are the potential issues with the current scheduling policy?
+% Future work: Modify the OpenMP scheduling to improve the parallel efficiency.
 
@@
 Future work could extent the depth of the analysis by including this information.
-\MD{TODO Cite paper on spatial/temporal for RP (ask yohan for references)}
+\MD{TODO Cite paper on spatial/temporal for RP}
 
@@
 \\\href{https://openneuro.org/datasets/ds004513/versions/1.0.2}{https://openneuro.org/datasets/ds004513/versions/1.0.2}.
-
-The container images for our profiling experiments are publicly on Docker Hub:
-\begin{itemize}
-	\item mathdugre/cmake:debug-info
-	\item mathdugre/intel-compilers:debug-info
-	\item mathdugre/afni:debug-info
-	\item mathdugre/ants:debug-info
-	\item mathdugre/fsl:debug-info
-	\item mathdugre/freesurfer:debug-info
-	\item mathdugre/fmriprep:debug-info
-\end{itemize}
-
-Our profiling results from VTune are publicly available at:
-\\\href{URL}{\MD{TODO add Zenodo link}}
+% TODO add compiled containers.
+% TODO add csv files from profiled runs.
+% The above can be distributed on DockerHub and Zenodo respectively.
 
@@
 \\\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
-\MD{TODO Update URL to slashbin org, when created.}
+% TODO Update URL to slashbin org, when created.
 
 \section*{Acknowledgement}
-\MD{TODO}
-% NSERC PGS-D
-% Concordia Bursary (find name)
-% Any from others?
-% ...
+% TODO
 
@@
 \onecolumn
-\section*{Supplementary Materials}
+\section*{Supplementary material}
 \begin{figure*}
@@
 }
+
 % CSV tables
+% ANTS brainExtraction
+% \csvtable{tables/hotspots-1thread-ants-brainExtraction.csv}{ANTS brainExtraction (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-brainExtraction}
+% \csvtable{tables/hotspots-1thread-ants-brainExtraction-fp.csv}{ANTS brainExtraction-fp (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-brainExtraction-fp}
+
+% ANTS registrationSyN
+% \csvtable{tables/hotspots-1thread-ants-registrationSyN.csv}{ANTS registrationSyN (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-registrationSyN}
+% \csvtable{tables/hotspots-1thread-ants-registrationSyN-fp.csv}{ANTS registrationSyN-fp (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-ants-registrationSyN-fp}
+
+% TODO find a way to display the FS table
+FreeSurfer recon-all
 % \csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
@@
 
+% FSL FAST
+% \csvtable{tables/hotspots-1thread-fsl-fast.csv}{FSL FAST (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-fast}
+
+% FSL FLIRT
+% \csvtable{tables/hotspots-1thread-fsl-flirt.csv}{FSL FLIRT (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-flirt}
+
+% FSL MCFLIRT
+% \csvtable{tables/hotspots-1thread-fsl-mcflirt.csv}{FSL MCFLIRT (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-fsl-mcflirt}
+
 \end{document}

### 2024-02-13 run_vtune.sh

@@ -7,4 +7,5 @@ if [[ -z ${1:+x} ]]; then
     exit 1
 fi
 
-vtune-backend --web-port 8080 --allow-remote-access --data-directory $1
+source $HOME/intel/oneapi/setvars.sh --force
+vtune-backend --web-port 8080 --allow-remote-access --data-directory $1 --reset-passphrase

### 2024-02-13 sbatch/fmriprep/anat-fs.sh

@@ -1,61 +1,0 @@
-#!/bin/bash
-
-#SBATCH -J fmriprep-anat-fs
-#SBATCH --array=1
-#SBATCH --time=UNLIMITED
-#SBATCH --exclusive
-#SBATCH --nodes=1
-#SBATCH --ntasks=1
-#SBATCH --cpus-per-task=1
-#SBATCH --mem=0
-# -------------------------
-#SBATCH -o log/%x-%A-%a.out
-#SBATCH -e log/%x-%A-%a.err
-# -------------------------
-set -e
-set -u
-
-# Setup environment and parse args.
-source ./sbatch/pre_run.sh fmriprep anat-fs -j ${SLURM_CPUS_PER_TASK} $@
-
-# Create an empty directory for output
-OUTPUT_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-outputs
-SUBJECT_OUTPUT=${OUTPUT_DIR}/sub-${SUBJECT_ID}
-if [ -d "${SUBJECT_OUTPUT}" ]; then rm -Rf ${SUBJECT_OUTPUT}; fi
-mkdir -p ${SUBJECT_OUTPUT}
-
-# Create WORK_DIR for fMRIPrep
-WORK_DIR=$(dirname ${SLURM_TMPDIR})/fmriprep-${RANDOM_STRING}
-mkdir -p ${WORK_DIR}
-
-SIF_OPTION+=" -B ${SUBJECT_OUTPUT}:/output"
-SIF_OPTION+=" -B ${WORK_DIR}:/workdir"
-export SIF_OPTION+=" -B $HOME/.freesurfer.txt:/opt/freesurfer/.license"
-cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
-FS_LICENSE=/opt/freesurfer/.license
-fmriprep \
-    /data \
-    /output \
-    participant \
-    --work-dir /workdir \
-    --skip_bids_validation \
-    --participant-label ${SUBJECT_ID} \
-    --nthreads ${NTHREAD} \
-    --omp-nthreads ${NTHREAD} \
-    --anat-only
-EOT
-
-export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
-source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
-
-export SKIP_REPORT=1
[…]

### 2024-02-13 sbatch/fmriprep/full.sh

@@ -31,6 +31,7 @@ mkdir -p ${WORK_DIR}
 SIF_OPTION+=" -B ${SUBJECT_OUTPUT}:/output"
 SIF_OPTION+=" -B ${WORK_DIR}:/workdir"
 export SIF_OPTION+=" -B $HOME/.freesurfer.txt:/opt/freesurfer/.license"
+export VTUNE_OPTION+=" -finalization-mode=deferred"
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 FS_LICENSE=/opt/freesurfer/.license
 fmriprep \
@@ -44,7 +45,7 @@ fmriprep \
     --omp-nthreads ${NTHREAD}
 EOT
 
-export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+export APPTAINERENV_OMP_NUM_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2024-03-25 paper/paper.tex
Finished first draft.

--- 2024-02-13 paper/paper.tex
@@
 \begin{abstract}
-	% TODO																
+	MRI pre-processing is a critical step for neuroimaging analysis.
+However, the computational cost of MRI pre-processing pipelines is a major bottleneck for large cohort studies and some clinical applications.
+While HPCs and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers.
+Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.
+Using Intel VTune profiler, we characterize the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes.
+First, we confirmed that few functions contribute to most of the CPU time.
+Second, we found that linear interpolation is the main bottleneck for the applications.
+Data access is also  a substantial bottleneck for the pipelines.
+Last, we found a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all.
+Our results provide a reference for future efforts to optimize MRI pre-processing workflows.
 \end{abstract}
@@
 Therefore, it can lead to under-utilization of the available resources.
-
-%  GPUs
-\MD{TODO GPUs / Accelerators}
 
@@
 
-\subsubsection{ANTS registrationSyN}
-% Spatial Normalization
-% SyN Registration: https://doi.org/10.1016/j.media.2007.06.004
-% TODO Mention the option for single and double precision.
+We use the default parameters for fMRIPrep and most of the sub-workflows. However, for FreeSurfer recon-all, we mimic the fMRIPrep specific 3 stages workflow.
 
-\subsubsection{ANTS brainExtraction}
-% ANTS brainExtraction
-% 1. Truncate image intensity
-% 2. N4 bias correction: https://doi.org/10.1109/TMI.2010.2046908
-% 3. SyN Registration: : https://doi.org/10.1016/j.media.2007.06.004
-% 4. Warp prior to image using affine
-% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/
-% TODO Mention the option for single and double precision.
-% TODO OASIS template citation
-
-\subsubsection{FSL FAST}
-% DOI: https://doi.org/10.1109/42.906424
-% Segmentation
-
-\subsubsection{FreeSurfer recon-all}
-% Surface Reconstruction: https://doi.org/10.1006/nimg.1998.0395
-% Refer to the fMRIPrep specific 3 stages workflow.
-
-\subsubsection{FSL FLIRT}
-% Registration
-% FSL FLIRT: https://doi.org/10.1006/nimg.2002.1132, https://doi.org/10.1016/S1361-8415(01)00036-6, https://doi.org/10.1016/j.neuroimage.2009.06.060
-
-\subsubsection{FSL MCFLIRT}
-% Head-Motion Correction
-% FSL MCFLIRT: https://doi.org/10.1006/nimg.2002.1132
-
-\subsubsection{fMRIPrep}
-\MD{TODO import fMRIPrep citation boilerplate from "./fmriprep-citations"}
+We profile both the single and double precision version for ANTs registrationSyN and ANTs brainExtraction. ANTs uses ITK templates to implement the single precision alternative.
 
@@
 We discuss our approach to these challenges in the next section.
-\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for refereces}
+\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for references}
 
@@
 The store bound occurs when a store operation stall the CPU.
-Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory boudnd characterization~\cite{Intel2006-lc}.
+Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}.
 We use the convention from~\cite{Kukunas2015-jd} for the equations.
@@
 Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e.
-makspan.
+makespan.
 There is a significant difference between the makespan of different applications.
@@
 
-%%%%%%%%%% # Linear Interpolation
-% Theoretical improvements using reduced precision. If time permits, toy experiment with ITK single precision data type at different locations.
-% 1. Winograd algorithm for fast convolution.
-% - It could be used to speed up the current pipeline. Tradeoff with precision? How does it compare to reduced precision?
-% 2. Improvement to interpolation would help the current algorithms, as well as the DL algorithm using CNN.
 \subsection{Importance of Linear Interpolation}
-\MD{TODO}
+Our results show that linear interpolation the primary bottleneck for all pipelines, except FreeSurfer recon-all.
+While each interpolation have low computational cost, the large amount of operations make it a significant bottleneck.
 
-%%%%%%%%%% # Single vs. Double Precision
-% 1. Not as simple as using different data type. It could affect
-% - Convergence
-% - Accuracy
-% - Runtime (maybe even slower)
-% 1.5 Convergence is the sane in our case (same number of iterations).
-% 2. Why is it slower in our case?
-% -? Possibly less optimized code for single precision.
-% -? Possibly memory placement
-% -? ...
-% 3. 
+There is an extensive literature on interpolation.
+Unfortunately, large amount of optimization techniques are not applicable to MRI pre-processing pipelines, because of the rotation during registration which requires dynamic kernels for interpolation.
+Moreover, the high-resolution of the data and iterative refinement of the registration require a large number of interpolations to be performed.
+
+The authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation.
+The review algorithm trade-off accuracy for a lower computational cost.
+We think it would be interesting for future work to explore these different alternatives to accelerate interpolation within the pipelines.
+However, special attention will be required to balance the accuracy and the performance of the algorithm.
+
+Similarly, we propose future efforts to explore reduced- or mixed-precision techniques to accelerate interpolation.
+Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements.
+
+Improvements to interpolation in 3D MRI imaging would have a large impact.
+It would also benefit other fields of studies using interpolation for high-resolution imaging with iterative refinement, such as climate simulation, astronomical imaging, or 3D image reconstruction.
+
 \subsection{ANTS: A tale of precision}
-\MD{TODO + change title}
+From our results, we observe a slow-down for the single precision version of ANTs pipelines.
+A single iteration takes longer to execute when using single-precision compared to double precision.
+After a further analysis of the ANTs registrationSyN single-precision pipeline, we found that the function\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} contributing most of the CPU time does computations in double precision.
+Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast to back to float in ANTS.
+We think this bug is the cause of the slow-down.
+We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base.
+This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base.
+Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
 
@@
 This would lead to faster runtime time and higher CPU usage ration in HPC allocations.
+
+\subsection{VTune results finalization}
+Using VTune to collect profiling results does not add a large overhead in term of time and resources utilization.
+However, a finalization step is required to query the data and generate a VTune report.
+We found this step to be both time consuming and resource intensive.
+For example, finalizing the results from a single FreeSurfer recon-all execution requires up to \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline.
+This significant difference in resource requirement creates a challenge to profile pipeline with long runtime, due to the potential difficulty to obtain access to compute nodes with enough RAM.
+This can be further challenging for researchers in early career or in developing countries, which might not have access to large HPC infrastructure.
+One way to mitigate this challenge is to lower the sampling rate during the profiling.
+However, to reach a similar level of memory consumption, the sampling rate must be significantly reduced.
+This would lead to a potential loss of information in the profiling results.
 
@@
 \section{Conclusion}
-\MD{TODO}
+In this paper, we profiled several MRI pre-processing applications: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep.
+We confirmed our hypothesis that few functions contribute to a large amount of CPU time.
+We found that most applications suffered from memory bound.
+Linear interpolation was the primary bottleneck for these applications.
+We discovered a bug in ITK which lead the implementation of ANTS registration in single-precision to use float for its computation.
+We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP.
+Last, we discuss the challenge of profiling long running applications with VTune, due to computational resource requirements to obtain results.
+
+We suggest that future effort focus on reduced precision techniques, which could reduce memory bound and accelerate interpolation computation.
+We note that the use of reduced precision is not trivial, as seen with the ANTS bug.
+Therefore, careful attention would be required to find a balance between performance and accuracy, while ensuring bugs are not introduced.
+We also suggest that future work should study the different configuration of OpenMP scheduling in FreeSurfer recon-all to improve the parallel efficiency.
+This would lead to faster runtime time and higher CPU usage ration in HPC allocations.
+
+We hope that this work serves as a reference for future work to optimize MRI pre-processing pipelines.
 
@@
 \\\href{URL}{\MD{TODO add Zenodo link}}
+% Waiting on data transfer from Compute Canada
 
@@
 \section*{Acknowledgement}
-\MD{TODO}
-% NSERC PGS-D
-% Concordia Bursary (find name)
-% Any from others?
-% ...
+Mathieu Dugr'e was jointly funded by the Natural Sciences and Engineering Research Council of Canada (NSERC) through the Postgraduate Scholarship-Doctoral (PGS-D) program and the Graduate Doctoral Fellowship Award from Concordia University.
+\MD{Q: Any other?}
 
@@
 \section*{Supplementary Materials}
-\begin{figure*}
+\MD{Not sure if these add value?}
+\label{sec:supplementary}
+
+\begin{figure*}[h]
 	\centering
@@
 \end{figure*}
-
-\label{sec:supplementary}
 

### 2024-03-25 container/build.sh

@@ -1,0 +1,16 @@
+#!/bin/sh
+set -e
+set -u
+
+# Compilers
+docker build -t mathdugre/cmake:debug-info -f cmake.Dockerfile .
+docker build -t mathdugre/intel-compilers:debug-info -f intel-compilers.Dockerfile .
+
+# Sub-workflows
+docker build -t mathdugre/afni:debug-info -f afni.Dockerfile .
+docker build -t mathdugre/ants:debug-info -f ants.Dockerfile .
+docker build -t mathdugre/fsl:debug-info -f cmake.Dockerfile .
+docker build -t mathdugre/freesurfer:debug-info -f freesurfer.Dockerfile .
+
+# fMRIPrep
+docker build -t mathdugre/fmriprep:debug-info -f fmriprep.Dockerfile .

### 2024-03-25 sbatch/afni/3dTshift.sh

@@ -25,7 +25,7 @@ cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
     /data/sub-${SUBJECT_ID}/ses-open/func/sub-${SUBJECT_ID}_ses-open_task-rest_bold.nii.gz
 EOT
 
-export SINGULARITYENV_OMP_NUM_THREADS=$NTHREAD
+export APPTAINERENV_OMP_NUM_THREADS=$NTHREAD
 source ./sbatch/vtune.sh bash ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 
 source ./sbatch/post_run.sh

### 2024-03-25 sbatch/generate_vtune_report.sh

@@ -1,0 +1,40 @@
+#!/bin/bash
+
+#SBATCH -J generate-vtune-summary
+#SBATCH --array=1
+#SBATCH --time=4:00:00
+#SBATCH --nodes=1
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem=4G
+#SBATCH --partition=gpu
+# -------------------------
+#SBATCH -o log-makespan/%x-%A-%a.out
+#SBATCH -e log-makespan/%x-%A-%a.err
+# -------------------------
+set -e
+set -u
+
+[[ -z ${SLURM_ARRAY_TASK_ID:+x} ]] && SLURM_ARRAY_TASK_ID=1
+SIF_IMG=$1
+TAR_FILE=$(sed -n ${SLURM_ARRAY_TASK_ID}p $2)
+
+PARENT_FOLDER=$(dirname $TAR_FILE)
+FILENAME=$(basename $TAR_FILE)
+EXTRACTED_FOLDER=${FILENAME%%.*}
+
+echo "[INFO] Fixing VTune report: $TAR_FILE"
+cd $PARENT_FOLDER
+tar xzf $FILENAME
+# Generate VTune report
+singularity exec --cleanenv \
+    -B ~/intel/oneapi/vtune/latest/:/vtune \
+    -B $PARENT_FOLDER:$PARENT_FOLDER \
+    $SIF_IMG \
+    /vtune/bin64/vtune \
+    -report summary \
+    -r $PARENT_FOLDER/$EXTRACTED_FOLDER \
+    -report-output $PARENT_FOLDER/$EXTRACTED_FOLDER.makespan \
+    -format csv \
+    -csv-delimiter tab
+rm -r $EXTRACTED_FOLDER

### 2024-03-27 paper/paper.tex

@@
 \title{An Analysis of Performance Bottlenecks in MRI Pre-Processing}
-\author{Mathieu Dugr\'e, Yohan Chatelain, Tristan Glatard}
+\author{Mathieu Dugr\'e, Yohan Chatelain, Tristan Glatard\\Department of Computer Science and Software Engineering, Concordia University, Montreal, QC, Canada}
 
@@
 \begin{abstract}
-	MRI pre-processing is a critical step for neuroimaging analysis.
+	Magnetic Resonance Image (MRI) pre-processing is a critical step for neuroimaging analysis.
 However, the computational cost of MRI pre-processing pipelines is a major bottleneck for large cohort studies and some clinical applications.
-While HPCs and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers.
-Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.
+While High-Performance Computing (HPC) and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers.
+Therefore, it is critical to understand the performance bottlenecks of MRI pre-processing pipelines to improve their performance.
 Using Intel VTune profiler, we characterize the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes.
-First, we confirmed that few functions contribute to most of the CPU time.
-Second, we found that linear interpolation is the main bottleneck for the applications.
-Data access is also  a substantial bottleneck for the pipelines.
-Last, we found a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all.
+We found that few functions contribute to most of the CPU time, and that linear interpolation is the largest contributor.
+Data access is also  a substantial bottleneck.
+Last, we identified a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all.
 Our results provide a reference for future efforts to optimize MRI pre-processing workflows.
@@
 Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed.
-For those reasons, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.
+Therefore, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.
 
 % HPCs
-Researcher broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts.
+\TG{Paragraphs 2 to 5 could be condensed in a more concise way.}
+Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts.
 HPC systems provide a large number of CPU cores and high-bandwidth memory to process subjects in parallel.
-However, HPCs might not be available to researchers in developing country.
+However, HPCs might not be available to researchers in developing countries \TG{revise that}.
 Similarly, cloud computing allows scaling the number of core and memory to process large cohort.
-However, cloud computing requires expertise to setup and can be costly for large cohort.
+However, cloud computing requires expertise to setup and can be costly for large cohorts.
 Moreover, data transfer can be slow and might be impossible due to ethic concerns related to human data.
@@
 In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing.
-For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} with the effort to provide Deep Learning alternative to traditional algorithms.
+For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} to provide Deep Learning alternatives to traditional algorithms.
 The main benefit of those tools is the faster runtime they provide.
-Moreover, A study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
+% Moreover, a study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
 However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive.
@@
 % Alternative solutions
-While those techniques can provide significant performance improvements, their additional hardware requirement can be costly and unaccessible to some researchers.
+While those techniques can provide significant performance improvements, their additional hardware requirements can be costly and unaccessible to some researchers.
 Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements.
@@
 % - No comprehensive study of the performance bottleneck of MRI pre-processing.
-The neuroimaging community is well aware that MRI pre-precessing pipelines require long computation time.
+The neuroimaging community is well aware that MRI pre-processing pipelines require long computation time.
 However, to the best of our knowledge, there is no comprehensive study of the performance bottlenecks in MRI pre-processing.
-In this paper, we characterize computational cost of several commonly adopted MRI pre-processing pipelines.
+In this paper, we characterize the computational cost of several commonly adopted MRI pre-processing pipelines.
 The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.
@@
 % - Focus on MRI pre-processing.
-Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain.
-Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI).
-While other modalities such as electroencephalogram (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRIs for their broad adoption and non-invasive property.
+MRI is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain.
+Three main MRI modalities exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI).
+While other modalities such as electroencephalography (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRIs for their broad adoption and non-invasiveness.
 Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.
@@
 Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.
+\TG{Move this to methods}
 
@@
 The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level.
-We profile the application using two threading configuration: single-threaded and 32 threads.
+We profile the application using two threading configuration: single-threaded and 32 threads \TG{add justification}.
 We aggregate subjects' results for each pipeline to obtain an average performance profile.
@@
 \subsection{Pipelines}
-In this paper, we study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow.
+We study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow.
 First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance.

### 2024-03-27 paper/paper.tex
[FIX] comments from Tristan.

@@
 \usepackage{cite}
+\usepackage{graphicx}
 \usepackage{xcolor}
@@
 \hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
-\usepackage{graphicx}
 \graphicspath{{figures}}
@@
 \usepackage{amsmath}
-\usepackage{import}
 
+\usepackage{color,soul}
 \newcommand{\TG}[1]{\color{blue}\textsc{From Tristan: }#1\color{black}}
@@
 Although established pre-processing pipelines exist, they commonly require expensive computation and produce large output and intermediate data.
-This has been a major bottleneck to study large subject cohorts in neuroimaging.
-Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed.
+\HL{These are a well-known challenges to study large subject cohorts in neuroimaging.} Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed.
 Therefore, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.
@@
 % HPCs
-\TG{Paragraphs 2 to 5 could be condensed in a more concise way.}
-Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts.
-HPC systems provide a large number of CPU cores and high-bandwidth memory to process subjects in parallel.
-However, HPCs might not be available to researchers in developing countries \TG{revise that}.
-Similarly, cloud computing allows scaling the number of core and memory to process large cohort.
-However, cloud computing requires expertise to setup and can be costly for large cohorts.
-Moreover, data transfer can be slow and might be impossible due to ethic concerns related to human data.
-In both cases, the benefits from parallelization are limited by the parallel efficiency of the pipeline and the number of subjects to process.
-Therefore, it can lead to under-utilization of the available resources.
+Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts \HL{leveraging} a large number of CPU cores and high-bandwidth memory to process subjects in parallel.
+However, \HL{it can be challenging to run HPCs in developing countries due to power restrictions, national network infrastructure, and lack of training and funding}~\cite{NRG-SC23}.
+\HL{Similarly, researcher use cloud computing, but it requires expertise to setup, can be costly for large studies, and} data transfer can be slow and might be impossible due to ethic concerns related to human data.
+\HL{In both cases, limitation from the parallel efficiency of the pipeline can lead to under-utilization of the available resources.}
 
@@
 In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing.
-For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} to provide Deep Learning alternatives to traditional algorithms.
-The main benefit of those tools is the faster runtime they provide.
+For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} to provide Deep Learning alternatives to traditional algorithms\HL{, providing faster runtime.}
 % Moreover, a study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
 However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive.
-Additionally, the results from DL models can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.
+Additionally, \HL{their results} can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.
 
 % Alternative solutions
-While those techniques can provide significant performance improvements, their additional hardware requirements can be costly and unaccessible to some researchers.
-Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements.
-However, their implementation can require substantial effort.
+\HL{These techniques require costly hardware to be effective, which can be unacessible to underrepresented researchers.} Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements but require substantial effort to implement.
 Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.
@@
 % - No comprehensive study of the performance bottleneck of MRI pre-processing.
-The neuroimaging community is well aware that MRI pre-processing pipelines require long computation time.
-However, to the best of our knowledge, there is no comprehensive study of the performance bottlenecks in MRI pre-processing.
-In this paper, we characterize the computational cost of several commonly adopted MRI pre-processing pipelines.
+\HL{To the best of our knowledge, this is the first study that} characterize the computational cost of several commonly adopted MRI pre-processing pipelines.
 The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.
@@
 % - data: OpenNeuro ds004513
-We use the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
+% We use the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
 We chose this dataset for its wide range of age and equal distribution of biological sex.
@@
 The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level.
-We profile the application using two threading configuration: single-threaded and 32 threads \TG{add justification}.
+We profile the application using \HL{-- single-threaded and 32 threads to simplify the analysis and study the the multi-threading performance, repsectively}.
 We aggregate subjects' results for each pipeline to obtain an average performance profile.
@@
 
-\begin{figure*}[h!]
+\begin{figure*}[ht]
 	\noindent
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\centering
@@
 }
-\begin{table*}[h]
+\begin{table*}[ht]
 	\centering
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\includegraphics[width=\linewidth]{figures/makespan-ants.png}
@@
 
-\begin{figure*}[h]
+\begin{figure*}[ht]
 	\centering
@@
 \newcommand{\csvtable}[3]{
-	\begin{table}[h]
+	\begin{table}[ht]
 		\resizebox*{\textwidth}{!}{

### 2024-03-28 paper/paper.tex
Additional comments

--- 2024-03-27 paper/paper.tex
@@
 We study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow.
-First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance.
-Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}.
-We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset.
-Moreover, the statistical analysis of the fMRIPrep workflow are not computationally expensive and are not the focus of this paper.
+First, we profile the entire workflow to obtain a coarse view of the workflow's performance.
+Furthermore, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}.
+We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset \TG{add a few words to explain why}.
+\TG{Information is missing to explain what is the statistical analysis in fMRIPrep.
+Likewise, it's unclear what ANTS registration, FSL Flirt, etc are and how they are combined in fMRIPrep.
+This section should start with a higher-level overview of fMRIPrep.
+You should also explain that fMRIPrep's building blocks are widely used in neuroimaging.} Moreover, the statistical analysis of the fMRIPrep workflow is not computationally expensive and is not the focus of this paper.
 
-We use the default parameters for fMRIPrep and most of the sub-workflows. However, for FreeSurfer recon-all, we mimic the fMRIPrep specific 3 stages workflow.
+We use the default parameters for fMRIPrep and most of the sub-workflows.
+However, for FreeSurfer recon-all, we mimic the fMRIPrep specific 3 stages workflow \TG{This is cryptic, it needs more context.}.
 
-We profile both the single and double precision version for ANTs registrationSyN and ANTs brainExtraction. ANTs uses ITK templates to implement the single precision alternative.
+We profile both the single and double precision version for ANTs registrationSyN and ANTs brainExtraction.
+ANTs uses ITK templates to implement the single precision alternative.
+\TG{You should re-order this paragraph as follows: (1) ANTs registration has two implementations, a double precision one and a single precision one.
+(2) This is actually coming from ITK.
+(3) We benchmark both implementations.}
 
 \subsection{Data}
-To ensure the performance profiles of the applications derived from our work are inclusive of different populations, we used data with a wide range of age and equal distribution of sex (biological).
-Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.
-
+To ensure that our performance profiles are inclusive of different populations, we used data with a wide range of age and equal distribution of sex.
 We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
@@
 The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session.
-We profiled the application using both cohort on the eyes open session, unless explicitly specified.
+We profiled the application using both cohorts and the eyes open session, unless explicitly specified \TG{is it ``explicitly specified'' anywhere?
+If not, remove this.}.
 
 \subsection{Profiling}
-Profiling MRI pre-processing pipelines present a set of challenges.
-The neuroimaging community develops pipelines using several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages.
+Profiling MRI pre-processing pipelines \TG{the terms pipelines, workflfows, applications are used interchangeably, let's harmonize and use only 1 term (I don't mind which one)} raises a few challenges.
+Neuroimaging pipelines use several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages.
 Pipelines are generally complex and computationally expensive.
 
-The Intel VTune profiler solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level.
+The Intel VTune profiler addresses these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level.
 However, common profiling challenges remain.
-Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules.
-Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds.
+Profilers require source code to be compiled with debug symbols to report interpretable information about the different functions and modules.
+Moreover, different input data can substantially impact the performance results due to conditional branching in the application and convergence thresholds.
 We discuss our approach to these challenges in the next section.
@@
 First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
-For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Apptainer image at execution.
+For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} \TG{Could this be called a listing rather than an algorithm?} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.
 
@@
 We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}.
-Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application.
+Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application \TG{This discussion on Pareto principle should be moved to the results section}.
 This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
@@
 Last, we profile the application using two threading configuration: single-threaded and 32 threads.
-On the one hand, the single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling.
-On the other hand, we use all available CPU cores from a compute node (i.e.
-32) to study the multi-threading performance of the application.
+The single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling.
+The 32-thread approach uses all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the application.
 This is a more realistic usage for the user of the application.
@@
 The data generated by the VTune profiler is written to a shared file system, for administrative reasons.
-Theoretically, this increase the profiling overhead.
-However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as the data from applications.
+In principle, this could increase the profiling overhead.
+However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as data produced by the applications.
 
 \subsection{Metrics Definition}
-We define the makespan as the elapsed time from the start to the end of the application.
-The CPU time refers to the total time spent by the CPU, across all core, to execute the application.
 VTune provides several metrics to characterize the performance of an application.
-We describe the memory metrics used in this paper.
-
+The makespan is the elapsed time from the start to the end of the application.
+The CPU time refers to the total time spent by the CPU, across all cores, to execute the application.
 \begin{figure*}[h!]
@@
 \end{figure*}	
-
 The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load.
@@
 The store bound occurs when a store operation stall the CPU.
-Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}.
+Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc} \TG{This figure should go to appendix, it's not critical to the understanding of the paper}.
 We use the convention from~\cite{Kukunas2015-jd} for the equations.
 
-The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level; i.e.
-the data is fetched from disk.
+The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level, that is, when the data is being fetched from disk.
 
 \subsection{Infrastructure}
-For our experiments, we used the \textit{Slashbin cluster} at Concordia University.
+We used the \textit{Slashbin cluster} at Concordia University.
 The compute nodes are configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), Rocky~Linux~8.9 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}.
@@
 First, we present a performance overview of the pipelines.
-We then dive into more detailed analysis observed from the hotspots of the pipelines.
+We then analysis specific hotspots in the pipelines.
 Unless specified explicitly, we only present the profile data from using 1-thread since the results from 32-threads are similar.
 
-\subsection{Long-tail Distribution}
-Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time is spent in less than 1.5\% of the function across all pipelines.
-While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a pareto distribution.
-We plotted the distribution of the functions' CPU time on a log-log scale and found it to be closer to a Zipf's law distribution with exponent $\alpha=1.8$.
+\subsection{CPU time distribution across functions was long tailed}
+Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time was spent in less than 1.5\% of the function across all pipelines.
+While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a Pareto distribution.
+We plotted the distribution of the functions' CPU time on a log-log scale and found it to be closer \TG{``closer'' is too vague: is it well approximated by a Zipf's law?
+Can you provide a p-value of this claim?} to a Zipf's law distribution with exponent $\alpha=1.8$.
+\TG{Add a few sentences to explain why this is interesting.}
 
@@
 The x-axis percentage of functions ordered by decreasing CPU time.
-The data includes all functions from all pipelines.}
+The data includes all functions from all pipelines.
+}
 	\label{fig:long-tail-distribution}
@@
 Overall, this shows that the input data does not substantially affect the makespan of those pipelines.
+\TG{not sure what to make from this result, I don't find it particularly noticeable, I think I would omit it.}
 
@@
 \subsection{Impact of Memory Bounded Functions}
-Table~\ref{extab:memory-bound} shows that fetching data from disk is the main source of cycles where the CPU is starved.
-Following, The L1 bound has the highest cache-level contribution to the memory bound of the pipelines.
+Table~\ref{extab:memory-bound} shows that fetching data from disk is the main source of cycles where the CPU is starved \TG{I don't see disk in table 1}.
+Following, the L1 bound has the highest cache-level contribution to the memory bound of the pipelines.
 ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT.
@@
 We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.
+\TG{Add discussion: what is interesting about this result?}
 
@@
 In other words, few interpolation functions are contribute to a large portion of the CPU time.
+\TG{This should be moved up in the results, I think right after Figure 2, as this is one of the most interesting results.} \TG{Add a few sentences to explain why this is interesting.}
 % There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)
@@
 Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.
+\TG{This is too inconclusive, you should give all the details here.}
 
@@
 \subsection{FreeSurfer: Thread-Synchronization}
-Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant different in the pipeline profile between single-threaded and multi-threaded execution.
-The multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime.
+Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded execution.
+In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime.
 FreeSurfer uses the multi-processing API provided by OpenMP.
@@
 
-While the makespan for FreeSurfer recon-all decrease with an increase in the number of threads, the benefits are limited (Figure~\ref{fig:freesurfer-threading}).
+While the makespan for FreeSurfer recon-all decreases with an increase in the number of threads, the benefits are limited (Figure~\ref{fig:freesurfer-threading}).
 The parallel efficiency decreases from 66\% with two threads down to 5\% with 32 threads.
-While the Amdhals' law plays an impact on the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} indicates that the thread Synchronization has a larger impact.
+While Amdhals' law plays an impact on the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} indicates that the thread Synchronization has a larger impact.
 
@@
 \section{Discussion}
+
+\TG{I think the discussion should be merged in the results.
+I know it's unconventional but I think it would make the paper easier to read and more interestingly.
+You could add a short conclusion that summarizes the main take-aways from the results.}
+
 \subsection{Few functions account for most of the CPU time}
@@
 Thus, 20\% of the functions would contribute to 80\% of the total CPU time.
-However, Figure~\ref{fig:long-tail-distribution} shows that the distribution of CPU time is closer to a Zipf's law with exponent $\alpha=1.8$.
+However, Figure~\ref{fig:long-tail-distribution} shows \TG{Figure 2 doesn't ``show'' that.
+Rephrase.} that the distribution of CPU time is closer to a Zipf's law with exponent $\alpha=1.8$.
 This means that the CPU time of a function is approximately inversely proportional to its rank.
@@
 In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
+\TG{You could cite some of Valerie's works in this direction.}
 
@@
 
-There is an extensive literature on interpolation.
-Unfortunately, large amount of optimization techniques are not applicable to MRI pre-processing pipelines, because of the rotation during registration which requires dynamic kernels for interpolation.
+There is an extensive literature on interpolation \TG{cite it}.
+Unfortunately, large amount of optimization techniques are not applicable to MRI pre-processing pipelines, because of the rotation during registration \TG{explain how interpolation is related to registration.
+In your presentation of fmriprep, explain what registration is and the role it plays in the pipelines.} \TG{this is a bit cryptic and I don't think you want to state that here.
+You may want to say that it would be interesting to further investigate these techniques in the future.} which requires dynamic kernels for interpolation.
 Moreover, the high-resolution of the data and iterative refinement of the registration require a large number of interpolations to be performed.
@@
 The authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation.
-The review algorithm trade-off accuracy for a lower computational cost.
+They review algorithm trade-off accuracy for a lower computational cost.
 We think it would be interesting for future work to explore these different alternatives to accelerate interpolation within the pipelines.
@@
 It would also benefit other fields of studies using interpolation for high-resolution imaging with iterative refinement, such as climate simulation, astronomical imaging, or 3D image reconstruction.
+\TG{You could summarize the previous 3 paragraphs, I find them a bit verbose.}
 
@@
 Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
+\TG{Can you report the bug and add a link to the issue here?}
 
@@
 In our experiments, FreeSurfer recon-all showed a significant decrease in parallel efficiency with an increase in the number of threads.
-While the Amdhals' law plays an impact on the maximum efficiency, the OpenMP thread synchronization played the largest impact in our results.
-The codebase use 83 \textit{static} scheduling policy out of 92 total.
+While Amdhals' law plays an impact on the maximum efficiency, the OpenMP thread synchronization played the largest impact in our results.
+The codebase uses 83 \textit{static} scheduling policy out of 92 total \TG{explain what these numbers represent}.
 \textit{Static} scheduling assign chunks to threads in round-robin fashion.
 This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency.
-We supposed that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads.
+We speculated that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads.
 Unfortunately, we failed to recompile FreeSurfer by naively changing the \textit{static} OpenMP scheduling to \textit{dynamic}.
-This remain challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
+This remains challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
+\TG{Could you file a bug report in freesurfer on this issue and refer to it here?}
 
@@
 \subsection{Limitations}
+\TG{You should frame the limitations as future work in the conclusion. These are not really limitations.}
 The dataset we chose contains wide range of age and equal distribution of sex.
@@
 Future work should include data from multiple sites to ensure the generalization of the performance bottleneck we found.
+\TG{Is there evidence showing that this would provide additional insights on the benchmark?
+I think it might be interesting to study subjects that fail QC but I'm not sure about other populations.}
 
-Our analysis omits the space or time of execution for function calls.
+Our analysis omits the space or time of execution for function calls \TG{I don't understand what this means}.
 This information could provide additional insight for optimization.
@@
 In this paper, we profiled several MRI pre-processing applications: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep.
-We confirmed our hypothesis that few functions contribute to a large amount of CPU time.
+We confirmed our hypothesis \TG{don't frame it as a hypothesis, it is an observation} that few functions contribute to a large amount of CPU time \TG{, which [explain why it is interesting to know]}.
 We found that most applications suffered from memory bound.

### 2024-04-02 paper/paper.tex

@@
 The CPU time refers to the total time spent by the CPU, across all cores, to execute the application.
-\begin{figure*}[h!]
+
+\begin{figure*}[ht]
 	\noindent

### 2024-04-17 paper/paper.tex
[FIX] comments from Tristan.

@@
 \usepackage{listings}
-\renewcommand{\lstlistingname}{Algorithm}
+\renewcommand{\lstlistingname}{Listing}
 \usepackage[binary-units=true]{siunitx}
@@
 Last, we identified a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all.
-Our results provide a reference for future efforts to optimize MRI pre-processing workflows.
+Our results provide a reference for future efforts to optimize MRI pre-processing pipelines.
 \end{abstract}
@@
 \HL{To the best of our knowledge, this is the first study that} characterize the computational cost of several commonly adopted MRI pre-processing pipelines.
-The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.
+The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing pipelines.
 
@@
 Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data.
-We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} workflow as it a comprehensive pre-processing pipeline for both sMRI and fMRI.
-fMRIPrep combines several pipelines to produce a complete pre-processing workflow.
-We also study its sub-workflow separately to provide a finer grain analysis of the performance bottleneck.
-
-% - data: OpenNeuro ds004513
-% We use the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
-We chose this dataset for its wide range of age and equal distribution of biological sex.
-Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.
-\TG{Move this to methods}
+We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} pipeline as it a comprehensive pre-processing application for both sMRI and fMRI.
+fMRIPrep combines several pipelines to produce a complete pre-processing pipeline.
+We also study its sub-pipeline separately to provide a finer grain analysis of the performance bottleneck.
 
@@
 The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level.
-We profile the application using \HL{-- single-threaded and 32 threads to simplify the analysis and study the the multi-threading performance, repsectively}.
+We profile the pipeline using \HL{-- single-threaded and 32 threads to simplify the analysis and study the the multi-threading performance, repsectively}.
 We aggregate subjects' results for each pipeline to obtain an average performance profile.
-We compiled each application with debug information to obtain human readable information on function and module names in VTune.
+We compiled each pipeline with debug information to obtain human readable information on function and module names in VTune.
 
@@
 \subsection{Pipelines}
-We study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow.
-First, we profile the entire workflow to obtain a coarse view of the workflow's performance.
-Furthermore, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}.
-We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset \TG{add a few words to explain why}.
-\TG{Information is missing to explain what is the statistical analysis in fMRIPrep.
-Likewise, it's unclear what ANTS registration, FSL Flirt, etc are and how they are combined in fMRIPrep.
-This section should start with a higher-level overview of fMRIPrep.
-You should also explain that fMRIPrep's building blocks are widely used in neuroimaging.} Moreover, the statistical analysis of the fMRIPrep workflow is not computationally expensive and is not the focus of this paper.
 
-We use the default parameters for fMRIPrep and most of the sub-workflows.
-However, for FreeSurfer recon-all, we mimic the fMRIPrep specific 3 stages workflow \TG{This is cryptic, it needs more context.}.
+fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, which is built on top of several well-known neuroimaging toolboxes such as ANTs, FSL, FreeSurfer and AFNI.
+Although the processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity corretion and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain.
+When enabled, FreeSurfer recon-all~\cite{Dale1999-wu} is used for surface reconstruction.
+We mimic the fMRIPrep specific 3 \HL{phases reconall pre-processing, which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3}.
+The fMRI pipeline generally performs motion correction with FSL MCFLIRT~\cite{Jenkinson2002-od}, slice timing correction with AFNI 3dtshift~\cite{Cox1996-nk}, and co-registration with FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}.
 
-We profile both the single and double precision version for ANTs registrationSyN and ANTs brainExtraction.
-ANTs uses ITK templates to implement the single precision alternative.
-\TG{You should re-order this paragraph as follows: (1) ANTs registration has two implementations, a double precision one and a single precision one.
-(2) This is actually coming from ITK.
-(3) We benchmark both implementations.}
+We profile the entire fMRIPrep pipeline to obtain a coarse view of the pipeline's performance bottleneck.
+Additionlly, we profile the sub-pipelines of fMRIPrep to characterize the performance at a finer grain.
+While we used the same sub-pipeline and order as in fMRIPrep, we used the default parameters for most of the sub-pipelines which may vary from the default parameters of fMRIPrep.
+We omitted some sub-pipelines of fMRIPrep as they were not compatible with our dataset.
+For example, time-slicing correction was already performed.
+
+\HL{ANTs brainExtraction and registrationSyN provides the user with the option to use a single or double precision version of the pipeline, leveraging the Insight Segmentation and Registration Toolkit (ITK)}~\cite{Yoo2002-ve}.
+\HL{We profile both version to understand the performance benefits of using reduced precision in the pipelines.}
 
@@
 The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session.
-We profiled the application using both cohorts and the eyes open session, unless explicitly specified \TG{is it ``explicitly specified'' anywhere?
-If not, remove this.}.
 
 \subsection{Profiling}
-Profiling MRI pre-processing pipelines \TG{the terms pipelines, workflfows, applications are used interchangeably, let's harmonize and use only 1 term (I don't mind which one)} raises a few challenges.
+Profiling MRI pre-processing pipelines raises a few challenges.
 Neuroimaging pipelines use several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages.
@@
 Profilers require source code to be compiled with debug symbols to report interpretable information about the different functions and modules.
-Moreover, different input data can substantially impact the performance results due to conditional branching in the application and convergence thresholds.
+Moreover, different input data can substantially impact the performance results due to conditional branching in the pipeline and convergence thresholds.
 We discuss our approach to these challenges in the next section.
@@
 First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
-For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} \TG{Could this be called a listing rather than an algorithm?} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.
+For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{lst:vtune_example} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.
 
@@
 		language={sh},
-		caption={Profiling an application within a Apptainer container, using VTune profiler.},
+		caption={Profiling a pipeline within a Apptainer container, using VTune profiler.},
 		captionpos=b,
-		label=alg:vtune_example,
+		label=lst:vtune_example,
 		numbers=left,
@@
 
-Second, we aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile.
-The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes.
-We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}.
-Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application \TG{This discussion on Pareto principle should be moved to the results section}.
-This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
-We report the mean and standard deviation of the execution time for those functions.
+Last, we profile the pipeline using two threading configuration: single-threaded and 32 threads.
+The single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling.
+The 32-thread approach uses all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the pipeline.
+This is a more realistic usage for the user of the pipeline.
 
-Last, we profile the application using two threading configuration: single-threaded and 32 threads.
-The single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling.
-The 32-thread approach uses all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the application.
-This is a more realistic usage for the user of the application.
-
-Before profiling each application, we transfer the input data from the shared file system to the compute node.
+Before profiling each pipeline, we transfer the input data from the shared file system to the compute node.
 After profiling, we transfer back the output to the shared file system.
@@
 In principle, this could increase the profiling overhead.
-However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as data produced by the applications.
+However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as data produced by the pipelines.
 
 \subsection{Metrics Definition}
-VTune provides several metrics to characterize the performance of an application.
-The makespan is the elapsed time from the start to the end of the application.
-The CPU time refers to the total time spent by the CPU, across all cores, to execute the application.
+VTune provides several metrics to characterize the performance of a pipeline.
+The makespan is the elapsed time from the start to the end of the pipeline.
+The CPU time refers to the total time spent by the CPU, across all cores, to execute the pipeline.
 
-\begin{figure*}[ht]
-	\noindent
-	\begin{minipage}
-		{0.5\textwidth}
-		\begin{align*}
-			C & = CPU\_CLK\_UNHALTED.THREAD                \\
-			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
-			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
-			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
-			W & = MEM\_L3\_WEIGHT                          \\
-			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
-			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
-			M & = \frac{W \times R}{H + R}                 \\
-			N & = \frac{H}{H + W \times R}                 
-		\end{align*}
-	\end{minipage}
-	\begin{minipage}
-		{0.5\textwidth}
-		\begin{align}
-			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
-			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
-			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
-			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
-			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
-		\end{align}
-	\end{minipage}
-	\caption{Intel performance monitoring events and derived memory metrics}
-	\label{fig:memory-metrics}
-\end{figure*}	
 The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load.
@@
 The store bound occurs when a store operation stall the CPU.
-Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc} \TG{This figure should go to appendix, it's not critical to the understanding of the paper}.
+Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}.
 We use the convention from~\cite{Kukunas2015-jd} for the equations.
@@
 
+\HL{We aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile.
+The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes.
+We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature}~\cite{Kukunas2015-jd}.
+\HL{Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the pipeline.
+This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
+We report the mean and standard deviation of the execution time for those functions.}
+
 \subsection{CPU time distribution across functions was long tailed}
@@
 While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a Pareto distribution.
-We plotted the distribution of the functions' CPU time on a log-log scale and found it to be closer \TG{``closer'' is too vague: is it well approximated by a Zipf's law?
-Can you provide a p-value of this claim?} to a Zipf's law distribution with exponent $\alpha=1.8$.
-\TG{Add a few sentences to explain why this is interesting.}
+Using the \textit{powerlaw} Python package~\cite{Alstott2014-gr} we found that the data fit a Zipf's law with exponent $\alpha=1.46$ (see Figure~\ref{fig:zips-law}).
+Therefore the CPU time of a function is approximately inversely proportional to its rank.
+This is consistent with the Pareto's principle, but is more extreme.
+\HL{In other words, most of the CPU time is spent in very few functions.
+By consequence, future efforts to optimize this minimal set of functions could result in significant performance improvement.}
 
@@
 
-\subsection{Makespan: A performance overview}
-Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e.
-makespan.
-There is a significant difference between the makespan of different applications.
-This is expected as the pipelines perform different tasks.
-The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation.
-Overall, this shows that the input data does not substantially affect the makespan of those pipelines.
-\TG{not sure what to make from this result, I don't find it particularly noticeable, I think I would omit it.}
+\subsection{Main Bottleneck: Linear Interpolation}
+Table~\ref{extab:interpolation} shows that interpolation is a critical part for all pipelines except FreeSurfer recon-all, contributing between 32.20\% to 62.70\% of the total CPU time.
+Overall, the number of functions using interpolation is low with less than 20 in each pipelines.
+In other words, few interpolation functions contribute to a large portion of the CPU time.
+While each interpolation have low computational cost, the large amount of operations make it a significant bottleneck.
+Therefore, small optimization to the interpolation operation would bring substantial benefits to these pipelines.
 
-\begin{figure}[t]
+\csvnames{csvcol}{1=\pipeline, 2=\nfunc, 3=\cputime}
+\csvstyle{interpolation}{
+	tabular = |r|c|c|,
+	table head = \hline Pipeline & \# of functions & \% CPU Time \\
+	& with interpolation & from interpolation\\\hline\hline,
+	late after line = \\\hline,
+	respect all,
+	csvcol
+}
+\begin{table*}[ht]
 	\centering
-	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
-	\caption{Average makespan of the pipeline while using 1-thread. The error bars show the standard deviation. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
-	\label{fig:makespan-1thread}
-\end{figure}
+	\csvreader[interpolation]{tables/interpolation.csv}{}
+	{\pipeline & \nfunc & \tablenum[round-precision=2]{\cputime}}
+	\caption{Contribution of interpolation to the pipelines' total CPU time.
+The percentage is the average sum of CPU time of functions using interpolation.
+The data includes all functions; not only the top 80\% of the CPU time.}
+	\label{extab:interpolation}
+\end{table*}
+
+Despite an extensive literature on linear interpolation \MD{TODO cite it}, no particular optimization techniques are available in the MRI pre-processing pipelines we profiled.
+We think it would be interesting to explore some simple optimization techniques.
+For example, the authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation.
+We think it would be interesting for future work to explore these different alternatives to accelerate interpolation in the pipelines, while considering the trade-off between accuracy and computational cost.
+Reduced- or mixed-precision techniques could also be explored to accelerate interpolation, with similar trade-off.
+Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements.
+Improving the performance of interpolation would have a large impact for 3D MRI imaging processing, and other fields such as climate simulation, astronomical imaging, or 3D image reconstruction.
 
 \subsection{Impact of Memory Bounded Functions}
-Table~\ref{extab:memory-bound} shows that fetching data from disk is the main source of cycles where the CPU is starved \TG{I don't see disk in table 1}.
+Table~\ref{extab:memory-bound} shows that fetching data from disk \HL{(Memory bound - sum of other metrics)} is the main source of cycles where the CPU is starved.
 Following, the L1 bound has the highest cache-level contribution to the memory bound of the pipelines.
@@
 We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.
-\TG{Add discussion: what is interesting about this result?}
+
+Optimizing the data access of the pipeline could reduce the number of data load.
+This would potentially have a large impact on the performance of the pipelines.
+Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound.
+To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing.
+In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
+\MD{TODO TG: You could cite some of Valerie's works in this direction.}
 
@@
 
-\subsection{Main Bottleneck: Linear Interpolation}
-Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all.
-For those pipelines, interpolation contributes between 32.20\% to 62.70\% of the total CPU time.
-Overall, the number of functions using interpolation is low with less than 20 in each applications.
-In other words, few interpolation functions are contribute to a large portion of the CPU time.
-\TG{This should be moved up in the results, I think right after Figure 2, as this is one of the most interesting results.} \TG{Add a few sentences to explain why this is interesting.}
-% There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)
-
-\csvnames{csvcol}{1=\pipeline, 2=\nfunc, 3=\cputime}
-\csvstyle{interpolation}{
-	tabular = |r|c|c|,
-	table head = \hline Pipeline & \# of functions & \% CPU Time \\
-	& with interpolation & from interpolation\\\hline\hline,
-	late after line = \\\hline,
-	respect all,
-	csvcol
-}
-\begin{table*}[ht]
-	\centering
-	\csvreader[interpolation]{tables/interpolation.csv}{}
-	{\pipeline & \nfunc & \tablenum[round-precision=2]{\cputime}}
-	\caption{Contribution of interpolation to the applications' total CPU time.
-The percentage is the average sum of CPU time of functions using interpolation.
-The data includes all functions; not only the top 80\% of the CPU time.}
-	\label{extab:interpolation}
-\end{table*}
-
-\subsection{ANTS: Single vs. Double Precision}
+\subsection{ANTs: Single vs. Double Precision}
 Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision.
@@
 
-The number of iteration between both version of ANTS registrationSyN is similar.
-In fact, for the latest two stages, they were the same across all but one execution.
-However, The single precision version of ANTS registrationSyN took longer per iteration than the double precision version (Figure~\ref{fig:mean-time-per-iteration-ants}).
+ANTs registrationSyN is a three stage process (Euler transform, affine transform, and SyN registration), each having 4 level with a specified number of maximum iterations.
+We focus on the SyN registration stage since it is multiple order longer to excecute.
+The number of iterations between both version of ANTs registrationSyN is similar.
+In fact, for the last two levels, the maximum specified number of iterations is reached across all but one execution.
+However, the single precision version of ANTS registrationSyN took \HL{approximately 20-25\%} longer per iteration than the double precision version for the last registration stage (Figure~\ref{fig:mean-time-per-iteration-ants}).
 Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.
-\TG{This is too inconclusive, you should give all the details here.}
+
+After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the template\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of its function to be in double precision, with it's \textit{EvaluateAtContinuousIndex} contributing most of the CPU time.
+Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast back to float in ANTS.
+We think this bug causes the slow-down.
+We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base.
+This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base.
+Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
+We reported the issue on ITK GitHub repository at: \href{}{} \MD{TODO open issue}
 
@@
 Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded execution.
-In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime.
+In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime.
 FreeSurfer uses the multi-processing API provided by OpenMP.
@@
 
-\section{Discussion}
-
-\TG{I think the discussion should be merged in the results.
-I know it's unconventional but I think it would make the paper easier to read and more interestingly.
-You could add a short conclusion that summarizes the main take-aways from the results.}
-
-\subsection{Few functions account for most of the CPU time}
-Our initial assumption was that the distribution of CPU time would follow a Pareto distribution.
-Thus, 20\% of the functions would contribute to 80\% of the total CPU time.
-However, Figure~\ref{fig:long-tail-distribution} shows \TG{Figure 2 doesn't ``show'' that.
-Rephrase.} that the distribution of CPU time is closer to a Zipf's law with exponent $\alpha=1.8$.
-This means that the CPU time of a function is approximately inversely proportional to its rank.
-This is consistent with the Pareto's principle, but is more extreme.
-Therefore, future efforts to optimize the main bottlenecks of MRI pre-processing pipelines could lead to enormous performance improvements.
-
-\subsection{Wasted CPU cycles due to Memory Bound}
-Overall, our results show a high percent of memory bound across all pipelines.
-Optimizing the data access of the application could reduce the number of data load.
-This would potentially have a large impact on the performance of the pipelines.
-Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound.
-To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing.
-In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
-\TG{You could cite some of Valerie's works in this direction.}
-
-\subsection{Importance of Linear Interpolation}
-Our results show that linear interpolation the primary bottleneck for all pipelines, except FreeSurfer recon-all.
-While each interpolation have low computational cost, the large amount of operations make it a significant bottleneck.
-
-There is an extensive literature on interpolation \TG{cite it}.
-Unfortunately, large amount of optimization techniques are not applicable to MRI pre-processing pipelines, because of the rotation during registration \TG{explain how interpolation is related to registration.
-In your presentation of fmriprep, explain what registration is and the role it plays in the pipelines.} \TG{this is a bit cryptic and I don't think you want to state that here.
-You may want to say that it would be interesting to further investigate these techniques in the future.} which requires dynamic kernels for interpolation.
-Moreover, the high-resolution of the data and iterative refinement of the registration require a large number of interpolations to be performed.
-
-The authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation.
-They review algorithm trade-off accuracy for a lower computational cost.
-We think it would be interesting for future work to explore these different alternatives to accelerate interpolation within the pipelines.
-However, special attention will be required to balance the accuracy and the performance of the algorithm.
-
-Similarly, we propose future efforts to explore reduced- or mixed-precision techniques to accelerate interpolation.
-Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements.
-
-Improvements to interpolation in 3D MRI imaging would have a large impact.
-It would also benefit other fields of studies using interpolation for high-resolution imaging with iterative refinement, such as climate simulation, astronomical imaging, or 3D image reconstruction.
-\TG{You could summarize the previous 3 paragraphs, I find them a bit verbose.}
-
-\subsection{ANTS: A tale of precision}
-From our results, we observe a slow-down for the single precision version of ANTs pipelines.
-A single iteration takes longer to execute when using single-precision compared to double precision.
-After a further analysis of the ANTs registrationSyN single-precision pipeline, we found that the function\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} contributing most of the CPU time does computations in double precision.
-Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast to back to float in ANTS.
-We think this bug is the cause of the slow-down.
-We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base.
-This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base.
-Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
-\TG{Can you report the bug and add a link to the issue here?}
-
-\subsection{OpenMP bottleneck in FreeSurfer recon-all}
-In our experiments, FreeSurfer recon-all showed a significant decrease in parallel efficiency with an increase in the number of threads.
-While Amdhals' law plays an impact on the maximum efficiency, the OpenMP thread synchronization played the largest impact in our results.
-The codebase uses 83 \textit{static} scheduling policy out of 92 total \TG{explain what these numbers represent}.
-\textit{Static} scheduling assign chunks to threads in round-robin fashion.
+OpenMP offers multiple scheduling policies to assign chunks to threads including -- \textit{static} scheduling assigns chunks to threads in round-robin fashion and \textit{dynamic} which assign chunks to threads as they become available for work.
+In the codebase 83 out of 92 tasks use the \textit{static} scheduling policy.
 This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency.
@@
 This remains challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
-\TG{Could you file a bug report in freesurfer on this issue and refer to it here?}
+We opened an issue on FreeSurfer GitHub repository at: \href{}{} \MD{TODO open issue}
 
@@
 \subsection{VTune results finalization}
-Using VTune to collect profiling results does not add a large overhead in term of time and resources utilization.
-However, a finalization step is required to query the data and generate a VTune report.
+\HL{VTune profiler offered a rich and extensive amount of data for our analysis, which would have been difficult to retrieve otherwise.
+Thanks to it's low performance overhead, we were able to profile the pipeline with minimal impact on the runtime, allowing the use of more subjects to obtain a more accurate performance profile.
+The documentation provided by VTune is extensive and well-documented, which helped us to understand the different metrics and events available.}
+
+Unfortunately, VTune requires a finalization step is required to query the data and generate a report.
 We found this step to be both time consuming and resource intensive.
-For example, finalizing the results from a single FreeSurfer recon-all execution requires up to \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline.
+For example, finalizing the results from a single FreeSurfer recon-all execution can requires over \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline.
 This significant difference in resource requirement creates a challenge to profile pipeline with long runtime, due to the potential difficulty to obtain access to compute nodes with enough RAM.
@@
 This would lead to a potential loss of information in the profiling results.
+\HL{Because of these limitations, we were unable to finalize the results of fMRIPrep for analysis.}
 
-\subsection{Limitations}
-\TG{You should frame the limitations as future work in the conclusion. These are not really limitations.}
+\section{Conclusion}
+In this paper, we profiled several MRI pre-processing pipelines: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep.
+We \HL{observe} that few functions contribute to a large amount of CPU time \HL{which presents the opportunity to optimize a practical amount of functions to potentially obtain significant speed-up}.
+We found that most pipelines suffered from memory bound.
+Linear interpolation was the primary bottleneck for these pipelines.
+We discovered a bug in ITK which lead the implementation of ANTS registration in single-precision to use float for its computation.
+We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP.
+Last, we discuss the challenge of profiling long running pipelines with VTune, due to computational resource requirements to obtain results.
+
 The dataset we chose contains wide range of age and equal distribution of sex.
@@
 Therefore, the performance bottleneck we found might not be applicable to data pre-processing of pathological patients.
-Moreover, the dataset is from a single site.
-This could also introduce bias from the scanner model and data acquisition protocol used.
-Future work should include data from multiple sites to ensure the generalization of the performance bottleneck we found.
-\TG{Is there evidence showing that this would provide additional insights on the benchmark?
-I think it might be interesting to study subjects that fail QC but I'm not sure about other populations.}
-
-Our analysis omits the space or time of execution for function calls \TG{I don't understand what this means}.
+\HL{It would be interesting to study the impact of QC and/or pathologies on performance.} Moreover, our profiling with VTune does not account for the spatiality (e.g.
+for loop) or temporality (e.g.
+start vs.
+end of convergence) of the functions.
 This information could provide additional insight for optimization.
-For example, spatial and temporal dimensions were shown to be important in the field of reduced precision.
-Future work could extent the depth of the analysis by including this information.
-\MD{TODO Cite paper on spatial/temporal for RP (ask yohan for references)}
-
-\section{Conclusion}
-In this paper, we profiled several MRI pre-processing applications: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep.
-We confirmed our hypothesis \TG{don't frame it as a hypothesis, it is an observation} that few functions contribute to a large amount of CPU time \TG{, which [explain why it is interesting to know]}.
-We found that most applications suffered from memory bound.
-Linear interpolation was the primary bottleneck for these applications.
-We discovered a bug in ITK which lead the implementation of ANTS registration in single-precision to use float for its computation.
-We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP.
-Last, we discuss the challenge of profiling long running applications with VTune, due to computational resource requirements to obtain results.
+Future work could extend the depth of the analysis by including this information.
 
@@
 \label{sec:code-availability}
-The code to compile, profile the pipelines, and generate figures is publicly available at:
+The code download the data, compile and profile the pipelines, and generate the figures is publicly available at:
 \\\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
@@
 \section*{Acknowledgement}
-Mathieu Dugr'e was jointly funded by the Natural Sciences and Engineering Research Council of Canada (NSERC) through the Postgraduate Scholarship-Doctoral (PGS-D) program and the Graduate Doctoral Fellowship Award from Concordia University.
+Mathieu Dugr\'e was jointly funded by the Natural Sciences and Engineering Research Council of Canada (NSERC) through the Postgraduate Scholarship-Doctoral (PGS-D) program and the Graduate Doctoral Fellowship Award from Concordia University.
 \MD{Q: Any other?}
@@
 \section*{Supplementary Materials}
+\begin{figure*}[ht]
+	\noindent
+	\begin{minipage}
+		{0.5\textwidth}
+		\begin{align*}
+			C & = CPU\_CLK\_UNHALTED.THREAD                \\
+			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
+			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
+			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
+			W & = MEM\_L3\_WEIGHT                          \\
+			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
+			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
+			M & = \frac{W \times R}{H + R}                 \\
+			N & = \frac{H}{H + W \times R}                 
+		\end{align*}
+	\end{minipage}
+	\begin{minipage}
+		{0.5\textwidth}
+		\begin{align}
+			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
+			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
+			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
+			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
+			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
+		\end{align}
+	\end{minipage}
+	\caption{Intel performance monitoring events and derived memory metrics}
+	\label{fig:memory-metrics}
+\end{figure*}
+
+\MD{From Tristan: If possible, include the colorful figure in the paper.}
+
 \MD{Not sure if these add value?}
@@
 	\centering
-	\includegraphics{figures/global-zipf_law.png}
-	\caption{Distribution of the functions' CPU time compare to a Zipf's law distribution with $\alpha=1.8$.
+	\includegraphics{figures/zipf_law.png}
+	\caption{Distribution of the functions' CPU time compare to a Zipf's law distribution with $\alpha=1.46$.
 The y-axis shows the average CPU time for a function.
@@
 The data includes all functions from all pipelines.}
-	\label{sup-fig:zips-law}
+	\label{fig:zips-law}
 \end{figure*}
@@
 % CSV tables
-% \csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
-\csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}
+% \csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the pipeline makespan.}{extab:hotspots-1thread-freesurfer-reconall}
+\csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the pipeline makespan.}{extab:hotspots-32threads-freesurfer-reconall}
 

### 2024-04-17 sbatch/pre_run.sh
[FIX] unbound variable `VTUNE_OPTIONS`.

@@ -2,6 +2,7 @@
 set -e
 set -u
 
+export VTUNE_OPTION=""
 TOOLKIT=$1; shift
 PIPELINE=$1; shift
 USAGE="

### 2024-04-17 paper/paper.tex
[ADD] link to ITK issue.

@@
 Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
-We reported the issue on ITK GitHub repository at: \href{}{} \MD{TODO open issue}
+We reported the issue on ITK GitHub repository (see issue \#4593)~\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}}.
 
@@
 Our profiling results from VTune are publicly available at:
-\\\href{URL}{\MD{TODO add Zenodo link}}
-% Waiting on data transfer from Compute Canada
+\\\href{https://doi.org/10.5281/zenodo.10987491}{https://doi.org/10.5281/zenodo.10987491}
+\MD{TODO finalize the metadata entry and publish}
 

### 2024-04-17 paper/paper.tex

@@
 Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
-We reported the issue on ITK GitHub repository (see issue \#4593)~\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}}.
+We reported the issue on ITK GitHub repository (see issue \#4593~\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}}).
 

### 2024-05-01 paper/paper.tex
[UPDATE] section titles
[ADD] FreeSurfer GitHub issue.

@@
 \subsection{Pipelines}
-
 fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, which is built on top of several well-known neuroimaging toolboxes such as ANTs, FSL, FreeSurfer and AFNI.
@@
 This remains challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
-We opened an issue on FreeSurfer GitHub repository at: \href{}{} \MD{TODO open issue}
+We opened an issue on FreeSurfer GitHub repository (see issue\#1199~\footnote{\href{https://github.com/freesurfer/freesurfer/issues/1199}{https://github.com/freesurfer/freesurfer/issues/1199}})
 

### 2024-05-09 paper/paper.tex
updates in abstract

--- 2024-04-17 paper/paper.tex
@@
 While High-Performance Computing (HPC) and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers.
-Therefore, it is critical to understand the performance bottlenecks of MRI pre-processing pipelines to improve their performance.
-Using Intel VTune profiler, we characterize the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes.
-We found that few functions contribute to most of the CPU time, and that linear interpolation is the largest contributor.
-Data access is also  a substantial bottleneck.
-Last, we identified a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all.
+Therefore, it is important to understand the performance bottlenecks of MRI pre-processing pipelines to improve their performance.
+Using Intel VTune profiler, we characterized the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes.
+We found that few functions contributed to most of the CPU time, and that linear interpolation was the largest contributor.
+Data access was also  a substantial bottleneck.
+We identified a bug in the ITK library that impacts the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all.
 Our results provide a reference for future efforts to optimize MRI pre-processing pipelines.
@@
 
-\section{Results}
+\section{Results \& Discussion}
 In this section we present the aggregated profiling data obtained with VTune.
@@
 
-\subsection{VTune results finalization}
+\subsection{Experiencing using VTune}
 \HL{VTune profiler offered a rich and extensive amount of data for our analysis, which would have been difficult to retrieve otherwise.

### 2024-05-22 paper/paper.tex
Various edits

--- 2024-05-09 paper/paper.tex
@@
 Pre-processing of Magnetic Resonance Imaging (MRI) data is a critical step in neuroimaging.
-Although established pre-processing pipelines exist, they commonly require expensive computation and produce large output and intermediate data.
-\HL{These are a well-known challenges to study large subject cohorts in neuroimaging.} Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed.
-Therefore, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.
+Although established pre-processing pipelines exist, they commonly require extensive amounts of computation and produce large volumes of output and intermediate data.
+Such requirements create challenges when studying large neuroimaging cohorts, and they limit clinical applicability when timely data analysis is needed.
+Therefore, the neuroimaging community is constantly seeking for ways to improve the computational performance of MRI pre-processing pipelines.
+In this paper, we characterize the computational cost of several commonly adopted MRI pre-processing pipelines, providing a reference benchmark for future efforts to optimize MRI pre-processing pipelines.
 
-% HPCs
-Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts \HL{leveraging} a large number of CPU cores and high-bandwidth memory to process subjects in parallel.
+% % HPCs
+% Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts \HL{leveraging} a large number of CPU cores and high-bandwidth memory to process subjects in parallel.
 However, \HL{it can be challenging to run HPCs in developing countries due to power restrictions, national network infrastructure, and lack of training and funding}~\cite{NRG-SC23}.
@@
 
-% DL
-In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing.
+% % DL
+% In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing.
 For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} to provide Deep Learning alternatives to traditional algorithms\HL{, providing faster runtime.}
-% Moreover, a study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
-However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive.
+% % Moreover, a study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
+% However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive.
 Additionally, \HL{their results} can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.
 
-% Alternative solutions
-\HL{These techniques require costly hardware to be effective, which can be unacessible to underrepresented researchers.} Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements but require substantial effort to implement.
+% % Alternative solutions
+% \HL{These techniques require costly hardware to be effective, which can be unacessible to underrepresented researchers.} Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements but require substantial effort to implement.
 Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.
@@
 % - No comprehensive study of the performance bottleneck of MRI pre-processing.
-\HL{To the best of our knowledge, this is the first study that} characterize the computational cost of several commonly adopted MRI pre-processing pipelines.
-The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing pipelines.
 
@@
 Three main MRI modalities exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI).
-While other modalities such as electroencephalography (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRIs for their broad adoption and non-invasiveness.
+While other modalities such as electroencephalography (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRI for its broad adoption in research and non-invasiveness.
 Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.
@@
 fMRIPrep combines several pipelines to produce a complete pre-processing pipeline.
-We also study its sub-pipeline separately to provide a finer grain analysis of the performance bottleneck.
+We study fMRIPrep's sub-pipelines separately, to provide a finer grained analysis of the performance bottlenecks.
 
 % - VTune profiler
-To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}.
-The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level.
-We profile the pipeline using \HL{-- single-threaded and 32 threads to simplify the analysis and study the the multi-threading performance, repsectively}.
+To profile the pipelines, we use the Intel VTune tool~\cite{vtune-profiler}, a multi-language profiler with low performance overhead that provides runtime information at the level of functions.
+We profile the pipeline in single-threaded mode and using 32 threads to account for parallel executions.
 We aggregate subjects' results for each pipeline to obtain an average performance profile.
-We compiled each pipeline with debug information to obtain human readable information on function and module names in VTune.
 
@@
 \subsection{Pipelines}
-
-fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, which is built on top of several well-known neuroimaging toolboxes such as ANTs, FSL, FreeSurfer and AFNI.
-Although the processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity corretion and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain.
+fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, built on top of several well-known neuroimaging toolboxes such as ANTs, FSL, FreeSurfer and AFNI \TG{expand acronyms and add references for each toolbox}.
+Although processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity correction and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain.
 When enabled, FreeSurfer recon-all~\cite{Dale1999-wu} is used for surface reconstruction.
-We mimic the fMRIPrep specific 3 \HL{phases reconall pre-processing, which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3}.
+We mimic the fMRIPrep specific 3 phases reconall pre-processing, \TG{I don't understand this sentence} which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3.
 The fMRI pipeline generally performs motion correction with FSL MCFLIRT~\cite{Jenkinson2002-od}, slice timing correction with AFNI 3dtshift~\cite{Cox1996-nk}, and co-registration with FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}.
 
-We profile the entire fMRIPrep pipeline to obtain a coarse view of the pipeline's performance bottleneck.
-Additionlly, we profile the sub-pipelines of fMRIPrep to characterize the performance at a finer grain.
+We profiled the entire fMRIPrep pipeline to obtain a coarse view of the pipeline's performance bottleneck.
+Additionally, to characterize the performance at a finer grain, we profiled the sub-pipelines of fMRIPrep, namely \TG{add list}.
 While we used the same sub-pipeline and order as in fMRIPrep, we used the default parameters for most of the sub-pipelines which may vary from the default parameters of fMRIPrep.
 We omitted some sub-pipelines of fMRIPrep as they were not compatible with our dataset.
-For example, time-slicing correction was already performed.
+For example, slice-timing correction was already performed in our data.
 
-\HL{ANTs brainExtraction and registrationSyN provides the user with the option to use a single or double precision version of the pipeline, leveraging the Insight Segmentation and Registration Toolkit (ITK)}~\cite{Yoo2002-ve}.
-\HL{We profile both version to understand the performance benefits of using reduced precision in the pipelines.}
+ANTs brainExtraction and registrationSyN are available in single or double precision, leveraging the Insight Segmentation and Registration Toolkit (ITK)~\cite{Yoo2002-ve}.
+We profiled both versions to understand the performance benefits of using reduced precision in the pipelines.
 
@@
 To ensure that our performance profiles are inclusive of different populations, we used data with a wide range of age and equal distribution of sex.
-We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
+We used the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
 The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts.
-The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed.
+The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two sessions: eyes open and eyes closed.
 The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session.
+\TG{add information on image resolution}
 
@@
 
-First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
-For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{lst:vtune_example} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.
+First, to obtain human-readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
+For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}}.
+\lstlistingname~\ref{lst:vtune_example} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.
 
@@
 		language={sh},
-		caption={Profiling a pipeline within a Apptainer container, using VTune profiler.},
+		caption={Profiling a pipeline within an Apptainer container, using VTune profiler.},
 		captionpos=b,
@@
 
-Last, we profile the pipeline using two threading configuration: single-threaded and 32 threads.
-The single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling.
-The 32-thread approach uses all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the pipeline.
-This is a more realistic usage for the user of the pipeline.
+We profiled the pipeline using two threading configuration: single-threaded and 32 threads.
+The single-threaded approach simplified the analysis and limited the potential I/O overhead on the profiling.
+The 32-thread approach used all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the pipeline.
+This configuration provided a more realistic usage of the pipeline.
 
-Before profiling each pipeline, we transfer the input data from the shared file system to the compute node.
-After profiling, we transfer back the output to the shared file system.
-This limits profiling variability due to I/O contingency.
+Before profiling each pipeline, we transferred the input data from the shared file system to the compute node.
+After profiling, we transferred back the output to the shared file system.
+This limited profiling variability due to I/O contingency.
 
-The data generated by the VTune profiler is written to a shared file system, for administrative reasons.
-In principle, this could increase the profiling overhead.
-However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as data produced by the pipelines.
+The data generated by the VTune profiler was written to a shared file system, to simplify post-processing.
+In principle, this could have increased the profiling overhead.
+However, in practice there was no impact since the data generated by the VTune profiler was very small and was not written on the same file system as data produced by the pipelines.
 
@@
 The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load.
-L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache.
-Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively.
-The store bound occurs when a store operation stall the CPU.
-Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}.
+L1 bound occurs when a demand load stalls CPU cores although the data already reside in L1 cache.
+Similarly, L2, L3, and DRAM bound occurs when a demand load stalls the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively.
+The store bound occurs when a store operation stalls the CPU.
+Supplementary Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}.
 We use the convention from~\cite{Kukunas2015-jd} for the equations.
@@
 
-\HL{We aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile.
-The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes.
-We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature}~\cite{Kukunas2015-jd}.
-\HL{Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the pipeline.
-This heuristic allows us to focus our efforts on the performance critical sections of a pipeline.
-We report the mean and standard deviation of the execution time for those functions.}
+We aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile.
+Consistently with~\cite{Kukunas2015-jd}, we hypothesized that the Pareto principle applied to the performance of MRI pre-processing pipelines, we therefore only studied the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the pipeline.
+This heuristic allowed us to focus our efforts on the performance critical sections of a pipeline.
+We report the mean and standard deviation of the execution time for those functions.
 
-\subsection{CPU time distribution across functions was long tailed}
+\subsection{CPU time distribution across functions was long-tailed}
 Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time was spent in less than 1.5\% of the function across all pipelines.
-While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a Pareto distribution.
-Using the \textit{powerlaw} Python package~\cite{Alstott2014-gr} we found that the data fit a Zipf's law with exponent $\alpha=1.46$ (see Figure~\ref{fig:zips-law}).
-Therefore the CPU time of a function is approximately inversely proportional to its rank.
-This is consistent with the Pareto's principle, but is more extreme.
-\HL{In other words, most of the CPU time is spent in very few functions.
-By consequence, future efforts to optimize this minimal set of functions could result in significant performance improvement.}
+While consistent with the Pareto principle, the distribution of the functions' CPU time did not follow a Pareto distribution.
+Using the \textit{powerlaw} Python package~\cite{Alstott2014-gr}, we found that the data fit a Zipf law with exponent $\alpha=1.46$ (Supplementary Figure~\ref{fig:zips-law}).
+Therefore, the CPU time of a function was approximately inversely proportional to its rank.
+Thus, future efforts to optimize this minimal set of functions could result in significant performance improvements.
 
@@
 The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage.
-The x-axis percentage of functions ordered by decreasing CPU time.
+The x-axis is the percentage of functions ordered by decreasing CPU time.
 The data includes all functions from all pipelines.
@@
 \subsection{Main Bottleneck: Linear Interpolation}
-Table~\ref{extab:interpolation} shows that interpolation is a critical part for all pipelines except FreeSurfer recon-all, contributing between 32.20\% to 62.70\% of the total CPU time.
-Overall, the number of functions using interpolation is low with less than 20 in each pipelines.
-In other words, few interpolation functions contribute to a large portion of the CPU time.
-While each interpolation have low computational cost, the large amount of operations make it a significant bottleneck.
-Therefore, small optimization to the interpolation operation would bring substantial benefits to these pipelines.
+Table~\ref{extab:interpolation} shows that interpolation was a critical part for all pipelines except FreeSurfer recon-all, contributing between 32.20\% to 62.70\% of the total CPU time.
+Overall, the number of functions using interpolation was low, with less than 20 in each pipeline: few interpolation functions contributed to a large portion of the CPU time.
+While each interpolation had low computational cost, the large amount of operations resulted in a significant bottleneck.
+Therefore, optimizing interpolation would bring substantial benefits to these pipelines.
 
@@
 	table head = \hline Pipeline & \# of functions & \% CPU Time \\
-	& with interpolation & from interpolation\\\hline\hline,
+	& with interpolation & spent in interpolation\\\hline\hline,
 	late after line = \\\hline,
@@
 The percentage is the average sum of CPU time of functions using interpolation.
-The data includes all functions; not only the top 80\% of the CPU time.}
+The data includes all functions from all pipelines.}
 	\label{extab:interpolation}
@@
 
-Despite an extensive literature on linear interpolation \MD{TODO cite it}, no particular optimization techniques are available in the MRI pre-processing pipelines we profiled.
-We think it would be interesting to explore some simple optimization techniques.
-For example, the authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation.
-We think it would be interesting for future work to explore these different alternatives to accelerate interpolation in the pipelines, while considering the trade-off between accuracy and computational cost.
-Reduced- or mixed-precision techniques could also be explored to accelerate interpolation, with similar trade-off.
-Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements.
-Improving the performance of interpolation would have a large impact for 3D MRI imaging processing, and other fields such as climate simulation, astronomical imaging, or 3D image reconstruction.
-
 \subsection{Impact of Memory Bounded Functions}
-Table~\ref{extab:memory-bound} shows that fetching data from disk \HL{(Memory bound - sum of other metrics)} is the main source of cycles where the CPU is starved.
-Following, the L1 bound has the highest cache-level contribution to the memory bound of the pipelines.
-ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT.
-L2 and L3 bound is low for all pipeline, although highest for FreeSurfer recon-all.
-The DRAM bound is high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision.
-Store bound is low for all pipelines.
+Table~\ref{extab:memory-bound} shows that fetching data from disk (Memory bound - sum of other metrics) \TG{I don't understand this subtraction} was the main cause for starving the CPU.
+Next, the L1 bound had the highest cache-level contribution to the memory bound of the pipelines.
+ANTS pipelines had the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT.
+L2 and L3 bound was low for all pipelines, although higher for FreeSurfer recon-all.
+The DRAM bound was high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision.
+Store bound was low for all pipelines.
 We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.
@@
 In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
-\MD{TODO TG: You could cite some of Valerie's works in this direction.}
 
@@
 The values are the summation of the metric weighted by each function CPU time.
-This represent the percentage of the total CPU time stalled by each metrics.
-We show the average value across all subjects execution with one thread.}
+This represents the percentage of the total CPU time stalled by each metrics \TG{not sure what this means}.
+\TG{Memory bound is not the sum of all metrics} We show the average value across all subjects execution with one thread.}
 	\label{extab:memory-bound}
@@
 Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision.
-For both pipelines, the makespan is significantly higher in single precision.
-This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv}.
+For both pipelines, the makespan was significantly higher in single precision than in double precision.
+This was unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv} than in double precision.
 
@@
 
-ANTs registrationSyN is a three stage process (Euler transform, affine transform, and SyN registration), each having 4 level with a specified number of maximum iterations.
-We focus on the SyN registration stage since it is multiple order longer to excecute.
-The number of iterations between both version of ANTs registrationSyN is similar.
-In fact, for the last two levels, the maximum specified number of iterations is reached across all but one execution.
-However, the single precision version of ANTS registrationSyN took \HL{approximately 20-25\%} longer per iteration than the double precision version for the last registration stage (Figure~\ref{fig:mean-time-per-iteration-ants}).
-Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.
+ANTs registrationSyN is a three-stage process (Euler transform, affine transform, and SyN registration), where each stage has 4 levels \TG{levels of what?} with a specified number of maximum iterations.
+We focused on the SyN registration stage since it was substantially longer to execute than the other stages.
+The number of iterations between both versions of ANTs registrationSyN was similar.
+In fact, for the last two levels, the maximum specified number of iterations was reached across all but one execution.
+However, the single precision version of ANTS registrationSyN took approximately 20-25\% longer per iteration than the double precision version for the last registration stage (Figure~\ref{fig:mean-time-per-iteration-ants}).
+Therefore, the slowdown was not due to the convergence of the algorithm, but rather the execution time for each iteration \TG{add data on this to Figure}.
 
-After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the template\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of its function to be in double precision, with it's \textit{EvaluateAtContinuousIndex} contributing most of the CPU time.
-Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast back to float in ANTS.
+After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the use of templates in ANTs' C++ code\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of its \TG{clarify what 'its' refers to} function to be in double precision, with \textit{EvaluateAtContinuousIndex} contributing most of the CPU time.
+Therefore, the pipeline wastes CPU cycles to convert the input data to double-precision in ITK, and potentially to cast back to single-precision in ANTs.
 We think this bug causes the slow-down.
 We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base.
-This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base.
-Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
-We reported the issue on ITK GitHub repository (see issue \#4593~\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}}).
+This example shows that the use of reduced precision is not trivial and requires a deep understanding of the code base.
+Moreover, benchmarks should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
+We reported this issue in the ITK GitHub repository\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}}.
 
@@
 	\includegraphics[width=\linewidth]{figures/ants-registrationSyN-iteration-mean.png}
-	\caption{Mean time per iteration for ANTS registrationSyN in double and single precision.
+	\caption{Average time per iteration for ANTS registrationSyN in double and single precision.
 Only the SyN Registration stage is shown as the two earlier stage were near zero time.
@@
 \subsection{FreeSurfer: Thread-Synchronization}
-Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded execution.
-In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime.
-FreeSurfer uses the multi-processing API provided by OpenMP.
-Moreover, the y-axis scale is multi folds larger than for the single-threaded executions.
-This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.
+Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded executions.
+In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP \TG{expand acronym} was a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime.
+FreeSurfer used the multi-processing API provided by OpenMP.
+Moreover, the y-axis scale is multi folds larger than for the single-threaded execution, indicating that the multi-threaded execution was significantly slower than the single-threaded execution.
 
@@
 	\end{subfigure}
-	\caption{FreeSurfer recon-all hotspots analysis.
-The y-axis show the average CPU time spent in each function, with error bars showing the standard deviation.
-The x-axis show the function ordered by decreasing CPU time grouped by module.
-We omit the function names for clarity.
+	\caption{FreeSurfer recon-all analysis.
+The y-axis shows the average CPU time spent in each function, with error bars showing the standard deviation \TG{across subjects?}.
+The x-axis shows the function ordered by decreasing CPU time grouped by module.
+We omitted function names for clarity.
 The function ID are dependent to each plot.
-The supplementary materials depicts the mapping of the function ID to the function name for each plot.}
+Supplementary materials show the mapping of the function ID to the function name for each plot.}
 	\label{fig:hotspots-freesurfer-reconall}
@@
 
-While the makespan for FreeSurfer recon-all decreases with an increase in the number of threads, the benefits are limited (Figure~\ref{fig:freesurfer-threading}).
-The parallel efficiency decreases from 66\% with two threads down to 5\% with 32 threads.
-While Amdhals' law plays an impact on the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} indicates that the thread Synchronization has a larger impact.
+While the makespan for FreeSurfer recon-all decreased with an increase in the number of threads, the benefits were limited (Figure~\ref{fig:freesurfer-threading}).
+The parallel efficiency decreased from 66\% with two threads down to 5\% with 32 threads.
+While Amdhal's law limits the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} indicates that thread Synchronization had a larger impact \TG{how does the figure indicate that?}.
 
@@
 	\includegraphics[width=\linewidth]{figures/makespan-freesurfer.png}
-	\caption{Makespan and parallel efficiency of FreeSurfer recon-all while varying the number of threads between 1 to 32.
+	\caption{Makespan and parallel efficiency of FreeSurfer recon-all while varying the number of threads from 1 to 32.
 The left y-axis shows the makespan in seconds, while the right y-axis shows the parallel efficiency in percent.
@@
 
-OpenMP offers multiple scheduling policies to assign chunks to threads including -- \textit{static} scheduling assigns chunks to threads in round-robin fashion and \textit{dynamic} which assign chunks to threads as they become available for work.
-In the codebase 83 out of 92 tasks use the \textit{static} scheduling policy.
+OpenMP offers multiple scheduling policies to assign chunks to threads, including \textit{static} scheduling that assigns chunks to threads in round-robin fashion, and \textit{dynamic} which assign chunks to threads as they become available for work.
+In the codebase, 83 out of 92 tasks use the \textit{static} scheduling policy.
 This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency.
 We speculated that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads.
-Unfortunately, we failed to recompile FreeSurfer by naively changing the \textit{static} OpenMP scheduling to \textit{dynamic}.
+Unfortunately, we failed to recompile FreeSurfer having changed the \textit{static} OpenMP scheduling to \textit{dynamic}.
 This remains challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
-We opened an issue on FreeSurfer GitHub repository at: \href{}{} \MD{TODO open issue}
+We opened an issue on FreeSurfer GitHub repository to report this problem\footnote{\href{https://github.com/freesurfer/freesurfer/issues/1199}{https://github.com/freesurfer/freesurfer/issues/1199}}.
 
@@
 \subsection{Experiencing using VTune}
-\HL{VTune profiler offered a rich and extensive amount of data for our analysis, which would have been difficult to retrieve otherwise.
-Thanks to it's low performance overhead, we were able to profile the pipeline with minimal impact on the runtime, allowing the use of more subjects to obtain a more accurate performance profile.
-The documentation provided by VTune is extensive and well-documented, which helped us to understand the different metrics and events available.}
+The VTune profiler provided a rich and extensive amount of data for our analysis, which would have been difficult to retrieve otherwise.
+Thanks to its low performance overhead, we were able to profile the pipeline with minimal impact on the runtime, allowing the use of more subjects to obtain a more accurate performance profile.
+The documentation provided by VTune is extensive and well-documented, which helped us to understand the different metrics and events available.
 
-Unfortunately, VTune requires a finalization step is required to query the data and generate a report.
-We found this step to be both time consuming and resource intensive.
-For example, finalizing the results from a single FreeSurfer recon-all execution can requires over \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline.
+VTune requires a finalization step to query the data and generate a report.
+We found this step to be both time-consuming and resource intensive.
+For example, finalizing the results from a single FreeSurfer recon-all execution can require over \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline.
 This significant difference in resource requirement creates a challenge to profile pipeline with long runtime, due to the potential difficulty to obtain access to compute nodes with enough RAM.
-This can be further challenging for researchers in early career or in developing countries, which might not have access to large HPC infrastructure.
 One way to mitigate this challenge is to lower the sampling rate during the profiling.
-However, to reach a similar level of memory consumption, the sampling rate must be significantly reduced.
+However, to reach a similar level of memory consumption \TG{than what?}, the sampling rate must be significantly reduced.
 This would lead to a potential loss of information in the profiling results.
-\HL{Because of these limitations, we were unable to finalize the results of fMRIPrep for analysis.}
+Because of these limitations, we were unable to finalize the results of fMRIPrep \TG{do you mean the entire fmriprep pipeline?} for analysis.
 
 \section{Conclusion}
-In this paper, we profiled several MRI pre-processing pipelines: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep.
-We \HL{observe} that few functions contribute to a large amount of CPU time \HL{which presents the opportunity to optimize a practical amount of functions to potentially obtain significant speed-up}.
+In this paper, we profiled several MRI pre-processing pipelines: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep \TG{was it finalized?}.
+We observed that few functions contribute to a large amount of CPU time which offers opportunities to optimize a practical amount of functions to potentially obtain significant speed-up values.
 We found that most pipelines suffered from memory bound.
 Linear interpolation was the primary bottleneck for these pipelines.
-We discovered a bug in ITK which lead the implementation of ANTS registration in single-precision to use float for its computation.
+We discovered a bug in ITK which leads the implementation of ANTS registration in single-precision to use double precision in practice.
 We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP.
-Last, we discuss the challenge of profiling long running pipelines with VTune, due to computational resource requirements to obtain results.
+Lastly, we discussed the challenge of profiling long-running pipelines with VTune, due to computational resource requirements.
 
@@
 However, it only contains healthy subjects.
-Therefore, the performance bottleneck we found might not be applicable to data pre-processing of pathological patients.
-\HL{It would be interesting to study the impact of QC and/or pathologies on performance.} Moreover, our profiling with VTune does not account for the spatiality (e.g.
+It would be interesting to study the impact of image quality and pathologies on performance.
+Moreover, our profiling with VTune did not account for the spatiality (e.g.
 for loop) or temporality (e.g.
@@
 
-We suggest that future effort focus on reduced precision techniques, which could reduce memory bound and accelerate interpolation computation.
-We note that the use of reduced precision is not trivial, as seen with the ANTS bug.
-Therefore, careful attention would be required to find a balance between performance and accuracy, while ensuring bugs are not introduced.
-We also suggest that future work should study the different configuration of OpenMP scheduling in FreeSurfer recon-all to improve the parallel efficiency.
-This would lead to faster runtime time and higher CPU usage ration in HPC allocations.
+Future optimization efforts could focus on reduced precision techniques, to reduce memory bound and accelerate interpolation computation.
+We note that the use of reduced precision is not trivial though, as seen with the ANTs bug.
+Therefore, careful attention would be required to find a balance between performance and accuracy.
+We also suggest that future work study the different configurations of OpenMP scheduling in FreeSurfer recon-all, to improve parallel efficiency.
+This would lead to faster executions time and higher CPU usage in HPC allocations.
 
@@
 
+% \TG{Move this paragraph to future work}
+% Despite an extensive literature on linear interpolation \MD{TODO cite it}, no particular optimization techniques are available in the MRI pre-processing pipelines we profiled.
+We think it would be interesting to explore simple optimization techniques.
+For example, the authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation.
+We think it would be interesting for future work to explore these different alternatives to accelerate interpolation in the pipelines, while considering the trade-off between accuracy and computational cost.
+Reduced- or mixed-precision techniques could also be explored to accelerate interpolation, with similar trade-off.
+Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements.
+Improving the performance of interpolation would have a large impact for 3D MRI imaging processing, and other fields such as climate simulation, astronomical imaging, or 3D image reconstruction.
 \section{Data Availability}
@@
 
-The container images for our profiling experiments are publicly on Docker Hub:
+The container images for our profiling experiments are available on Docker Hub:
 \begin{itemize}
@@
 
-Our profiling results from VTune are publicly available at:
+Our profiling results from VTune are available at:
 \\\href{https://doi.org/10.5281/zenodo.10987491}{https://doi.org/10.5281/zenodo.10987491}
@@
 \label{sec:code-availability}
-The code download the data, compile and profile the pipelines, and generate the figures is publicly available at:
+The code to download the data, compile and profile the pipelines, and generate the figures is available at:
 \\\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
@@
 Mathieu Dugr\'e was jointly funded by the Natural Sciences and Engineering Research Council of Canada (NSERC) through the Postgraduate Scholarship-Doctoral (PGS-D) program and the Graduate Doctoral Fellowship Award from Concordia University.
-\MD{Q: Any other?}
+This work was partially funded by the Canada Research Chairs program.
+The computing infrastructure was provided by the Canada Foundation for Innovation.
 

### 2024-05-22 paper/paper.tex

@@
 When enabled, FreeSurfer recon-all~\cite{Dale1999-wu} is used for surface reconstruction.
-We mimic the fMRIPrep specific 3 phases reconall pre-processing, \TG{I don't understand this sentence} which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3.
+We benchmarked a recon-all command similar to  fMRIPrep's, which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3.
 The fMRI pipeline generally performs motion correction with FSL MCFLIRT~\cite{Jenkinson2002-od}, slice timing correction with AFNI 3dtshift~\cite{Cox1996-nk}, and co-registration with FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}.
@@
 \subsection{Impact of Memory Bounded Functions}
-Table~\ref{extab:memory-bound} shows that fetching data from disk (Memory bound - sum of other metrics) \TG{I don't understand this subtraction} was the main cause for starving the CPU.
+Table~\ref{extab:memory-bound} shows that fetching data from disk was the main cause for starving the CPU \TG{add data fetching time \% to table}.
 Next, the L1 bound had the highest cache-level contribution to the memory bound of the pipelines.
-ANTS pipelines had the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT.
+ANTs pipelines had the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT.
 L2 and L3 bound was low for all pipelines, although higher for FreeSurfer recon-all.
-The DRAM bound was high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision.
+The DRAM bound was high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTs brainExtraction in single and double precision.
 Store bound was low for all pipelines.
@@
 	\caption{Impact of data load on stalled CPU cycles.
-The values are the summation of the metric weighted by each function CPU time.
-This represents the percentage of the total CPU time stalled by each metrics \TG{not sure what this means}.
-\TG{Memory bound is not the sum of all metrics} We show the average value across all subjects execution with one thread.}
+For each metric (memory, L1, L2, L3, DRAM, and store), reported values are summations weighted by function CPU time, averaged over n=20 subjects, and obtained in single-threaded mode.
+They represent the percentage of the total CPU time stalled for the metric.}
 	\label{extab:memory-bound}
@@
 \subsection{ANTs: Single vs. Double Precision}
-Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision.
+Figure~\ref{fig:makespan-ants} shows the makespan of ANTs brainExtraction and ANTs registrationSyN in double and single precision.
 For both pipelines, the makespan was significantly higher in single precision than in double precision.
@@
 	\includegraphics[width=\linewidth]{figures/makespan-ants.png}
-	\caption{Comparison of makespan between double (blue) and single (orange) precision for ANTS brainExtraction and ANTS registrationSyN.}
+	\caption{Comparison of makespan between double (blue) and single (orange) precision for ANTs brainExtraction and ANTs registrationSyN.}
 	\label{fig:makespan-ants}
@@
 
-ANTs registrationSyN is a three-stage process (Euler transform, affine transform, and SyN registration), where each stage has 4 levels \TG{levels of what?} with a specified number of maximum iterations.
+ANTs registrationSyN is a three-stage process (Euler transform, affine transform, and SyN registration), where each stage is successively performed at 4 resolution levels with a specified number of maximum iterations.
 We focused on the SyN registration stage since it was substantially longer to execute than the other stages.
@@
 In fact, for the last two levels, the maximum specified number of iterations was reached across all but one execution.
-However, the single precision version of ANTS registrationSyN took approximately 20-25\% longer per iteration than the double precision version for the last registration stage (Figure~\ref{fig:mean-time-per-iteration-ants}).
-Therefore, the slowdown was not due to the convergence of the algorithm, but rather the execution time for each iteration \TG{add data on this to Figure}.
+However, the single precision version of ANTs registrationSyN took approximately 20-25\% longer per iteration than the double precision version for the last registration stage (Figure~\ref{fig:mean-time-per-iteration-ants}).
+Therefore, the slowdown was not due to a slower convergence of the algorithm, but rather to an increased execution time for each iteration.
 
-After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the use of templates in ANTs' C++ code\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of its \TG{clarify what 'its' refers to} function to be in double precision, with \textit{EvaluateAtContinuousIndex} contributing most of the CPU time.
+After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the use of templates in ANTs' C++ code\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of the interpolation function to be in double precision, with \textit{EvaluateAtContinuousIndex} contributing most of the CPU time.
 Therefore, the pipeline wastes CPU cycles to convert the input data to double-precision in ITK, and potentially to cast back to single-precision in ANTs.
@@
 	\includegraphics[width=\linewidth]{figures/ants-registrationSyN-iteration-mean.png}
-	\caption{Average time per iteration for ANTS registrationSyN in double and single precision.
-Only the SyN Registration stage is shown as the two earlier stage were near zero time.
-The error bars show the standard deviation.}
+	\caption{Average time per iteration for ANTs registrationSyN in double and single precision.
+Only the SyN Registration stage is shown as the two earlier stages were near zero time.
+The error bars show the standard deviation across n=20 subjects.}
 	\label{fig:mean-time-per-iteration-ants}
@@
 Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded executions.
-In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP \TG{expand acronym} was a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime.
+In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), the OpenMP multithreading library was a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime.
 FreeSurfer used the multi-processing API provided by OpenMP.
@@
 	\caption{FreeSurfer recon-all analysis.
-The y-axis shows the average CPU time spent in each function, with error bars showing the standard deviation \TG{across subjects?}.
+The y-axis shows the average CPU time spent in each function, with error bars showing the standard deviation across n=20 subjects.
 The x-axis shows the function ordered by decreasing CPU time grouped by module.
@@
 The parallel efficiency decreased from 66\% with two threads down to 5\% with 32 threads.
-While Amdhal's law limits the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} indicates that thread Synchronization had a larger impact \TG{how does the figure indicate that?}.
-
+While Amdhal's law limits the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} shows that most of the CPU time is spent in libiomp, which suggests that thread synchronization had a larger impact on parallel efficiency.
 \begin{figure}
@@
 One way to mitigate this challenge is to lower the sampling rate during the profiling.
-However, to reach a similar level of memory consumption \TG{than what?}, the sampling rate must be significantly reduced.
+However, the sampling rate would have to be significantly reduced for the finalization stage to require  \SI{16}{\giga\byte} of RAM or less.
 This would lead to a potential loss of information in the profiling results.
-Because of these limitations, we were unable to finalize the results of fMRIPrep \TG{do you mean the entire fmriprep pipeline?} for analysis.
+Because of these limitations, we were unable to finalize the results of the entire fMRIPrep pipeline and we only reported results for the sub-pipelines.
 
 \section{Conclusion}
-In this paper, we profiled several MRI pre-processing pipelines: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep \TG{was it finalized?}.
+In this paper, we profiled several MRI pre-processing pipelines used in the popular fMRIPrep tool: ANTs brainExtraction, ANTs registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all.
 We observed that few functions contribute to a large amount of CPU time which offers opportunities to optimize a practical amount of functions to potentially obtain significant speed-up values.
@@
 Linear interpolation was the primary bottleneck for these pipelines.
-We discovered a bug in ITK which leads the implementation of ANTS registration in single-precision to use double precision in practice.
+We discovered a bug in ITK which leads the implementation of ANTs registration in single-precision to use double precision in practice.
 We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP.

### 2024-05-22 paper/paper.tex
Small edits in conclusion

@@
 We observed that few functions contribute to a large amount of CPU time which offers opportunities to optimize a practical amount of functions to potentially obtain significant speed-up values.
-We found that most pipelines suffered from memory bound.
-Linear interpolation was the primary bottleneck for these pipelines.
-We discovered a bug in ITK which leads the implementation of ANTs registration in single-precision to use double precision in practice.
+We found that most pipelines suffered from memory bound, and that linear interpolation was the primary time bottleneck.
+We discovered a bug in ITK which leads the implementation of ANTs registration in single-precision to use double precision instead.
 We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP.
@@
 Future optimization efforts could focus on reduced precision techniques, to reduce memory bound and accelerate interpolation computation.
-We note that the use of reduced precision is not trivial though, as seen with the ANTs bug.
+We note that the use of reduced precision is not trivial, as seen with the ANTs bug.
 Therefore, careful attention would be required to find a balance between performance and accuracy.

### 2024-05-23 paper/paper.tex
[FIX] minor fixes.

@@
 \subsection{Pipelines}
-fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, built on top of several well-known neuroimaging toolboxes such as ANTs, FSL, FreeSurfer and AFNI \TG{expand acronyms and add references for each toolbox}.
+fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, built on top of several well-known neuroimaging toolboxes such as Advanced Normalization Tools (ANTs), FMRIB Software Library (FSL), FreeSurfer and Analysis of Functional NeuroImages (AFNI) \TG{add references for each toolbox}.
 Although processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity correction and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain.
@@
 
-\begin{minipage}{\linewidth}
+\begin{minipage}{0.9\linewidth}
 	\lstinputlisting[
@@
 \\\href{https://doi.org/10.5281/zenodo.10987491}{https://doi.org/10.5281/zenodo.10987491}
-\MD{TODO finalize the metadata entry and publish}
 

### 2024-05-27 paper/paper.tex
[FIX] comments.

@@
 \title{An Analysis of Performance Bottlenecks in MRI Pre-Processing}
-\author{Mathieu Dugr\'e, Yohan Chatelain, Tristan Glatard\\Department of Computer Science and Software Engineering, Concordia University, Montreal, QC, Canada}
+\author{Mathieu Dugr\'e*, Yohan Chatelain, Tristan Glatard
+\\Department of Computer Science and Software Engineering, Concordia University, Montreal, QC, Canada
+\\\{mathieu.dugre, yohan.chatelain, tristan.glatard\}@concordia.ca}
 
@@
 
-% % HPCs
-% Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts \HL{leveraging} a large number of CPU cores and high-bandwidth memory to process subjects in parallel.
-However, \HL{it can be challenging to run HPCs in developing countries due to power restrictions, national network infrastructure, and lack of training and funding}~\cite{NRG-SC23}.
-\HL{Similarly, researcher use cloud computing, but it requires expertise to setup, can be costly for large studies, and} data transfer can be slow and might be impossible due to ethic concerns related to human data.
-\HL{In both cases, limitation from the parallel efficiency of the pipeline can lead to under-utilization of the available resources.}
-
-% % DL
-% In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing.
-For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} to provide Deep Learning alternatives to traditional algorithms\HL{, providing faster runtime.}
-% % Moreover, a study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
-% However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive.
-Additionally, \HL{their results} can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.
-
-% % Alternative solutions
-% \HL{These techniques require costly hardware to be effective, which can be unacessible to underrepresented researchers.} Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements but require substantial effort to implement.
-Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.
-
-% - No comprehensive study of the performance bottleneck of MRI pre-processing.
-
 % - Focus on MRI pre-processing.
@@
 \subsection{Pipelines}
-fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, built on top of several well-known neuroimaging toolboxes such as Advanced Normalization Tools (ANTs), FMRIB Software Library (FSL), FreeSurfer and Analysis of Functional NeuroImages (AFNI) \TG{add references for each toolbox}.
+fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, built on top of several well-known neuroimaging toolboxes such as Advanced Normalization Tools (ANTs)~\cite{Avants2020-xx}, FMRIB Software Library (FSL)~\cite{Jenkinson2012-tq}, FreeSurfer~\cite{Fischl2012-cx} and Analysis of Functional NeuroImages (AFNI)~\cite{Cox1996-nk}.
 Although processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity correction and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain.
@@
 We profiled the entire fMRIPrep pipeline to obtain a coarse view of the pipeline's performance bottleneck.
-Additionally, to characterize the performance at a finer grain, we profiled the sub-pipelines of fMRIPrep, namely \TG{add list}.
+Additionally, to characterize the performance at a finer grain, we profiled the default sub-pipelines of fMRIPrep listed above.
 While we used the same sub-pipeline and order as in fMRIPrep, we used the default parameters for most of the sub-pipelines which may vary from the default parameters of fMRIPrep.
@@
 We used the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
-The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts.
+The anatomical and functional data from 20 health individuals was acquired from two different cohorts.
 The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two sessions: eyes open and eyes closed.
 The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session.
-\TG{add information on image resolution}
-
+Each subject has an anatomical image with resolution of 1mm and a dimension of $160 \times 240 \times 256$ voxels (approx.
+\SI{12}{\mebi\byte}).
+Additionally, the subjects have 300 functional images taken at 2 seconds interval with a resolution of 3mm and a dimension of $64 \times 64 \times 35$ voxels (approx.
+\SI{50}{\mebi\byte}).
 \subsection{Profiling}
@@
 We discuss our approach to these challenges in the next section.
-\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for references}
 
@@
 For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}}.
-\lstlistingname~\ref{lst:vtune_example} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.
+\lstlistingname~\ref{lst:vtune_example} demonstrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.
 
@@
 \subsection{Impact of Memory Bounded Functions}
-Table~\ref{extab:memory-bound} shows that fetching data from disk was the main cause for starving the CPU \TG{add data fetching time \% to table}.
+Table~\ref{extab:memory-bound} shows that fetching data from disk was the main cause for starving the CPU.
 Next, the L1 bound had the highest cache-level contribution to the memory bound of the pipelines.
@@
 
-\csvnames{csvcol}{1=\pipeline, 2=\mem, 3=\la, 4=\lb, 5=\lc, 6=\dram, 7=\store}
+\csvnames{csvcol}{1=\pipeline, 2=\mem, 3=\la, 4=\lb, 5=\lc, 6=\dram, 7=\store, 8=\disk}
 \csvstyle{memory_bound}{
-	tabular = |r|c|c|c|c|c|c|,
-	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound\\\hline\hline,
+	tabular = |r|c|c|c|c|c|c|c|,
+	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound & \% Disk Bound\\\hline\hline,
 	late after line = \\\hline,
@@
 	\csvreader[memory_bound]{tables/memory_bound.csv}{}
-	{\pipeline & \tablenum[round-precision=2]{\mem} & \tablenum[round-precision=2]{\la} & \tablenum[round-precision=2]{\lb} & \tablenum[round-precision=2]{\lc} & \tablenum[round-precision=2]{\dram} & \tablenum[round-precision=2]{\store}}
+	{\pipeline & \tablenum[round-precision=2]{\mem} & \tablenum[round-precision=2]{\la} & \tablenum[round-precision=2]{\lb} & \tablenum[round-precision=2]{\lc} & \tablenum[round-precision=2]{\dram} & \tablenum[round-precision=2]{\store} & \tablenum[round-precision=2]{\disk}}
 	\caption{Impact of data load on stalled CPU cycles.
@@
 For both pipelines, the makespan was significantly higher in single precision than in double precision.
-This was unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv} than in double precision.
+This was unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision than in double precision~\cite{Wang2018-jv}.
 
@@
 
-After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the use of templates in ANTs' C++ code\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of the interpolation function to be in double precision, with \textit{EvaluateAtContinuousIndex} contributing most of the CPU time.
+After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the use of templates in ITK's C++ code\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of the interpolation function to be in double precision, with \textit{EvaluateAtContinuousIndex} contributing most of the CPU time.
 Therefore, the pipeline wastes CPU cycles to convert the input data to double-precision in ITK, and potentially to cast back to single-precision in ANTs.
@@
 In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), the OpenMP multithreading library was a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime.
-FreeSurfer used the multi-processing API provided by OpenMP.
 Moreover, the y-axis scale is multi folds larger than for the single-threaded execution, indicating that the multi-threaded execution was significantly slower than the single-threaded execution.
@@
 The parallel efficiency decreased from 66\% with two threads down to 5\% with 32 threads.
-While Amdhal's law limits the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} shows that most of the CPU time is spent in libiomp, which suggests that thread synchronization had a larger impact on parallel efficiency.
+While Amdhal's law limits the maximum parallel efficiency of the pipeline, Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads} shows that most of the CPU time is spent in libiomp, which suggests that thread synchronization had a larger impact on parallel efficiency.
 \begin{figure}
@@
 This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency.
-We speculated that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads.
+We speculate that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads.
 Unfortunately, we failed to recompile FreeSurfer having changed the \textit{static} OpenMP scheduling to \textit{dynamic}.
@@
 We found this step to be both time-consuming and resource intensive.
-For example, finalizing the results from a single FreeSurfer recon-all execution can require over \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline.
+For example, finalizing the results from a single FreeSurfer recon-all execution can require over \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline\footnote{\href{https://surfer.nmr.mgh.harvard.edu/fswiki/SystemRequirements}{https://surfer.nmr.mgh.harvard.edu/fswiki/SystemRequirements}}.
 This significant difference in resource requirement creates a challenge to profile pipeline with long runtime, due to the potential difficulty to obtain access to compute nodes with enough RAM.
 One way to mitigate this challenge is to lower the sampling rate during the profiling.
-However, the sampling rate would have to be significantly reduced for the finalization stage to require  \SI{16}{\giga\byte} of RAM or less.
-This would lead to a potential loss of information in the profiling results.
+However, the sampling rate would have to be significantly reduced for the finalization stage to require a similar amount of RAM (\SI{16}{\giga\byte} or less), which would lead to a potential loss of information in the profiling results.
 Because of these limitations, we were unable to finalize the results of the entire fMRIPrep pipeline and we only reported results for the sub-pipelines.
@@
 \section{Conclusion}
-In this paper, we profiled several MRI pre-processing pipelines used in the popular fMRIPrep tool: ANTs brainExtraction, ANTs registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all.
+In this paper, we profiled several MRI pre-processing pipelines used in the popular fMRIPrep tool: ANTs brainExtraction, ANTs registrationSyN, FSL FAST, FSL FLIRT, FSL MCFLIRT, and FreeSurfer recon-all.
 We observed that few functions contribute to a large amount of CPU time which offers opportunities to optimize a practical amount of functions to potentially obtain significant speed-up values.
@@
 
-% \TG{Move this paragraph to future work}
-% Despite an extensive literature on linear interpolation \MD{TODO cite it}, no particular optimization techniques are available in the MRI pre-processing pipelines we profiled.
-We think it would be interesting to explore simple optimization techniques.
-For example, the authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation.
-We think it would be interesting for future work to explore these different alternatives to accelerate interpolation in the pipelines, while considering the trade-off between accuracy and computational cost.
-Reduced- or mixed-precision techniques could also be explored to accelerate interpolation, with similar trade-off.
-Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements.
-Improving the performance of interpolation would have a large impact for 3D MRI imaging processing, and other fields such as climate simulation, astronomical imaging, or 3D image reconstruction.
 \section{Data Availability}
@@
 \\\href{https://github.com/mathdugre/mri-bottleneck}{https://github.com/mathdugre/mri-bottleneck}
-\MD{TODO Update URL to slashbin org, when created.}
 
@@
 
-\MD{From Tristan: If possible, include the colorful figure in the paper.}
-
-\MD{Not sure if these add value?}
 \label{sec:supplementary}
@@
 	\centering
-	\includegraphics{figures/zipf_law.png}
+	\includegraphics[width=0.66\columnwidth]{figures/zipf_law.png}
 	\caption{Distribution of the functions' CPU time compare to a Zipf's law distribution with $\alpha=1.46$.
@@
 }
-\newcommand{\csvtable}[3]{
+
 \begin{table}[ht]
-		\resizebox*{\textwidth}{!}{
-			\csvreader[hotspot]{#1}{}
-			% \csvreader[hotspot, filter={\value{csvrow}<60}]{#1}{}
+	\centering
+	\resizebox*{0.66\columnwidth}{!}{
+		\csvreader[hotspot]{tables/hotspots-32threads-freesurfer-reconall.csv}{}
 		{\thecsvrow & \module & \func & \tablenum[round-precision=2, round-mode=places]{\mean}$\pm$\tablenum[round-precision=2, round-mode=places]{\std}}
 	}
-		\caption{#2}
-		\label{#3}
+	\caption{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the pipeline makespan.}
+	\label{extab:hotspots-32threads-freesurfer-reconall}
 \end{table}
-}
-% CSV tables
-% \csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the pipeline makespan.}{extab:hotspots-1thread-freesurfer-reconall}
-\csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the pipeline makespan.}{extab:hotspots-32threads-freesurfer-reconall}
 

### 2024-05-27 sbatch/ants/registrationSyN-fp.sh

@@ -22,6 +22,8 @@ export SIF_OPTION="-B $HOME/.cache/templateflow:/templateflow"
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 TMPLT="/templateflow/tpl-MNI152NLin2009cAsym/"
 
+mkdir -p /data/derivatives/ants/registrationSyN-fp/sub-${SUBJECT_ID}/ses-open/anat
+
 antsRegistrationSyN.sh \
     -d 3 \
     -f \${TMPLT}/tpl-MNI152NLin2009cAsym_res-01_desc-brain_T1w.nii.gz\

### 2024-05-27 sbatch/ants/registrationSyN.sh

@@ -22,6 +22,8 @@ export SIF_OPTION="-B $HOME/.cache/templateflow:/templateflow"
 cat <<EOT >> ${TMP_SCRIPT}/${RANDOM_STRING}.sh
 TMPLT="/templateflow/tpl-MNI152NLin2009cAsym/"
 
+mkdir -p /data/derivatives/ants/registrationSyN/sub-${SUBJECT_ID}/ses-open/anat
+
 antsRegistrationSyN.sh \
     -d 3 \
     -f \${TMPLT}/tpl-MNI152NLin2009cAsym_res-01_desc-brain_T1w.nii.gz\

### 2024-10-08 paper/paper.tex
restore paper.tex
Recover changes to paper.tex

@@
 \\Department of Computer Science and Software Engineering, Concordia University, Montreal, QC, Canada
-\\\{mathieu.dugre, yohan.chatelain, tristan.glatard\}@concordia.ca}
+\\research@mathdugre.me}
 
@@
 Therefore, it is important to understand the performance bottlenecks of MRI pre-processing pipelines to improve their performance.
-Using Intel VTune profiler, we characterized the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes.
-We found that few functions contributed to most of the CPU time, and that linear interpolation was the largest contributor.
+Using the Intel VTune profiler, we characterized the bottlenecks of several commonly used MRI preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes.
+We found few functions contributed to most of the CPU time, and that linear interpolation was the largest contributor.
 Data access was also a substantial bottleneck.
@@
 Pre-processing of Magnetic Resonance Imaging (MRI) data is a critical step in neuroimaging.
-Although established pre-processing pipelines exist, they commonly require extensive amounts of computation and produce large volumes of output and intermediate data.
+Although established preprocessing pipelines exist, they commonly require extensive amounts of computation and produce large volumes of output and intermediate data.
 Such requirements create challenges when studying large neuroimaging cohorts, and they limit clinical applicability when timely data analysis is needed.
 Therefore, the neuroimaging community is constantly seeking for ways to improve the computational performance of MRI pre-processing pipelines.
-In this paper, we characterize the computational cost of several commonly adopted MRI pre-processing pipelines, providing a reference benchmark for future efforts to optimize MRI pre-processing pipelines.
+In this paper, we characterize the computational cost of several commonly adopted MRI preprocessing pipelines, providing a reference benchmark for future efforts to optimize MRI pre-processing pipelines.
 
@@
 While other modalities such as electroencephalography (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRI for its broad adoption in research and non-invasiveness.
-Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.
 
@@
 Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data.
-We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} pipeline as it a comprehensive pre-processing application for both sMRI and fMRI.
+We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} pipeline, as it is a comprehensive pre-processing application for both sMRI and fMRI.
 fMRIPrep combines several pipelines to produce a complete pre-processing pipeline.
@@
 We profile the pipeline in single-threaded mode and using 32 threads to account for parallel executions.
-We aggregate subjects' results for each pipeline to obtain an average performance profile.
+We aggregate subjects' results for each pipeline to get an average performance profile.
 
@@
 \subsection{Pipelines}
-fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, built on top of several well-known neuroimaging toolboxes such as Advanced Normalization Tools (ANTs)~\cite{Avants2020-xx}, FMRIB Software Library (FSL)~\cite{Jenkinson2012-tq}, FreeSurfer~\cite{Fischl2012-cx} and Analysis of Functional NeuroImages (AFNI)~\cite{Cox1996-nk}.
-Although processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity correction and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain.
+fMRIPrep is a commonly used fMRI and sMRI preprocessing pipeline for neuroimaging, built on top of several well-known neuroimaging toolboxes such as Advanced Normalization Tools (ANTs)~\cite{Avants2020-xx}, FMRIB Software Library (FSL)~\cite{Jenkinson2012-tq}, FreeSurfer~\cite{Fischl2012-cx} and Analysis of Functional NeuroImages (AFNI)~\cite{Cox1996-nk}.
+Although processing steps may vary based on the type of input data, the sMRI pipeline uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity correction and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain.
 When enabled, FreeSurfer recon-all~\cite{Dale1999-wu} is used for surface reconstruction.
 We benchmarked a recon-all command similar to fMRIPrep's, which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3.
-The fMRI pipeline generally performs motion correction with FSL MCFLIRT~\cite{Jenkinson2002-od}, slice timing correction with AFNI 3dtshift~\cite{Cox1996-nk}, and co-registration with FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}.
+The fMRI pipeline performs motion correction with FSL MCFLIRT~\cite{Jenkinson2002-od}, slice timing correction with AFNI 3dtshift~\cite{Cox1996-nk}, and co-registration with FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}.
 
-We profiled the entire fMRIPrep pipeline to obtain a coarse view of the pipeline's performance bottleneck.
-Additionally, to characterize the performance at a finer grain, we profiled the default sub-pipelines of fMRIPrep listed above.
-While we used the same sub-pipeline and order as in fMRIPrep, we used the default parameters for most of the sub-pipelines which may vary from the default parameters of fMRIPrep.
-We omitted some sub-pipelines of fMRIPrep as they were not compatible with our dataset.
+We profiled the entire fMRIPrep pipeline to get a coarse view of the pipeline's performance bottleneck.
+To characterize the performance at a finer grain, we profiled the default sub-pipelines of fMRIPrep listed above.
+While we used the same sub-pipeline and order as in fMRIPrep, we used the default parameters for most of the sub-pipelines, which may vary from the default parameters of fMRIPrep.
+We omitted some sub-pipelines of fMRIPrep, as they were not compatible with our dataset.
 For example, slice-timing correction was already performed in our data.
@@
 We used the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}.
-The anatomical and functional data from 20 health individuals was acquired from two different cohorts.
+The anatomical and functional data from 20 healthy individuals were acquired from two different cohorts.
 The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two sessions: eyes open and eyes closed.
 The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session.
-Each subject has an anatomical image with resolution of 1mm and a dimension of $160 \times 240 \times 256$ voxels (approx.
+Each subject has an anatomical image with a resolution of \SI{1}{\milli\meter} and a dimension of $160 \times 240 \times 256$ voxels (approx.
 \SI{12}{\mebi\byte}).
-Additionally, the subjects have 300 functional images taken at 2 seconds interval with a resolution of 3mm and a dimension of $64 \times 64 \times 35$ voxels (approx.
+The subjects have 300 functional images taken at 2 seconds interval with a resolution of \SI{3}{\milli\meter} and a dimension of $64 \times 64 \times 35$ voxels (approx.
 \SI{50}{\mebi\byte}).
+
 \subsection{Profiling}
 Profiling MRI pre-processing pipelines raises a few challenges.
-Neuroimaging pipelines use several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages.
+Neuroimaging pipelines use several programming languages including, \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines using a combination of multiple languages.
 Pipelines are generally complex and computationally expensive.
 
-The Intel VTune profiler addresses these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level.
+The Intel VTune profiler addresses these challenges by offering multi-language support, low performance overhead, and information on functions at the runtime level.
 However, common profiling challenges remain.
 Profilers require source code to be compiled with debug symbols to report interpretable information about the different functions and modules.
-Moreover, different input data can substantially impact the performance results due to conditional branching in the pipeline and convergence thresholds.
+Different input data can substantially impact the performance results because of conditional branching in the pipeline and convergence thresholds.
 We discuss our approach to these challenges in the next section.
 
-First, to obtain human-readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
+First, to get human-readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image.
 For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}}.
@@
 
-We profiled the pipeline using two threading configuration: single-threaded and 32 threads.
+We profiled the pipeline using two threading configurations: single-threaded and 32 threads.
 The single-threaded approach simplified the analysis and limited the potential I/O overhead on the profiling.
-The 32-thread approach used all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the pipeline.
+The 32-thread approach used all available CPU cores from a computing node of our infrastructure to study the multi-threading performance of the pipeline.
 This configuration provided a more realistic usage of the pipeline.
 
-Before profiling each pipeline, we transferred the input data from the shared file system to the compute node.
+Before profiling each pipeline, we transferred the input data from the shared file system to the computing node.
 After profiling, we transferred back the output to the shared file system.
-This limited profiling variability due to I/O contingency.
+This limited profiling variability is due to I/O contingency.
 
@@
 In principle, this could have increased the profiling overhead.
-However, in practice there was no impact since the data generated by the VTune profiler was very small and was not written on the same file system as data produced by the pipelines.
+However, there was no impact, since the data generated by the VTune profiler was tiny and was not written on the same file system as data produced by the pipelines.
 
@@
 VTune provides several metrics to characterize the performance of a pipeline.
-The makespan is the elapsed time from the start to the end of the pipeline.
+makespan measures elapsed time from the start to the end of the pipeline.
 The CPU time refers to the total time spent by the CPU, across all cores, to execute the pipeline.
 
-The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load.
-L1 bound occurs when a demand load stalls CPU cores although the data already reside in L1 cache.
-Similarly, L2, L3, and DRAM bound occurs when a demand load stalls the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively.
-The store bound occurs when a store operation stalls the CPU.
-Supplementary Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}.
+The memory-bound metrics, \HL{in particular, focus on identifying bottlenecks where the CPU cannot proceed due to data unavailability.
+Specifically, memory-bound metrics count the number of cycles where the CPU is stalled waiting for in-flight memory demand loads to be satisfied.
+This indicates how efficiently or inefficiently the pipeline utilizes memory resources, highlighting scenarios where memory latency hinders performance.}
+
+\HL{L1 bound occurs when a demand load stalls CPU cores, even though the required data is already in the L1 cache.
+Similarly, L2, L3, and DRAM bound conditions occur when a demand load causes the CPU to stall due to a cache miss at a particular cache level, with the data instead residing in L2, L3, or DRAM, respectively.
+In each of these cases, the CPU is delayed while waiting for the requested data to be fetched from the next level in the memory hierarchy.}
+
+\HL{A store-bound condition arises when a store operation halts CPU execution, often due to insufficient memory bandwidth or resource conflicts.
+These metrics help isolate whether the performance bottleneck is due to memory bandwidth limitations or inefficient caching.} Supplementary Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}.
 We use the convention from~\cite{Kukunas2015-jd} for the equations.
 
-The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level, that is, when the data is being fetched from disk.
+\HL{The difference between the total memory-bound value and the sum of the other metrics indicates the proportion of CPU idle time caused by waiting for data that is missing from all cache levels, meaning the data is being retrieved from secondary storage, such as a disk.}
 
@@
 \section{Results \& Discussion}
-In this section we present the aggregated profiling data obtained with VTune.
-First, we present a performance overview of the pipelines.
-We then analysis specific hotspots in the pipelines.
+In this section, we present the aggregated profiling data collected using VTune.
+First, we present a performance overview of the pipelines, then analyze specific hotspots in the pipelines.
 Unless specified explicitly, we only present the profile data from using 1-thread since the results from 32-threads are similar.
-
-We aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile.
-Consistently with~\cite{Kukunas2015-jd}, we hypothesized that the Pareto principle applied to the performance of MRI pre-processing pipelines, we therefore only studied the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the pipeline.
+We aggregate the profiling results of each pipeline across all subjects to get an average performance profile.
+Consistently with~\cite{Kukunas2015-jd}, we hypothesized that the Pareto principle applied to performance of MRI pre-processing pipelines.
+We therefore only studied the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the pipeline.
 This heuristic allowed us to focus our efforts on the performance critical sections of a pipeline.
@@
 While consistent with the Pareto principle, the distribution of the functions' CPU time did not follow a Pareto distribution.
-Using the \textit{powerlaw} Python package~\cite{Alstott2014-gr}, we found that the data fit a Zipf law with exponent $\alpha=1.46$ (Supplementary Figure~\ref{fig:zips-law}).
+Using the \textit{powerlaw} Python package~\cite{Alstott2014-gr}, we found the data fit a Zipf law with an exponent $\alpha=1.46$ (Supplementary Figure~\ref{fig:zips-law}).
 Therefore, the CPU time of a function was approximately inversely proportional to its rank.
-Thus, future efforts to optimize this minimal set of functions could result in significant performance improvements.
+Thus, future efforts to optimize this minimal set of functions could bring significant performance improvements.
 
@@
 	\includegraphics[width=\linewidth]{figures/global-longtail.png}
-	\caption{Distribution of the functions' CPU time.
-The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage.
+	\caption{Distribution of CPU time for the functions.
+The left y-axis shows the average total CPU Time spent in a function, while the right y-axis shows the cumulative CPU time percentage.
 The x-axis is the percentage of functions ordered by decreasing CPU time.
-The data includes all functions from all pipelines.
-}
+Data includes functions from all pipelines.}
 	\label{fig:long-tail-distribution}
@@
 \subsection{Main Bottleneck: Linear Interpolation}
-Table~\ref{extab:interpolation} shows that interpolation was a critical part for all pipelines except FreeSurfer recon-all, contributing between 32.20\% to 62.70\% of the total CPU time.
-Overall, the number of functions using interpolation was low, with less than 20 in each pipeline: few interpolation functions contributed to a large portion of the CPU time.
+Table~\ref{extab:interpolation} shows that interpolation was a critical part of all pipelines except FreeSurfer recon-all, contributing between 32.20\% to 62.70\% of the total CPU time.
+Overall, the number of functions using interpolation was low, with less than 20 in each pipeline.
+Few interpolation functions dominated CPU time.
 While each interpolation had low computational cost, the large amount of operations resulted in a significant bottleneck.
@@
 	\caption{Contribution of interpolation to the pipelines' total CPU time.
-The percentage is the average sum of CPU time of functions using interpolation.
+The percentage is the averaged sum of the CPU time of functions using interpolation.
 The data includes all functions from all pipelines.}
@@
 \subsection{Impact of Memory Bounded Functions}
-Table~\ref{extab:memory-bound} shows that fetching data from disk was the main cause for starving the CPU.
-Next, the L1 bound had the highest cache-level contribution to the memory bound of the pipelines.
+Table~\ref{extab:memory-bound} shows that fetching data from disks was the main cause for starving the CPU\HL{, contributing an average of 6.60\% across pipelines}.
+Next, the L1 bound had the highest cache level contribution \HL{(5.58\% average)} to the memory bound of the pipelines.
 ANTs pipelines had the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT.
-L2 and L3 bound was low for all pipelines, although higher for FreeSurfer recon-all.
-The DRAM bound was high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTs brainExtraction in single and double precision.
+L2 and L3 bound were low \HL{(0.88\% and 1.17\% average)} for all pipelines, although higher for FreeSurfer recon-all.
+The DRAM bound was high for both FSL FLIRT \HL{(15.73\%)} and FreeSurfer recon-all \HL{(13.83\%)}, followed by ANTs brainExtraction in \HL{double (5.08\%)} and \HL{single (2.48\%)} precision.
 Store bound was low for all pipelines.
@@
 Optimizing the data access of the pipeline could reduce the number of data load.
-This would potentially have a large impact on the performance of the pipelines.
+This could have a large impact on pipelines performance.
 Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound.
-To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing.
+To reduce the memory bound impact of fetching data from disks, it could be possible to convert and store the data to a lower precision before pre-processing.
 In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
@@
 	\caption{Impact of data load on stalled CPU cycles.
-For each metric (memory, L1, L2, L3, DRAM, and store), reported values are summations weighted by function CPU time, averaged over n=20 subjects, and obtained in single-threaded mode.
+For each metric \HL{(Total, L1, L2, L3, DRAM, Store, and Disk Fetch)}, reported values are summations weighted by function CPU time, averaged over $n=20$ subjects, and collected in single-threaded mode.
 They represent the percentage of the total CPU time stalled for the metric.}
@@
 \subsection{ANTs: Single vs. Double Precision}
-Figure~\ref{fig:makespan-ants} shows the makespan of ANTs brainExtraction and ANTs registrationSyN in double and single precision.
+Figure~\ref{fig:makespan-ants} shows the makespan of ANTs brainExtraction and ANTs registrationSyN with double and single precision.
 For both pipelines, the makespan was significantly higher in single precision than in double precision.
@@
 
-After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the use of templates in ITK's C++ code\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of the interpolation function to be in double precision, with \textit{EvaluateAtContinuousIndex} contributing most of the CPU time.
-Therefore, the pipeline wastes CPU cycles to convert the input data to double-precision in ITK, and potentially to cast back to single-precision in ANTs.
-We think this bug causes the slow-down.
-We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base.
+After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the use of templates in ITK's C++ code\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of the interpolation function to be in double precision, with \textit{EvaluateAtContinuousIndex} contributing most to the CPU time.
+As a result, the pipeline wastes CPU cycles by converting the input data to double-precision in ITK and potentially casting it back to single-precision in ANTs.
+We think this bug causes the slowdown.
+We tried re-compiling a version of ITK to fix this issue, but we were unsuccessful because of the size and complexity of the code base.
 This example shows that the use of reduced precision is not trivial and requires a deep understanding of the code base.
-Moreover, benchmarks should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
+Benchmarks should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.
 We reported this issue in the ITK GitHub repository\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}}.
@@
 Only the SyN Registration stage is shown as the two earlier stages were near zero time.
-The error bars show the standard deviation across n=20 subjects.}
+The error bars show the standard deviation across $n=20$ subjects.}
 	\label{fig:mean-time-per-iteration-ants}
@@
 Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded executions.
-In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), the OpenMP multithreading library was a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime.
-Moreover, the y-axis scale is multi folds larger than for the single-threaded execution, indicating that the multi-threaded execution was significantly slower than the single-threaded execution.
+In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), the OpenMP multi-threading library was a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime.
+The y-axis scale is multi folds larger than for the single-threaded execution, showing that the multi-threaded execution was significantly slower than the single-threaded execution.
 
@@
 The x-axis shows the function ordered by decreasing CPU time grouped by module.
-We omitted function names for clarity.
-The function ID are dependent to each plot.
-Supplementary materials show the mapping of the function ID to the function name for each plot.}
+We only display module names and omitted function names for clarity.}
 	\label{fig:hotspots-freesurfer-reconall}
@@
 While Amdhal's law limits the maximum parallel efficiency of the pipeline, Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads} shows that most of the CPU time is spent in libiomp, which suggests that thread synchronization had a larger impact on parallel efficiency.
+
 \begin{figure}
@@
 This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency.
-We speculate that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads.
-Unfortunately, we failed to recompile FreeSurfer having changed the \textit{static} OpenMP scheduling to \textit{dynamic}.
+We speculate that \textit{dynamic} scheduling could improve the parallel efficiency by assigning chunks based on the current load of the threads.
+Unfortunately, we failed to re-compile FreeSurfer having changed the \textit{static} OpenMP scheduling to \textit{dynamic}.
 This remains challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.
@@
 The VTune profiler provided a rich and extensive amount of data for our analysis, which would have been difficult to retrieve otherwise.
-Thanks to its low performance overhead, we were able to profile the pipeline with minimal impact on the runtime, allowing the use of more subjects to obtain a more accurate performance profile.
+Thanks to its low performance overhead, we could profile the pipeline with minimal impact on the runtime, allowing the use of more subjects to get a more accurate performance profile.
 The documentation provided by VTune is extensive and well-documented, which helped us to understand the different metrics and events available.
@@
 We found this step to be both time-consuming and resource intensive.
-For example, finalizing the results from a single FreeSurfer recon-all execution can require over \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline\footnote{\href{https://surfer.nmr.mgh.harvard.edu/fswiki/SystemRequirements}{https://surfer.nmr.mgh.harvard.edu/fswiki/SystemRequirements}}.
-This significant difference in resource requirement creates a challenge to profile pipeline with long runtime, due to the potential difficulty to obtain access to compute nodes with enough RAM.
+For example, finalizing the results from a single FreeSurfer recon-all execution can require over \SI{2}{\tera\byte} of RAM, while the developers recommend using between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline\footnote{\href{https://surfer.nmr.mgh.harvard.edu/fswiki/SystemRequirements}{https://surfer.nmr.mgh.harvard.edu/fswiki/SystemRequirements}}.
+This significant difference in resource requirement creates a challenge to profile pipeline with long runtime, because of the potential difficulty to get access to computing nodes with enough RAM.
 One way to mitigate this challenge is to lower the sampling rate during the profiling.
 However, the sampling rate would have to be significantly reduced for the finalization stage to require a similar amount of RAM (\SI{16}{\giga\byte} or less), which would lead to a potential loss of information in the profiling results.
-Because of these limitations, we were unable to finalize the results of the entire fMRIPrep pipeline and we only reported results for the sub-pipelines.
+Because of these limitations, we could not collect the results of the entire fMRIPrep pipeline, and we only reported results for the sub-pipelines.
 
@@
 In this paper, we profiled several MRI pre-processing pipelines used in the popular fMRIPrep tool: ANTs brainExtraction, ANTs registrationSyN, FSL FAST, FSL FLIRT, FSL MCFLIRT, and FreeSurfer recon-all.
-We observed that few functions contribute to a large amount of CPU time which offers opportunities to optimize a practical amount of functions to potentially obtain significant speed-up values.
-We found that most pipelines suffered from memory bound, and that linear interpolation was the primary time bottleneck.
-We discovered a bug in ITK which leads the implementation of ANTs registration in single-precision to use double precision instead.
-We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP.
-Lastly, we discussed the challenge of profiling long-running pipelines with VTune, due to computational resource requirements.
+A few functions were found to contribute significantly to the CPU time, presenting opportunities to optimize these functions and potentially achieve significant speed-up.
+It was discovered that linear interpolation is the main time bottleneck, with most pipelines being affected by memory bound.
+We discovered a bug in ITK, which leads to the version of ANTs registration in single precision to use double precision instead.
+A potential bug in FreeSurfer recon-all was found, which limits the benefit of multi-threading with OpenMP.
+Last, we discussed the challenge of profiling long-running pipelines with VTune, because of computational resource requirements.
 
-The dataset we chose contains wide range of age and equal distribution of sex.
+The dataset we chose contains a wide range of age and equal distribution of sex.
 However, it only contains healthy subjects.
 It would be interesting to study the impact of image quality and pathologies on performance.
-Moreover, our profiling with VTune did not account for the spatiality (e.g.
+Our profiling with VTune did not account for the spatiality (e.g.
 for loop) or temporality (e.g.
@@
 Therefore, careful attention would be required to find a balance between performance and accuracy.
-We also suggest that future work study the different configurations of OpenMP scheduling in FreeSurfer recon-all, to improve parallel efficiency.
+We also suggest for future work to study the different configurations of OpenMP scheduling in FreeSurfer recon-all, to improve parallel efficiency.
 This would lead to faster executions time and higher CPU usage in HPC allocations.

### 2024-10-11 paper/paper.tex
Fix table and figures to improve clarity.

@@
 \csvstyle{memory_bound}{
-	tabular = |r|c|c|c|c|c|c|c|,
-	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound & \% Disk Bound\\\hline\hline,
+	tabular = |r|c||c|c|c|c|c|c|,
+	table head = \hline Pipeline &  Total Memory Bound \% & L1 \% & L2 \% &  L3 \% & DRAM \% & Store \% & Disk Fetch \%\\\hline\hline,
 	late after line = \\\hline,
@@
 	\caption{FreeSurfer recon-all analysis.
-The y-axis shows the average CPU time spent in each function, with error bars showing the standard deviation across n=20 subjects.
+Error bars show the standard deviation across n=20 subjects.
 The x-axis shows the function ordered by decreasing CPU time grouped by module.
